### 2017-09-21 paper/draft.tex
populate draft with first content

\documentclass[a4paper]{scrartcl}
\usepackage[english]{babel}
\usepackage{natbib}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\title{The relative roles of genetic, ecological and environmental factors determining lineage radiations}
\author{Ludwig Leidinger}
\date{\today}
%
\begin{document}
\maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}

\section{Material and Methods}

\section{Results}

\section{Discussion}

\addbibresource{refs.bib}
\printbibliography

\end{document}

### 2017-09-21 paper/draft.tex
lay out some intro

@@
 
-\title{The relative roles of genetic, ecological and environmental factors determining lineage radiations}
+\title{The relative roles of genetic, ecological and environmental factors in species radiations}
 \author{Ludwig Leidinger}
@@
 \section{Introduction}
+island biota have always fascinated and inspired researchers (Darwin, Wallace, ...)
 
+This interest in islands and their recognition as highly suitable model systems resulted in the formulation of
+important theoretical frameworks, foremost MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB).
+Later ... GDM.
+
+These theories have in common a focus on species numbers and respective rates of migration, extinction and speciation.
+While they are successful in describing species numbers, they are neutral in design (in the UNTB sense) and thus were not meant to consider species traits and compositions.
+
+Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
+island colonisation, while others remain one single species.
+
+These patterns have been explained theoretically [here, here and here], but usually one at a time.
+In the following, we aim to produce and thus explain these forementioned patterns using a single framework, and thus following the same set of (rather simple) rules.
+
+Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
+and what trait combinations constitute successful colonizer and radiator [this sounds like heating ;)] species, respectively.
 \section{Material and Methods}

### 2017-09-21 paper/draft.tex
lay out abstract

@@
 \begin{abstract}
+  Islands feature peculiar and fascinating biotas, often very distinct from any other island or continental area.
+  [Examples of radiations/good and poor radiators, island syndromes etc.]
+  Although island biogeography theory has seen an ever increasing development over the years,
+  the mechanisms underlying these patterns are not fully understood/ not yet explained in a direct causal relationship.
+  Specifically, some processes remain un- or under-explored, foremost of these adaption and micro-evolution.
+  We used an individual-based mechanistic model implementing these processes and other commonly investigated ecological and
+  environmental processes to assess the importance of drivers across different organisational and spatio-temporal levels
+  on species trait distribution on islands.
+  This enabled us to also take a closer look on the properties of lineages/species that actually managed to colonize islands,
+  and which of them resulted in radiations[~].
 
@@
 Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
-island colonisation, while others remain one single species.
+island colonisation, while others remain one single species (Price/Chase).
 

### 2017-09-22 paper/draft.tex
include some refs/ideas

@@
 Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
-island colonisation, while others remain one single species (Price/Chase).
+island colonisation, while others remain one single species (Price/Chase) (Price & Wagner 2004).
+
+Also: Ecological Opportunity and Adaptive Radiation, Stroud & Losos 2016: ``Does the concept have predictive value[...]?''
+Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity
 

### 2017-09-22 paper/draft.tex
include refs

--- 2017-09-21 paper/draft.tex
@@
 Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
-island colonisation, while others remain one single species (Price/Chase).
+island colonisation, while others remain one single species (Price/Chase) (eg. \cite{price2004speciationPrice}).
+
+Also: Ecological Opportunity and Adaptive Radiation, Stroud & Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
+Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs.
+phylogenetic similarity \cite{losos2008phylogenetic}
 

### 2017-09-22 paper/draft.tex
fix

@@
 
-Also: Ecological Opportunity and Adaptive Radiation, Stroud & Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
+Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
 Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs.

### 2017-09-22 MIbGxMCmod.jl
lay out iscompatible function

@@ -599,6 +599,23 @@ function compete!(world::Array{Patch,1})
     end
 end
 
+function iscompatible(mate::Individual, ind::Individual)
+    tolerance = ind.traits["reptol"]
+    length(ind.genome) != length(mate.genome) && return false
+    sum(x -> length(x.genes), ind.genome) != sum(x -> length(x.genes), mate.genome) && return false
+    indgenes = Gene[]
+    for chrm in ind.genome
+        append!(indgenes, chrm.genes)
+    end
+    sort!(indgenes, by = x -> x.id)
+    mategenes = Gene[]
+    for chrm in mate.genome
+        append!(mategenes, chrm.genes)
+    end
+    sort!(mategenes, by = x -> x.id)
+    map(x -> x.id, indgenes) != map(x -> x.id, mategenes) && return false
+end
+
 function findposspartners(world::Array{Patch,1}, ind::Individual, location::Tuple{Int64, Int64})
     ind.isnew = true
     radius = ind.traits["repradius"] + 0.5 # to account for cell width ... or not??
@@ -606,10 +623,11 @@ function findposspartners(world::Array{Patch,1}, ind::Individual, location::Tupl
     for x in -radius:radius, y in -radius:radius
         sqrt(x^2 + y^2) <= radius && push!(coordinates, (x + location[1], y + location[2]))
     end
-    map(x -> checkborderconditions!(world, x[1], x[2]), coordinates)
+    coordinates = map(x -> checkborderconditions!(world, x[1], x[2]), coordinates)
     posspartners = Individual[]
     map(x -> append!(posspartners, x.community), filter(x -> in(x.location, coordinates), world))
     filter!(x -> x.size >= x.traits["repsize"], posspartners)
+    filter!(x -> iscompatible(x, ind), posspartners)
     filter!(x -> !x.isnew, posspartners) # filter out mating individual
     ind.isnew = false
     posspartners

### 2017-09-22 MIbGxMCmod.jl
iscompatible function

@@ -614,6 +614,21 @@ function iscompatible(mate::Individual, ind::Individual)
     end
     sort!(mategenes, by = x -> x.id)
     map(x -> x.id, indgenes) != map(x -> x.id, mategenes) && return false
+    indgenesize = length(*map(x -> x.id, indgenes)...)
+    mategenesize = length(*map(x -> x.id, mategenes)...)
+    1 - (abs(mategenesize - indgenesize) / mategenesize) < tolerance && return false
+    basediffs = 0
+    for i in eachindex(indgenes)
+        for j in eachindex(indgenes[i].sequence)
+            try
+                indgenes[i].sequence[j] != mategenes[i].sequence[j] && basediffs += 1
+            catch
+                basediffs += 1
+            end
+        end
+    end
+    1 - (basediffs / mategenesize) < tolerance && return false
+    true
 end
 
 function findposspartners(world::Array{Patch,1}, ind::Individual, location::Tuple{Int64, Int64})

### 2017-09-22 MIbGxMCmod.jl
lint

@@ -79,204 +79,6 @@ Patch(id,location) = Patch(id,location,298,100,false,0,0,Individual[],false)
 ## Methods/Functions:
 #####################
 
-function readmapfile(filename::String)
-    println("Reading file \"$filename\"...")
-    mapstrings = String[]
-    open(filename) do file
-        mapstrings = readlines(file)
-    end
-    mapstrings = filter(x->!all(isspace,x),mapstrings) # remove empty lines
-    mapstrings = filter(x->x[1]!='#',mapstrings) # remove comment lines
-    mapsubstrings = map(split,mapstrings)
-    mapentries = map(x->map(String,x),mapsubstrings)
-    timesteps = 0
-    try
-        timesteps = parse(Int,filter(x->size(x,1)==1,mapentries)[1][1])
-    catch
-        timesteps = 1000
-        warn("your mapfile \"$filename\" does not include timestep information. Assumed $timesteps timesteps.")
-    end
-    mapentries = filter(x->size(x,1)>1,mapentries)
-    return timesteps,mapentries
-end
-
-function createworld(maptable::Array{Array{String,1},1}, settings::Dict{String,Any})
-    println("Creating world...")
-    world = Patch[]
-    area = 100 # CAVE: just for now...
-    for entry in maptable
-        size(entry,1) < 3 && error("please check your map file for incomplete or faulty entries. \n
-                Each line must contain patch information with at least \n
-                \t - a unique integer ID, \n
-                \t - an integer x coordinate, \n
-                \t - an integer y coordinate, \n
-                separated by a whitespace character (<ID> <x> <y>).")
-        id = parse(Int64, entry[1])
-        xcord = parse(Int64, entry[2])
-        ycord = parse(Int64, entry[3])
-        size(entry,1) > 3 ? temperature = parse(Float64, entry[4]) : temperature = 298.0
-        isisland = false
-        if size(entry,1) > 4 && contains(lowercase(entry[5]),"island")
-            isisland = true # islands do not receive an initial community
-        end
-        newpatch = Patch(id,(xcord,ycord),temperature,area,isisland)
-        if size(entry,1) > 5 && contains(lowercase(entry[6]),"isolated")
-            newpatch.isolated = true
-        end
-        !isisland && append!(newpatch.community,genesis(settings["linkage"]))
-        push!(world,newpatch)
-    end
-    world
-end
-
[…]

### 2017-09-22 MIbGxMCmod.jl
include tolerance info from options

@@ -24,6 +24,7 @@ const normconst = 1e10 # normalization constant to get biologically realistic or
 const mutscaling = 50 #parse(ARGS[2])
 const sexualreproduction = true
 const meangenes = 20 # mean number of genes per individual
+const mutationrate = 1e-3
 
 
 ## Types:
@@ -140,11 +141,11 @@ function mutate!(ind::Individual, temp::Float64)
                     end
                     gene.sequence[i] = newbase
                     for trait in gene.codes
+                        contains(trait.name, "rep") && contains(trait.name, "tol") && continue
                         trait.value == 0 && (trait.value = rand(Normal(0,0.01)))
                         newvalue = trait.value + rand(Normal(0, trait.value/mutscaling)) # CAVE: mutscaling! new value for trait
-                        (newvalue > 1 && contains(trait.name,"prob")) && (newvalue=1)
-                        (newvalue > 1 && contains(trait.name,"reptol")) && (newvalue=1)
                         newvalue < 0 && (newvalue=abs(newvalue))
+                        (newvalue > 1 && contains(trait.name,"prob")) && (newvalue=1)
                         while newvalue == 0 #&& contains(trait.name,"mut")
                             newvalue = trait.value + rand(Normal(0, trait.value/mutscaling))
                         end
@@ -476,14 +477,8 @@ function reproduce!(world::Array{Patch,1}, patch::Patch) #TODO: refactorize!
                 noffs = rand(Poisson(metaboffs))
                 if sexualreproduction
                     posspartners = findposspartners(world, patch.community[idx], patch.location)
-                    partneridx = idx
-                    try
-                        partneridx = rand(filter(x->x!=idx,posspartners))
-                    catch
-                        idx += 1
-                        continue
-                    end
-                    partnergenome = meiosis(patch.community[partneridx].genome, false) #CAVE: maybe move inside offspring loop?
+                    partner = rand(posspartners)
+                    partnergenome = meiosis(partner.genome, false) #CAVE: maybe move inside offspring loop?
                     mothergenome = meiosis(patch.community[idx].genome, true)
                     for i in 1:noffs
                         (size(partnergenome,1) < 1 || size(mothergenome,1) < 1) && continue
@@ -513,7 +508,7 @@ function reproduce!(world::Array{Patch,1})
     end
 end
 
-function createtraits(traitnames::Array{String,1})
[…]

### 2017-09-22 main.jl
edit comment header

@@ -5,7 +5,8 @@
 ##
 ## Main simulation
 ##
-## Synopsis: julia main.jl <MAPFILE> <RANDOMSEED> <TIMESTEPS>
+## For a list of options, run julia main.jl --help
+##
 ## <MAPFILE> is a textfile containing information about the simulation arena
 ## Every line describes one patch in the following format:
 ## <ID> <X-COORDINATE> <Y-COORDINATE> [<TYPE>]

### 2017-09-22 main.jl
lint

@@ -56,9 +56,9 @@ any(path -> path == thisDir, LOAD_PATH) || push!(LOAD_PATH, thisDir)
 #        "--flag1"
 #            help = "an option without argument, i.e. a flag"
 #            action = :store_true
-        "arg1"
-            help = "a positional argument"
-            required = false
+#        "arg1"
+#            help = "a positional argument"
+#            required = false
         end
     return parse_args(s)
 end
@@ -105,6 +105,6 @@ const replicates = startseed:startseed+nprocesses-1
 
 TT = STDOUT # save original STDOUT stream
 redirect_stdout()
-pmap(x->runit(true,mapfiles,x),replicates) # compilation/optimization run
+pmap(x->runit(true,allargs,x),replicates) # compilation/optimization run
 redirect_stdout(TT) # restore STDOUT
-pmap(x->runit(false,mapfiles,x),replicates)
+#pmap(x->runit(false,allargs,x),replicates)

### 2017-09-22 MIbGxMCmod.jl
fixes

@@ -477,7 +477,12 @@ function reproduce!(world::Array{Patch,1}, patch::Patch) #TODO: refactorize!
                 noffs = rand(Poisson(metaboffs))
                 if sexualreproduction
                     posspartners = findposspartners(world, patch.community[idx], patch.location)
-                    partner = rand(posspartners)
+                    try
+                        partner = rand(posspartners)
+                    catch
+                        idx += 1
+                        continue
+                    end
                     partnergenome = meiosis(partner.genome, false) #CAVE: maybe move inside offspring loop?
                     mothergenome = meiosis(patch.community[idx].genome, true)
                     for i in 1:noffs
@@ -588,7 +593,8 @@ function createchrs(nchrs::Int64,genes::Array{Gene,1})
     chromosomes
 end
 
-function genesis(ninds::Int64=100, meangenes::Int64=meangenes,
+function genesis(linkage::String="random", tolerance::String="evo",
+                 ninds::Int64=100, meangenes::Int64=meangenes,
                  traitnames::Array{String,1} = ["ageprob",
                                                 "dispmean",
                                                 "dispprob",
@@ -602,8 +608,7 @@ function genesis(ninds::Int64=100, meangenes::Int64=meangenes,
                                                 "reptol",
                                                 "seedsize",
                                                 "temptol",
-                                                "tempopt"], # minimal required traitnames
-                 linkage::String="random", tolerance::String="evo")
+                                                "tempopt"]) # minimal required traitnames
     community = Individual[]
     for ind in 1:ninds
         ngenes = rand(Poisson(meangenes))

### 2017-09-22 main.jl

@@ -84,7 +84,7 @@ end
     seed != 0 && srand(seed)
     mapfiles =  map(x->String(x),split(allargs["maps"],","))
     if firstrun
-        world=createworld([["1","1","1"]])
+        world=createworld([["1","1","1"]],settings)
         simulation(world, seed, "", 10)
     else
         for i in 1:length(mapfiles)

### 2017-09-22 MIbGxMCmod.jl
order of FUN defs

@@ -631,6 +631,27 @@ function genesis(linkage::String="random", tolerance::String="evo",
     community
 end
 
+function readmapfile(filename::String)
+    println("Reading file \"$filename\"...")
+    mapstrings = String[]
+    open(filename) do file
+        mapstrings = readlines(file)
+    end
+    mapstrings = filter(x->!all(isspace,x),mapstrings) # remove empty lines
+    mapstrings = filter(x->x[1]!='#',mapstrings) # remove comment lines
+    mapsubstrings = map(split,mapstrings)
+    mapentries = map(x->map(String,x),mapsubstrings)
+    timesteps = 0
+    try
+        timesteps = parse(Int,filter(x->size(x,1)==1,mapentries)[1][1])
+    catch
+        timesteps = 1000
+        warn("your mapfile \"$filename\" does not include timestep information. Assumed $timesteps timesteps.")
+    end
+    mapentries = filter(x->size(x,1)>1,mapentries)
+    return timesteps,mapentries
+end
+
 function createworld(maptable::Array{Array{String,1},1}, settings::Dict{String,Any})
     println("Creating world...")
     world = Patch[]
@@ -658,27 +679,6 @@ function createworld(maptable::Array{Array{String,1},1}, settings::Dict{String,A
         push!(world,newpatch)
     end
     world
-end
-
-function readmapfile(filename::String)
-    println("Reading file \"$filename\"...")
-    mapstrings = String[]
-    open(filename) do file
-        mapstrings = readlines(file)
-    end
-    mapstrings = filter(x->!all(isspace,x),mapstrings) # remove empty lines
-    mapstrings = filter(x->x[1]!='#',mapstrings) # remove comment lines
-    mapsubstrings = map(split,mapstrings)
-    mapentries = map(x->map(String,x),mapsubstrings)
-    timesteps = 0
-    try
-        timesteps = parse(Int,filter(x->size(x,1)==1,mapentries)[1][1])
-    catch
-        timesteps = 1000
-        warn("your mapfile \"$filename\" does not include timestep information. Assumed $timesteps timesteps.")
-    end
-    mapentries = filter(x->size(x,1)>1,mapentries)
-    return timesteps,mapentries
 end
 
 function updateworld!(world::Array{Patch,1},maptable::Array{Array{String,1},1}) #TODO: add functionality to remove patches!

### 2017-09-22 main.jl
re-enable proper simulation run

@@ -99,7 +99,7 @@ end
 
 ## Parallel stuff:
 const nprocesses = nworkers()
-allargs = parsecommandline()
+const allargs = parsecommandline()
 const startseed = allargs["seed"]
 const replicates = startseed:startseed+nprocesses-1
 
@@ -107,4 +107,4 @@ TT = STDOUT # save original STDOUT stream
 redirect_stdout()
 pmap(x->runit(true,allargs,x),replicates) # compilation/optimization run
 redirect_stdout(TT) # restore STDOUT
-#pmap(x->runit(false,allargs,x),replicates)
+pmap(x->runit(false,allargs,x),replicates)

### 2017-09-22 types.jl
disable Plot usage

@@ -1,308 +1,0 @@
-#!/usr/bin/env julia
-## contains all custom types necessary for eco-evo-env speciation island model
-
-## NOTE: all functions: make methods for ind/pop/island/world level!!!!
-
-module GeneInds
-
-using Distributions
-
-## Global(?) variables/constants:
-## sdmut: mutation standard deviation
-
-
-## Types:
-
-mutable struct Trait
-    name::String
-    value::Float64 # numerical value
-    strength::Float64 # mutation strength
-    codedby::Array{String,1}
-end
-
-mutable struct Gene
-    sequence::Array{Char,1} # contains gene base code
-    id::String # gene identifier
-    codes::Array{Trait,1}
-end
-
-struct Chromosome #CAVE! mutable?
-    genes::Array{Gene,1} # 1D array of genes
-    ##    origin::Bool # parental origin of chromosome (paternal/maternal)
-end
-
-
-mutable struct Individual
-    genome::Array{Chromosome,1} # genome = 2D array of chromosomes (>=1 sets)
-    traits::Array{Trait,1}
-    stage::String # demographic stage of individual
-    isnew::Bool # indicator whether individual is new to a patch or has already dispersed etc.
-    fitness::Float64 # reproduction etc. scaling factor representing life history
-    size::Float64 # body size/mass -> may replace stage.
-end
-
-mutable struct Patch
-    community::Array{Individual,1} # holds the population (1D) of prob not: present species (2nd D)
-    altitude::Float64 # altitude: corresponds to T
-    nichea::Float64 # additional niches,
-    nicheb::Float64 # e.g. precipitation
-end
-
-mutable struct Island # needs new type? -> rather world
-    patches::Array{Patch,2} # 2D grid
-end
-
-
-## methods:
-function compete!(patch::Patch)
-    while sum(map(x->x.size,patch.community)) > 1 # occupied area larger than available
-        victim = rand(1:size(patch.community,1))
-        splice!(patch.community, victim)
-    end
-end
-
-function reproduce!(patch::Patch)
-    idx = 1
-    while idx <= size(patch.community,1)
-        if patch.community[idx].size >= 0.5 * patch.community[idx].traits[find(x->x.name=="maxsize",
-                                                                               patch.community[idx].traits)][1].value
-            noffs = rand(Poisson(patch.community[idx].traits[find(x->x.name=="noffspring",
[…]

### 2017-11-02 paper/draft.tex
fix
fix biblatex/biber
fix citation
csquotes
csquotes after inputenc
new angle for intro

@@
 \usepackage[english]{babel}
-\usepackage{natbib}
+\usepackage[
+  backend=biber,
+  style=authoryear
+]{biblatex}
+\addbibresource{refs.bib}
+
 \usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc}
+\usepackage{csquotes}
 
@@
 Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
-island colonisation, while others remain one single species (Price/Chase) (eg. \cite{price2004speciationPrice}).
+island colonisation, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
 
@@
 Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
-and what trait combinations constitute successful colonizer and radiator [this sounds like heating ;)] species, respectively.
+and what trait combinations constitute successful colonizer and radiating species, respectively.
+
+[NEW ANGLE:]
+island biota have always fascinated and inspired researchers (Darwin, Wallace, ...).
+While often featuring unique species communities and harboring countless endemics,
+islands are especially threatened by climate change, human impact and invasive species.
+
+This interest in islands and their recognition as highly suitable model systems resulted in the formulation of
+important theoretical frameworks, foremost MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB).
+Later ... GDM.
+
+Because the study of islands biogeography takes into account evolutionary time frames (millions of years),
+comprehensive and meaningful data is difficult to obtain.
+In a recent review \cite{leidinger2017biodiversity}, we analyzed a powerful alternative to field experiments in island biogeography
+--- mechanistic (/simulation) models.
+Unfortunately, most island models failed to include more explicit genetic or micro evolution.
+Instead some models featured a concept known as ``protracted speciaton'' \cite{rosindellXXX}, where speciation occurs with a delay from the actual
+divergence event if the population remains isolated during that time.
+
 \section{Material and Methods}
@@
 
-\addbibresource{refs.bib}
 \printbibliography

### 2017-11-02 paper/draft.tex
extend new angle for intro (stub)

@@
 divergence event if the population remains isolated during that time.
+But nature is usually more complex than this and genetic mechanisms have important ramifications for speciation/radiation and species richness.
+[EXAMPLES!]
+
+Another shortfall of the reviewd models was the lack of integrating ecological niches and adaption.
+
+In the following, we present a model that remedies both of the adressed shortcomings by implementing genetic processes on a niche-explicit background
+in an island biogeography framework.
+We will adress the questions of what relative impact does (a) genetic linkage and (b) reproductive isolation have on emerging species numbers and diversity,
+(c) what ...
+We find that
 

### 2017-11-02 paper/draft.tex
extend new angle for intro (complete sentence)

@@
 
+Besides adaptive radiations(/evolution), changes in or different measures of geographic isolation (landscape complexity) can give rise to non-adaptive radiations.
+
 In the following, we present a model that remedies both of the adressed shortcomings by implementing genetic processes on a niche-explicit background
@@
 We will adress the questions of what relative impact does (a) genetic linkage and (b) reproductive isolation have on emerging species numbers and diversity,
-(c) what ...
-We find that
+and how (c) landscape heterogeneity (habitat niche diverstiy) and (d) topographical complexity differently affect radiatons.
 

### 2017-11-02 paper/draft.tex
lay out shortcomings in disc

@@
 
-Besides adaptive radiations(/evolution), changes in or different measures of geographic isolation (landscape complexity) can give rise to non-adaptive radiations.
+Besides adaptive radiations(/evolution), changes in or different measures of geographic isolation (landscape complexity) can also give rise to non-adaptive radiations.
 
@@
 \section{Discussion}
+Shortcomings. (/Processes/Factors not considered.)
+(Dynamic) resource availability: assumed constant in model, but petentially more variation, complexity in community dynamics.
+Interaction: in present model only indirect interaction via competition, but predation, parasitation, herbivory, mutualism will increase complexity.
+Especially pollination is assumed to be an important interaction, because having a high impact on evolution and establishment of species due to specialization.
+[maybe some words on trophic interactions.]
 

### 2017-11-02 paper/draft.tex
lay out m&ms

@@
 \section{Material and Methods}
+We use an individual based metacommunity model written in julia \cite{JULIALANG}.
+Each individual is characterized by a set of genetically controlled traits and is submitted to processes of
+establishment, competition, growth, (density independent) mortality, reproduction and dispersal.
+Each of which processes is affected by the individuals' characteristics.
+The simulation arena consists of a larger landmass (the mainland) holding an initial pool of evolving populations and a smaller, initially unpopulated island some distance apart.
+During the course of the simulation, individuals may disperse from the mainland to the island, thus colonizing it.
 
+A detailed protocol can be found in the appendix (\cite{grimm2010odd}, XXX).
+
+We record the complete state of our simulation world at the end of a simulation run and each time the arena changes (geomorphological dynamics).
+We additionally record every individual that colonizes the island to track it in the phylogeny.
+
+Using the thus simulated data, we calculate a phylogeny of the extant species, species numbers, etc. [WHAT ELSE?]
 \section{Results}

### 2017-11-22 paper/draft.tex
comment m&ms
further questions to intro
write some intro

@@
 \section{Introduction}
-island biota have always fascinated and inspired researchers (Darwin, Wallace, ...)
+%% island biota have always fascinated and inspired researchers (Darwin, Wallace, ...)
 
-This interest in islands and their recognition as highly suitable model systems resulted in the formulation of
-important theoretical frameworks, foremost MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB).
-Later ... GDM.
+%% This interest in islands and their recognition as highly suitable model systems resulted in the formulation of
+%% important theoretical frameworks, foremost MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB).
+%% Later ... GDM.
 
-These theories have in common a focus on species numbers and respective rates of migration, extinction and speciation.
-While they are successful in describing species numbers, they are neutral in design (in the UNTB sense) and thus were not meant to consider species traits and compositions.
+%% These theories have in common a focus on species numbers and respective rates of migration, extinction and speciation.
+%% While they are successful in describing species numbers, they are neutral in design (in the UNTB sense) and thus were not meant to consider species traits and compositions.
 
-Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
-island colonisation, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
+%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
+%% island colonisation, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
 
-Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
-Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs.
+%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
+%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs.
 phylogenetic similarity \cite{losos2008phylogenetic}
 
-These patterns have been explained theoretically [here, here and here], but usually one at a time.
-In the following, we aim to produce and thus explain these forementioned patterns using a single framework, and thus following the same set of (rather simple) rules.
+%% These patterns have been explained theoretically [here, here and here], but usually one at a time.
+%% In the following, we aim to produce and thus explain these forementioned patterns using a single framework, and thus following the same set of (rather simple) rules.
 
-Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
-and what trait combinations constitute successful colonizer and radiating species, respectively.
+%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
+%% and what trait combinations constitute successful colonizer and radiating species, respectively.
 
-[NEW ANGLE:]
-island biota have always fascinated and inspired researchers (Darwin, Wallace, ...).
-While often featuring unique species communities and harboring countless endemics,
-islands are especially threatened by climate change, human impact and invasive species.
+%% [NEW ANGLE:]
+Island biota have always fascinated and inspired researchers (Darwin, Wallace, ...). %TODO: refs
+Nowadays islands are understood as well suitable model systems for ecology and evolution due to their well definable characteristics like isolation. %TODO: ref
+However island biota are also increasingly at risk.
+While often featuring unique species communities and harboring countless endemics, %TODO: refs
+islands are especially threatened by climate change, human impact and invasive species. %TODO: refs
+This makes understanding the functioning and evolution of island communities a high priority objective.
 
-This interest in islands and their recognition as highly suitable model systems resulted in the formulation of
-important theoretical frameworks, foremost MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB).
-Later ... GDM.
+One of the first and to this day haighly influential formal theories of island biodiversity is MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB). %TODO: ref
+In it, species richness on islands is the result of opposing rates of immigration and extinction, both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
+This relationship was since successfully applied to a multitude of islands %TODO: refs
+, but also island like systems, such as coral reefs, lakes/ponds or forest fragments. %TODO: refs
+This general applicability, however, can not hide the fact that the ETIB fails to deliver a mechanistic explanation of the processes at play.
+It has particularly been criticized for its lack of considering speciation. %TODO: refs
 
-Because the study of islands biogeography takes into account evolutionary time frames (millions of years),
+About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM). %TODO: ref
+The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
+whith islands emerging from sea level, increasing over time in size and topographic complexity, before eroding and subsiding in the ocean again.
+In doing so, the model incorporates concepts of niche theory, speciation and species succession.
+Yet, there are still many processes and drivers deemed important, which are not considered by any formal theory.
+
+%%TODO: here come a list of processes with examples
+
+The problem of investigating these processes is the time frame needed for a proper evaluation.
+Because islands biogeography takes into account evolutionary time frames (millions of years),
 comprehensive and meaningful data is difficult to obtain.
@@
 Unfortunately, most island models failed to include more explicit genetic or micro evolution.
-Instead some models featured a concept known as ``protracted speciaton'' \cite{rosindellXXX}, where speciation occurs with a delay from the actual
+Instead, some models featured a concept known as ``protracted speciaton'' \cite{rosindellXXX}, where speciation occurs with a delay from the actual
 divergence event if the population remains isolated during that time.
@@
 and how (c) landscape heterogeneity (habitat niche diverstiy) and (d) topographical complexity differently affect radiatons.
+[further questions thus tackled: traits of colonizer species, ...]
 
@@
 A detailed protocol can be found in the appendix (\cite{grimm2010odd}, XXX).
+
+[SCENARIOS, REPLICATES]
 

### 2017-11-22 paper/draft.tex
add to M&M

@@
 We will adress the questions of what relative impact does (a) genetic linkage and (b) reproductive isolation have on emerging species numbers and diversity,
-and how (c) landscape heterogeneity (habitat niche diverstiy) and (d) topographical complexity differently affect radiatons.
+and how (c) landscape heterogeneity (habitat niche diversity) and (d) topographical complexity differently affect radiatons.
 [further questions thus tackled: traits of colonizer species, ...]
@@
 
-[SCENARIOS, REPLICATES]
+\subsection{scenarios}
+In the simulations, we modify the value and extent of
+(a) gene linkage,
+(b) reproductive isolation,
+(c) habitat diversity, and
+(d) topographical complexity
+to asses their relative contribution on community composition.
+\Cref{tab:scenarios} summarizes the parameters which were varied for the scenarios, their meaning and their values. %TODO: make table.
+For each parameter combination, we launched a number of %TODO: how much?
+replicates.
 
+\subsection{Data recording}
 We record the complete state of our simulation world at the end of a simulation run and each time the arena changes (geomorphological dynamics).
-We additionally record every individual that colonizes the island to track it in the phylogeny.
+Additionally, we store the state of the simulation arena each 1000 %TODO: how much exactly?
+time steps and record every individual that colonizes the island to track it in the phylogeny.
 

### 2017-11-22 MIbGxMCmod.jl
comments to iscompatible

@@ -455,9 +455,9 @@ end
 
 function iscompatible(mate::Individual, ind::Individual)
     tolerance = ind.traits["reptol"]
-    length(ind.genome) != length(mate.genome) && return false
-    sum(x -> length(x.genes), ind.genome) != sum(x -> length(x.genes), mate.genome) && return false
-    indgenes = Gene[]
+    length(ind.genome) != length(mate.genome) && return false # check equal chromosome numbers
+    sum(x -> length(x.genes), ind.genome) != sum(x -> length(x.genes), mate.genome) && return false # check equal gene numbers
+    indgenes = Gene[] # TODO: this could be a separate function
     for chrm in ind.genome
         append!(indgenes, chrm.genes)
     end
@@ -467,21 +467,21 @@ function iscompatible(mate::Individual, ind::Individual)
         append!(mategenes, chrm.genes)
     end
     sort!(mategenes, by = x -> x.id)
-    map(x -> x.id, indgenes) != map(x -> x.id, mategenes) && return false
+    map(x -> x.id, indgenes) != map(x -> x.id, mategenes) && return false # check equal gene identity
     indgenomesize = length(*(map(x -> String(x.sequence), indgenes)...))
     mategenomesize = length(*(map(x -> String(x.sequence), mategenes)...))
-    1 - (abs(mategenomesize - indgenomesize) / mategenomesize) < tolerance && return false
+    1 - (abs(mategenomesize - indgenomesize) / mategenomesize) < tolerance && return false # check length difference of genome
     basediffs = 0
     for i in eachindex(indgenes)
         for j in eachindex(indgenes[i].sequence)
             try
                 indgenes[i].sequence[j] != mategenes[i].sequence[j] && (basediffs += 1) # alternatively use bioinformatics tools
-            catch
+            catch # e.g., in case of differently long genomes
                 basediffs += 1
             end
         end
     end
-    1 - (basediffs / mategenomesize) < tolerance && return false
+    1 - (basediffs / mategenomesize) < tolerance && return false # CAVE: maybe compare w/ indgenome?
     true
 end
 

### 2017-11-22 MIbGxMCmod.jl
check findposspartners

@@ -487,7 +487,7 @@ end
 
 function findposspartners(world::Array{Patch,1}, ind::Individual, location::Tuple{Int64, Int64})
     ind.isnew = true
-    radius = floor(ind.traits["repradius"] + 0.5) # to account for cell width ... or not??
+    radius = floor(ind.traits["repradius"] + 0.5) # CAVE: to account for cell width ... or not??
     coordinates = Tuple[]
     for x = -radius:radius, y = -radius:radius
         sqrt(x^2 + y^2) <= radius && push!(coordinates, (x + location[1], y + location[2]))

### 2018-01-10 paper/draft.tex
additions to intro and disc

@@
 %%TODO: here come a list of processes with examples
+%% - genetic processes in general
+%% - isolation processes
 
@@
 [EXAMPLES!]
+
+For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
+seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al. 2013) % cite properly!
 
@@
 
+In reality linkage between genes is not a binary decision, but rather a consequence of the distance between those genes.
+The larger the distance, the higher the probability of crossing over during meiosis.
+The size of the resulting genomic blocks or islands, however, may increase to several megabases in diverging population ---
+a phenomenom coined as ``divergence hitchhiking'' (Via \& West 2008). % cite properly!
+This mechanism is thought to facilitate sympatric speciation by reducing the detrimental effects of gene flow and
+recombination between populations to speciation (Via 2011, but see Feder \& Nosil 2010).
+During incipient speciation, however, th
+
 \printbibliography

### 2018-01-11 paper/draft.tex
typo

@@
 
-One of the first and to this day haighly influential formal theories of island biodiversity is MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB). %TODO: ref
+One of the first and to this day highly influential formal theories of island biodiversity is MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB). %TODO: ref
 In it, species richness on islands is the result of opposing rates of immigration and extinction, both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
@@
 , but also island like systems, such as coral reefs, lakes/ponds or forest fragments. %TODO: refs
-This general applicability, however, can not hide the fact that the ETIB fails to deliver a mechanistic explanation of the processes at play.
+This general applicability, however, cannot hide the fact that the ETIB fails to deliver a mechanistic explanation of the processes at play.
 It has particularly been criticized for its lack of considering speciation. %TODO: refs

### 2018-01-11 paper/draft.tex
additions to intro

@@
 
+Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al. 2017).
+A subset of these questions deals with 
 One of the first and to this day highly influential formal theories of island biodiversity is MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB). %TODO: ref
@@
 --- mechanistic (/simulation) models.
-Unfortunately, most island models failed to include more explicit genetic or micro evolution.
+Unfortunately, most island models failed to include more explicit genetic or micro evolution processes.
+% reasons: underlying theories: Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al. 2017).
+%A subset of these questions deals with 
 Instead, some models featured a concept known as ``protracted speciaton'' \cite{rosindellXXX}, where speciation occurs with a delay from the actual
@@
 In the following, we present a model that remedies both of the adressed shortcomings by implementing genetic processes on a niche-explicit background
-in an island biogeography framework.
+using an island biogeography framework.
 We will adress the questions of what relative impact does (a) genetic linkage and (b) reproductive isolation have on emerging species numbers and diversity,

### 2018-01-16 MIbGxMCmod.jl
indent

@@ -173,7 +173,7 @@ function mutate!(patch::Patch, settings::Dict{String,Any})
 end
 
 function mutate!(world::Array{Patch, 1}, settings::Dict{String,Any})
-        pmap(x -> mutate!(x, settings), world)
+    pmap(x -> mutate!(x, settings), world)
 end
 
 function checkviability!(patch::Patch) # may consider additional rules... # maybe obsolete anyhow...

### 2018-01-16 MIbGxMCmod.jl
remove remnant of gene id

@@ -592,7 +592,7 @@ function creategenes(ngenes::Int64,traits::Array{Trait,1})
         end
     end
     if !any(map(x -> length(x.codes) == 0, genes)) # make sure there is a neutral gene!
-        push!(genes, Gene(String(rand(collect("acgt"), 20)), randstring(8), Trait[]))
+        push!(genes, Gene(String(rand(collect("acgt"), 20)), Trait[]))
     end
     genes
 end

### 2018-01-16 MIbGxMCmod.jl
fitness is multiplicative

@@ -220,27 +220,27 @@ function establish!(patch::Patch, nniches::Int64=1)
     idx = 1
     while idx <= size(patch.community,1)
         fitness = 1
-        if !traitsexist(patch.community[idx], ["temptol", "tempopt"])
+        if !traitsexist(patch.community[idx], ["temptol", "tempopt"]) # CAVE: should check for more traits!
             splice!(patch.community, idx) # kill it!
             continue
         elseif patch.community[idx].isnew || patch.community[idx].age == 0
             opt = patch.community[idx].traits["tempopt"]
             tol = patch.community[idx].traits["temptol"]
-            fitness -= 1 - gausscurve(opt, tol, temp)
+            fitness *= gausscurve(opt, tol, temp)
             fitness > 1 && (fitness = 1) # should be obsolete
             fitness < 0 && (fitness = 0) # should be obsolete
         end
         if nniches >= 2 && (patch.community[idx].isnew || patch.community[idx].age == 0)
             opt = patch.community[idx].traits["precopt"]
             tol = patch.community[idx].traits["prectol"]
-            fitness -= 1 - gausscurve(opt, tol, patch.nichea)
+            fitness *= gausscurve(opt, tol, patch.nichea)
             fitness > 1 && (fitness = 1) # should be obsolete
             fitness < 0 && (fitness = 0) # should be obsolete
         end
         if nniches == 3 && (patch.community[idx].isnew || patch.community[idx].age == 0)
             opt = patch.community[idx].traits["nicheopt"]
             tol = patch.community[idx].traits["nichetol"]
-            fitness -= 1 - gausscurve(opt, tol, patch.nicheb)
+            fitness *= gausscurve(opt, tol, patch.nicheb)
             fitness > 1 && (fitness = 1) # should be obsolete
             fitness < 0 && (fitness = 0) # should be obsolete
         end  

### 2018-01-16 MIbGxMCmod.jl
reduce mortality

@@ -21,7 +21,7 @@ export Patch, # types
 const boltz = 1.38064852e-23 # J/K = m2⋅kg/(s2⋅K)
 const act = 1e-19 # activation energy /J, ca. 0.63eV - Brown et al. 2004
 const growthrate = exp(25.2) # global base growth/biomass production from Brown et al. 2004
-const mortality = exp(22) # global base mortality from Brown et al. 2004 is 26.3, but competition introduces add. mort.
+const mortality = exp(20) # global base mortality from Brown et al. 2004 is 26.3, but competition introduces add. mort.
 const fertility = exp(23.8) # global base reproduction rate from Brown et al. 2004, alternatively 25.0
 const phylconstr = 50 #parse(ARGS[2])
 # const meangenes = 20 # mean number of genes per individual

### 2018-01-16 MIbGxMCmod.jl
comment

@@ -21,7 +21,7 @@ export Patch, # types
 const boltz = 1.38064852e-23 # J/K = m2⋅kg/(s2⋅K)
 const act = 1e-19 # activation energy /J, ca. 0.63eV - Brown et al. 2004
 const growthrate = exp(25.2) # global base growth/biomass production from Brown et al. 2004
-const mortality = exp(20) # global base mortality from Brown et al. 2004 is 26.3, but competition introduces add. mort.
+const mortality = exp(20) # global base mortality from Brown et al. 2004 is 26.3, but competition and dispersal introduce add. mort.
 const fertility = exp(23.8) # global base reproduction rate from Brown et al. 2004, alternatively 25.0
 const phylconstr = 50 #parse(ARGS[2])
 # const meangenes = 20 # mean number of genes per individual

### 2018-01-16 paper/draft.tex
checkout changes from parallel

@@
 %% [NEW ANGLE:]
-Island biota have always fascinated and inspired researchers (Darwin, Wallace, ...). %TODO: refs
-Nowadays islands are understood as well suitable model systems for ecology and evolution due to their well definable characteristics like isolation. %TODO: ref
-However island biota are also increasingly at risk.
+% Island biota have always fascinated and inspired researchers (Darwin, Wallace, ...). %TODO: refs
+Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation. %TODO: ref
+Due to global change, however, island biota are also increasingly at risk.
 While often featuring unique species communities and harboring countless endemics, %TODO: refs
@@
 Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al. 2017).
-A subset of these questions deals with 
+A subset of these questions deals with the topic of speciation and radiations on islands.
+Still, the proposed mechanisms so far include only ecological or geo-morphological processes,
+while genetic properties are hardly recognized in the island context.
+At the same time, an increasing number of studies investigates genetic processes and
+their effect on reproductive/ecological isolation of populations (e.g., Via).
+
 One of the first and to this day highly influential formal theories of island biodiversity is MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB). %TODO: ref

### 2018-01-16 paper/draft.tex
add to MM
MM edit

@@
 At the same time, an increasing number of studies investigates genetic processes and
-their effect on reproductive/ecological isolation of populations (e.g., Via).
+their effect on reproductive/ecological isolation of populations (e.g., Via & West 2008, Via 2001 and references therein).
 
@@
 establishment, competition, growth, (density independent) mortality, reproduction and dispersal.
-Each of which processes is affected by the individuals' characteristics.
+Each of which process is affected by the individuals' trait characteristics.
 The simulation arena consists of a larger landmass (the mainland) holding an initial pool of evolving populations and a smaller, initially unpopulated island some distance apart.
 During the course of the simulation, individuals may disperse from the mainland to the island, thus colonizing it.
+The complete state of the simulation is recorded every 1000 time steps for analysis.
 
@@
 \Cref{tab:scenarios} summarizes the parameters which were varied for the scenarios, their meaning and their values. %TODO: make table.
-For each parameter combination, we launched a number of %TODO: how much?
-replicates.
+Each parameter combination replicated %TODO: how much?
+times.
 

### 2018-02-19 paper/draft.tex
add citation
lay out figure dummies: linkage units v species and trait space

@@
 % Island biota have always fascinated and inspired researchers (Darwin, Wallace, ...). %TODO: refs
-Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation. %TODO: ref
+Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation. (e.g., Losos & Ricklefs 2009) %TODO: ref
 Due to global change, however, island biota are also increasingly at risk.
@@
 Using the thus simulated data, we calculate a phylogeny of the extant species, species numbers, etc. [WHAT ELSE?]
+
 \section{Results}
+
+\begin{figure}
+  \includegraphics{}
+  \caption{Number of linkage units and number of species per lineage XXX timesteps after colonization.
+    [Histogram (with threshold category: strongly diversified lineages) or scatter plot.]}
+  \label{specieslinkage}
+\end{figure}
+
+\begin{figure}
+  \includegraphics{}
+  \caption{Trait space characteristics of lineages.
+    (A) [Ordination] Trait space of successful colonizers compared to the mainland species pool. Each point represents a lineage.
+    Arrows denote the traits.
+    (B) [Ordination?] Linkage of traits of successful colonizers(/radiators?). Traits perpendicular to each other are inherited independently,
+  same direction indicate tight linkage of traits.}
+  \label{traitspace}
+\end{figure}
 

### 2018-02-19 paper/draft.tex
some updates and minor edits

@@
 Each of which process is affected by the individuals' trait characteristics.
-The simulation arena consists of a larger landmass (the mainland) holding an initial pool of evolving populations and a smaller, initially unpopulated island some distance apart.
+The simulation arena consists of a larger landmass (the mainland) holding an initial random species pool and a smaller, initially unpopulated island some distance apart.
 During the course of the simulation, individuals may disperse from the mainland to the island, thus colonizing it.
@@
 (a) gene linkage,
-(b) reproductive isolation,
+(b) reproductive isolation (or reproduction compatibility),
 (c) habitat diversity, and
@@
 \Cref{tab:scenarios} summarizes the parameters which were varied for the scenarios, their meaning and their values. %TODO: make table.
-Each parameter combination replicated %TODO: how much?
-times.
+Each parameter combination was replicated %TODO: how much?
+times, setting a different random seed for each replicate.
 
 \subsection{Data recording}
-We record the complete state of our simulation world at the end of a simulation run and each time the arena changes (geomorphological dynamics).
+We record the complete state of our simulation world at the start and end of a simulation run and each time the arena changes (geomorphological dynamics).
 Additionally, we store the state of the simulation arena each 1000 %TODO: how much exactly?
-time steps and record every individual that colonizes the island to track it in the phylogeny.
+time steps and record every individual that colonizes the island to track changes.
 
-Using the thus simulated data, we calculate a phylogeny of the extant species, species numbers, etc. [WHAT ELSE?]
+Using the thus simulated data, we calculate a phylogenies of the colonizing lineages, species numbers, trait spaces etc. [WHAT ELSE?]
 

### 2018-02-19 paper/draft.tex
fixes: load necessary packages etc.

@@
 
+\usepackage{graphicx}
 \usepackage[T1]{fontenc}
@@
 \usepackage{csquotes}
+\usepackage{cleveref}
 
@@
 %
+
 \begin{document}
@@
 % Island biota have always fascinated and inspired researchers (Darwin, Wallace, ...). %TODO: refs
-Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation. (e.g., Losos & Ricklefs 2009) %TODO: ref
+Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation. (e.g., Losos \& Ricklefs 2009) %TODO: ref
 Due to global change, however, island biota are also increasingly at risk.
@@
 At the same time, an increasing number of studies investigates genetic processes and
-their effect on reproductive/ecological isolation of populations (e.g., Via & West 2008, Via 2001 and references therein).
+their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein).
 
@@
 to asses their relative contribution on community composition.
-\Cref{tab:scenarios} summarizes the parameters which were varied for the scenarios, their meaning and their values. %TODO: make table.
+\cref{tab:scenarios} summarizes the parameters which were varied for the scenarios, their meaning and their values. %TODO: make table.
 Each parameter combination was replicated %TODO: how much?
@@
 \begin{figure}
-  \includegraphics{}
+  %\includegraphics{}
   \caption{Number of linkage units and number of species per lineage XXX timesteps after colonization.
@@
 \begin{figure}
-  \includegraphics{}
+  %\includegraphics{}
   \caption{Trait space characteristics of lineages.

### 2018-02-19 paper/draft.tex
restructure and update intro

@@
 Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation. (e.g., Losos \& Ricklefs 2009) %TODO: ref
-Due to global change, however, island biota are also increasingly at risk.
 While often featuring unique species communities and harboring countless endemics, %TODO: refs
@@
 This makes understanding the functioning and evolution of island communities a high priority objective.
-
-Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al. 2017).
-A subset of these questions deals with the topic of speciation and radiations on islands.
-Still, the proposed mechanisms so far include only ecological or geo-morphological processes,
-while genetic properties are hardly recognized in the island context.
-At the same time, an increasing number of studies investigates genetic processes and
-their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein).
 
@@
 In doing so, the model incorporates concepts of niche theory, speciation and species succession.
-Yet, there are still many processes and drivers deemed important, which are not considered by any formal theory.
+Yet, there are still many processes and drivers deemed important, which are not considered by any formal theory and often little investigated.
+
+Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al. 2017).
+A subset of these questions deals with the topic of speciation and radiations on islands.
+Still, the proposed mechanisms so far include only ecological or geo-morphological processes,
+while genetic properties are hardly recognized in the island context.
+At the same time, an increasing number of studies investigates genetic processes and
+their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein).
+
+The problem of investigating these processes is the time frame and study extent needed for a proper evaluation.
+Because islands biogeography takes into account evolutionary time frames (millions of years),
+comprehensive and meaningful data is difficult to obtain.
+
+In a recent review \cite{leidinger2017biodiversity}, we analyzed a powerful alternative to field experiments in island biogeography: mechanistic (/simulation) models.
+Although many of the reviewed models integrate multiple processes at a time, unfortunately, most models failed to include more explicit genetic or micro evolution processes.
+Instead, some models used concepts such as ``protracted speciaton'' \cite{rosindellXXX},
+where speciation occurs with a delay from the actual
+divergence event if the population remains isolated during that time.
+But nature is usually more complex than this and genetic mechanisms have important ramifications for speciation/radiation and species richness.
 
@@
 
-The problem of investigating these processes is the time frame needed for a proper evaluation.
-Because islands biogeography takes into account evolutionary time frames (millions of years),
-comprehensive and meaningful data is difficult to obtain.
-In a recent review \cite{leidinger2017biodiversity}, we analyzed a powerful alternative to field experiments in island biogeography
---- mechanistic (/simulation) models.
-Unfortunately, most island models failed to include more explicit genetic or micro evolution processes.
-% reasons: underlying theories: Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al. 2017).
-%A subset of these questions deals with 
-Instead, some models featured a concept known as ``protracted speciaton'' \cite{rosindellXXX}, where speciation occurs with a delay from the actual
-divergence event if the population remains isolated during that time.
-But nature is usually more complex than this and genetic mechanisms have important ramifications for speciation/radiation and species richness.
-[EXAMPLES!]
-
 For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
@@
 
-Another shortfall of the reviewd models was the lack of integrating ecological niches and adaption.
+Another shortfall of the reviewed models was the lack of integrating ecological niches and adaption.
 
@@
 
-In the following, we present a model that remedies both of the adressed shortcomings by implementing genetic processes on a niche-explicit background
+In the following, we present a model that remedies both of the adressed shortcomings
+by implementing genetic processes on a niche-explicit background
 using an island biogeography framework.
-We will adress the questions of what relative impact does (a) genetic linkage and (b) reproductive isolation have on emerging species numbers and diversity,
-and how (c) landscape heterogeneity (habitat niche diversity) and (d) topographical complexity differently affect radiatons.
-[further questions thus tackled: traits of colonizer species, ...]
+We will adress the questions of what relative impact does (a) genetic linkage and
+(b) reproductive isolation have on emerging species numbers and diversity,
+and (c) which genetic and functional trait characteristics distinguish successful colonizer,
+strong competitor and diversifying lineages.
 

### 2018-02-20 paper/draft.tex
restructure intro

@@
 
-%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
-%% island colonisation, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
-
-%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
-%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs.
-phylogenetic similarity \cite{losos2008phylogenetic}
-
 %% These patterns have been explained theoretically [here, here and here], but usually one at a time.
@@
 
-%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
-%% and what trait combinations constitute successful colonizer and radiating species, respectively.
-
 %% [NEW ANGLE:]
-% Island biota have always fascinated and inspired researchers (Darwin, Wallace, ...). %TODO: refs
 Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation. (e.g., Losos \& Ricklefs 2009) %TODO: ref
@@
 A subset of these questions deals with the topic of speciation and radiations on islands.
+%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
+%% island colonisation, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
+
+%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
+%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs.
+phylogenetic similarity \cite{losos2008phylogenetic}
 Still, the proposed mechanisms so far include only ecological or geo-morphological processes,
@@
 
-In the following, we present a model that remedies both of the adressed shortcomings
+In the following, we present a mechanistic model that remedies both of the adressed shortcomings
 by implementing genetic processes on a niche-explicit background
-using an island biogeography framework.
+using an island biogeography framework.%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
+%% and what trait combinations constitute successful colonizer and radiating species, respectively.
 We will adress the questions of what relative impact does (a) genetic linkage and

### 2018-02-20 paper/draft.tex
add some refs to commented text

@@
 A subset of these questions deals with the topic of speciation and radiations on islands.
-%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
-%% island colonisation, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
+%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact of explosively radiating lineages (Hughes \& Eastwood 2006, Kocher 2004)
+%%, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
 
@@
 phylogenetic similarity \cite{losos2008phylogenetic}
-Still, the proposed mechanisms so far include only ecological or geo-morphological processes,
+Still, the proposed mechanisms so far include only ecological or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
 while genetic properties are hardly recognized in the island context.

### 2018-02-20 paper/draft.tex
add some refs in intro, extend disc

@@
 While often featuring unique species communities and harboring countless endemics, %TODO: refs
-islands are especially threatened by climate change, human impact and invasive species. %TODO: refs
+islands are especially threatened by climate change, human impact and invasive species (Fordham \& Brook 2010). %TODO: refs
 This makes understanding the functioning and evolution of island communities a high priority objective.
@@
 
+In this work we concentrated on genetic influences on speciation.
+However, there is ongoing debate on how strongly genetic factors also impact extinction processes (Jamieson 2007).
+Seeing that observed speciation rates are in fact net speciation rates highlights the need to mechanistically explore
+these processes as well.
+In fact, genetic processes might play an important role in the issue of extinction debt. %% cite Ludmilla, ask for refs
+
 \printbibliography

### 2018-02-20 paper/draft.tex
comment
extend M&M

@@
 
+Building on the model of Cabral et al. (2017a, b), most of the biological rates are controlled following the metabolic theory
+of ecology (MTE, Brown et al.\ 2004).
+In MTE, a biological rate $b$ depends on the temperature $T$ and individual mass $M$, scaling a base rate $b_0$ as:
+\begin{equation}
+  $b=b_0*M^{3/4}*exp(-E_A/{k_B*T})$
+  \label{eq:MTE}
+\end{equation}
+with $E_A$ the activation energy and $k_B$ the Boltzman constant.
+
 A detailed protocol can be found in the appendix (\cite{grimm2010odd}, XXX).
@@
 Especially pollination is assumed to be an important interaction, because having a high impact on evolution and establishment of species due to specialization.
-[maybe some words on trophic interactions.]
+%[maybe some words on trophic interactions.]
 

### 2018-02-20 paper/draft.tex
fix
fix equation
add ref against protracted spec. to intro

@@
 
-Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al. 2017).
+Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al.\ 2017).
 A subset of these questions deals with the topic of speciation and radiations on islands.
@@
 divergence event if the population remains isolated during that time.
-But nature is usually more complex than this and genetic mechanisms have important ramifications for speciation/radiation and species richness.
+But nature is usually more complex than this (cf.\ Herben et al.\ 2005) and genetic mechanisms can have important ramifications for speciation/radiation and species richness.
 
@@
 \begin{equation}
-  $b=b_0*M^{3/4}*exp(-E_A/{k_B*T})$
+  b=b_0*M^{3\over4}*e^{-{E_A}\over{k_B*T}}
   \label{eq:MTE}

### 2018-02-20 paper/draft.tex
add a lot of refs

@@
 In it, species richness on islands is the result of opposing rates of immigration and extinction, both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
-This relationship was since successfully applied to a multitude of islands %TODO: refs
-, but also island like systems, such as coral reefs, lakes/ponds or forest fragments. %TODO: refs
-This general applicability, however, cannot hide the fact that the ETIB fails to deliver a mechanistic explanation of the processes at play.
-It has particularly been criticized for its lack of considering speciation. %TODO: refs
+This relationship was since successfully applied to a multitude of islands (Lomolina \& Weiser 2001, Triantis et al.\ 2011)%TODO: refs
+, but also island like systems, such as coral reefs, lakes/ponds or forest fragments (Kelly et al.\ 1989, Peay et al.\ 2007, ). %TODO: refs
+This general applicability (Lomolino 2000), however, cannot hide the fact that the ETIB fails to deliver a mechanistic explanation of the processes at play.
+It has particularly been criticized for its lack of considering speciation (see e.g.\ Losos \& Schluter 2000). %TODO: refs
 
-About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM). %TODO: ref
+About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM, Whittaker et al.\ 2007, 2008). %TODO: ref
 The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
@@
 divergence event if the population remains isolated during that time.
-But nature is usually more complex than this (cf.\ Herben et al.\ 2005) and genetic mechanisms can have important ramifications for speciation/radiation and species richness.
+But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005) and genetic mechanisms can have important ramifications for speciation/radiation and species richness.
 
@@
 For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
-seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al. 2013) % cite properly!
+seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al.\ 2013) % cite properly!
 

### 2018-02-20 paper/draft.tex
restructure flow of intro

@@
 
+The problem of investigating these processes is the time frame and study extent needed for a proper evaluation.
+Because islands biogeography takes into account evolutionary time frames (millions of years),
+comprehensive and meaningful data is difficult to obtain.
+
+In a recent review %\cite{leidinger2017biodiversity}
+(Leidinger \& Sarmento Cabral 2017), we analyzed a powerful alternative to field experiments in island biogeography: mechanistic (/simulation) models.
+Although many of the reviewed models integrate multiple processes at a time, unfortunately, most models failed to include more explicit genetic or micro evolution processes.
+Instead, some models used concepts such as ``protracted speciaton'' (Rosindell \& Harmon 2011) %\cite{rosindellXXX},
+where speciation occurs with a delay from the actual
+divergence event if the population remains isolated during that time.
+But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005) and genetic mechanisms can have important ramifications for speciation/radiation and species richness.
+
 Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al.\ 2017).
@@
 their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein).
-
-The problem of investigating these processes is the time frame and study extent needed for a proper evaluation.
-Because islands biogeography takes into account evolutionary time frames (millions of years),
-comprehensive and meaningful data is difficult to obtain.
-
-In a recent review \cite{leidinger2017biodiversity}, we analyzed a powerful alternative to field experiments in island biogeography: mechanistic (/simulation) models.
-Although many of the reviewed models integrate multiple processes at a time, unfortunately, most models failed to include more explicit genetic or micro evolution processes.
-Instead, some models used concepts such as ``protracted speciaton'' \cite{rosindellXXX},
-where speciation occurs with a delay from the actual
-divergence event if the population remains isolated during that time.
-But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005) and genetic mechanisms can have important ramifications for speciation/radiation and species richness.
-
 %%TODO: here come a list of processes with examples
@@
 %% - isolation processes
+Besides arthropods (Howthorne \& Via 2001), these processes can be observed in plants as well.
+For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
+seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al.\ 2013). % cite properly!
+Considering these phenomena, we expect speciation rates to be highest at intermediate levels of gene linkage --- low
+levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption.
 
-For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
-seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al.\ 2013) % cite properly!
+%Another shortfall of the reviewed models was the lack of integrating ecological niches and adaption.
 
-Another shortfall of the reviewed models was the lack of integrating ecological niches and adaption.
+%Besides adaptive radiations(/evolution), changes in or different measures of geographic isolation (landscape complexity) can also give rise to non-adaptive radiations.
 
-Besides adaptive radiations(/evolution), changes in or different measures of geographic isolation (landscape complexity) can also give rise to non-adaptive radiations.
-
-In the following, we present a mechanistic model that remedies both of the adressed shortcomings
+To address this hypothesis, we present a mechanistic model that tries to remedy above mentioned shortcomings
 by implementing genetic processes on a niche-explicit background
@@
 \section{Material and Methods}
-We use an individual based metacommunity model written in julia \cite{JULIALANG}.
+We use an individual based metacommunity model written in julia %\cite{JULIALANG}.
 Each individual is characterized by a set of genetically controlled traits and is submitted to processes of

### 2018-02-20 paper/draft.tex
update abstract

@@
   Islands feature peculiar and fascinating biotas, often very distinct from any other island or continental area.
-  [Examples of radiations/good and poor radiators, island syndromes etc.]
   Although island biogeography theory has seen an ever increasing development over the years,
@@
   Specifically, some processes remain un- or under-explored, foremost of these adaption and micro-evolution.
-  We used an individual-based mechanistic model implementing these processes and other commonly investigated ecological and
+  We use an individual-based mechanistic model implementing genetic and micro-evolutionary processes and other commonly investigated ecological and
   environmental processes to assess the importance of drivers across different organisational and spatio-temporal levels
   on species trait distribution on islands.
-  This enabled us to also take a closer look on the properties of lineages/species that actually managed to colonize islands,
-  and which of them resulted in radiations[~].
+  Specifically, we address the question why some lineages diverge stronger than others.
+  We found that lineages with intermediate levels of gene linkage evolve into a higher number of distinct populations
+  than low or high gene linkage lineages.
+  The model enabled us to also take a closer look on the properties of lineages/species that actually managed to colonize islands,
+  and which of them resulted in radiations.
 
@@
   \caption{Number of linkage units and number of species per lineage XXX timesteps after colonization.
-    [Histogram (with threshold category: strongly diversified lineages) or scatter plot.]}
+    [Histogram (with threshold category: strongly diversified lineages) or scatter plot. Two panels/color codes: high and low degree of environmental heterogeneity]}
   \label{specieslinkage}

### 2018-02-22 MIbGxMCmod.jl
fix gausscurve NaN results when c=0

@@ -209,8 +209,14 @@ function traitsexist(ind::Individual, traitnames::Array{String, 1})
 end
 
 function gausscurve(b::Float64, c::Float64, x::Float64, a::Float64=1.0)
-    a != 1.0 && (a = 1 / (c * sqrt(2 * pi)))
-    y = a * exp(-(x-b)^2/(2*c^2))
+    if c != 0 && a != 1.0
+        a = 1 / (c * sqrt(2 * pi))
+        y = a * exp(-(x-b)^2/(2*c^2))
+    elseif c != 0
+        y = a * exp(-(x-b)^2/(2*c^2))
+    else
+        y = 0
+    end
 end
 
 """

### 2018-02-22 MIbGxMCmod.jl
fix establishment set fitness to 1 for old individuals

@@ -228,33 +228,38 @@ function establish!(patch::Patch, nniches::Int64=1)
     temp = patch.altitude
     idx = 1
     while idx <= size(patch.community,1)
-        fitness = 1
-        if !traitsexist(patch.community[idx], ["temptol", "tempopt"]) # CAVE: should check for more traits!
-            splice!(patch.community, idx) # kill it!
-            continue
-        elseif patch.community[idx].isnew || patch.community[idx].age == 0
+        if patch.community[idx].isnew || patch.community[idx].age == 0
+            fitness = 1
+            if !traitsexist(patch.community[idx], ["temptol", "tempopt"]) # CAVE: should check for more traits!
+                splice!(patch.community, idx) # kill it!
+                continue
+            end
             opt = patch.community[idx].traits["tempopt"]
             tol = patch.community[idx].traits["temptol"]
             fitness *= gausscurve(opt, tol, temp, 0.0)
             fitness > 1 && (fitness = 1) # should be obsolete
             fitness < 0 && (fitness = 0) # should be obsolete
+            if nniches >= 2
+                if !traitsexist(patch.community[idx], ["prectol", "precopt"]) # CAVE: should check for more traits!
+                    splice!(patch.community, idx) # kill it!
+                    continue
+                end
+                opt = patch.community[idx].traits["precopt"]
+                tol = patch.community[idx].traits["prectol"]
+                fitness *= gausscurve(opt, tol, patch.nichea, 0.0)
+                fitness > 1 && (fitness = 1) # should be obsolete
+                fitness < 0 && (fitness = 0) # should be obsolete
+            end
+            if nniches == 3
+                opt = patch.community[idx].traits["nicheopt"]
+                tol = patch.community[idx].traits["nichetol"]
+                fitness *= gausscurve(opt, tol, patch.nicheb, 0.0)
+                fitness > 1 && (fitness = 1) # should be obsolete
+                fitness < 0 && (fitness = 0) # should be obsolete
+            end  
+            patch.community[idx].fitness = fitness
+            patch.community[idx].isnew = false
         end
-        if nniches >= 2 && (patch.community[idx].isnew || patch.community[idx].age == 0)
-            opt = patch.community[idx].traits["precopt"]
-            tol = patch.community[idx].traits["prectol"]
-            fitness *= gausscurve(opt, tol, patch.nichea, 0.0)
-            fitness > 1 && (fitness = 1) # should be obsolete
[…]

### 2018-02-22 MIbGxMCmod.jl
make competition random

@@ -457,7 +457,7 @@ end
 function compete!(patch::Patch)
     sort!(patch.community, by = x -> x.fitness)
     while sum(map(x -> x.size, patch.community)) >= patch.area # occupied area larger than available
-        victim = rand(Geometric()) + 1
+        victim = rand(1:length(patch.community)) # rand(Geometric()) + 1 # fitness sorted?
         victim > length(patch.community) && (victim = length(patch.community))
         splice!(patch.community, victim)
     end

### 2018-02-22 paper/draft.tex
add dummy figure, comment in disc
ecomod meeting

@@
 
-\title{The relative roles of genetic, ecological and environmental factors in species radiations}
+\title{The relative roles of genetic, ecological and environmental factors %/processes?
+  in insular species radiations}
 \author{Ludwig Leidinger}
@@
 divergence event if the population remains isolated during that time.
-But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005) and genetic mechanisms can have important ramifications for speciation/radiation and species richness.
-
+But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005) and genetic mechanisms can have important implications for speciation/radiation and species richness.
+%as a matter of fact, most explanations still ecological/environmental (Pati)
 Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al.\ 2017).
@@
 Considering these phenomena, we expect speciation rates to be highest at intermediate levels of gene linkage --- low
-levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption.
+levels leads to dissolving phenotype populations %explain!
+, whereas high levels hinder rapid adaption.
 
@@
 by implementing genetic processes on a niche-explicit background
-using an island biogeography framework.%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
+using an island biogeography framework.
+%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
 %% and what trait combinations constitute successful colonizer and radiating species, respectively.
@@
 
+\begin{figure}
+  %\includegraphics{}
+  \caption{Exemplary phylogenies of lineages without co-occurring competing lineages calculated from neutral sequences of
+    extant individuals.
+    (A) Lineage with low gene linkage,
+    (B) lineage with intermediate gene linkage,
+    (C) lineage with high gene linkage.}
+  \label{phylogenies}
+\end{figure}
+
 \section{Discussion}
@@
 recombination between populations to speciation (Via 2011, but see Feder \& Nosil 2010).
-During incipient speciation, however, th
+%During incipient speciation, however, th
 

### 2018-04-26 src/types.jl
add a number to chromosome

@@ -7,6 +7,7 @@ end
 
 mutable struct Chromosome
     genes::Array{Gene, 1}
+    number::int64
     maternal::Bool # parental origin of chromosome
 end
 

### 2018-04-26 src/subprocs.jl
use chromosome number

@@ -17,7 +17,7 @@ function meiosis(genome::Array{Chromosome,1}, maternal::Bool, mtraits::Dict{Stri
         else
             append!(ptraitnames, vcat(map(x -> x.codes, genome[i].genes)...))
         end
-        push!(gamete, Chromosome(genome[i].genes, maternal))
+        push!(gamete, Chromosome(genome[i].genes, genome[i].number, maternal))
     end
     mtraitnames = unique(mtraitnames)
     ptraitnames = unique(ptraitnames)
@@ -291,15 +291,15 @@ function createchrs(nchrs::Int64,genes::Array{Gene,1})
         chromosomes = Chromosome[]
         for chr in 1:nchrs
             if chr==1 # first chromosome
-                push!(chromosomes, Chromosome(genes[1:chrsplits[chr]], true))
+                push!(chromosomes, Chromosome(genes[1:chrsplits[chr]], chr, true))
             elseif chr==nchrs # last chromosome
-                push!(chromosomes, Chromosome(genes[(chrsplits[chr-1]+1):end], true))
+                push!(chromosomes, Chromosome(genes[(chrsplits[chr-1]+1):end], chr, true))
             else
-                push!(chromosomes, Chromosome(genes[(chrsplits[chr-1]+1):chrsplits[chr]], true))
+                push!(chromosomes, Chromosome(genes[(chrsplits[chr-1]+1):chrsplits[chr]], chr, true))
             end
         end
     else # only one chromosome
-        chromosomes = [Chromosome(genes, true)]
+        chromosomes = [Chromosome(genes, 1, true)]
     end
     secondset = deepcopy(chromosomes)
     for chrm in secondset

### 2018-04-26 src/subprocs.jl
use chromosome numbers to sort genome prior meiosis

@@ -1,6 +1,7 @@
 # Subsidiary functions for GeMM
 
 function meiosis(genome::Array{Chromosome,1}, maternal::Bool, mtraits::Dict{String,Float64}, ptraits::Dict{String,Float64}) # TODO: include further dynamics, errors...
+    sort!(genome, by = x -> x.number)
     firstset = find(x->x.maternal,genome) # CAVE: chromosomes could still be mismatched!
     secondset = find(x->!x.maternal,genome)
     length(firstset) != length(secondset) && return (Chromosome[], mtraits) # CAVE: more elegant solution...

### 2018-04-26 src/types.jl
fix typo

@@ -7,7 +7,7 @@ end
 
 mutable struct Chromosome
     genes::Array{Gene, 1}
-    number::int64
+    number::Int64
     maternal::Bool # parental origin of chromosome
 end
 

### 2018-04-26 src/subprocs.jl
fixes to mate filtering, break after one mate is found

@@ -171,11 +171,12 @@ function findposspartners(world::Array{Patch,1}, ind::Individual, location::Tupl
         for mate in community
             mate.age == 0 && continue
             mate.lineage != ind.lineage && continue
-            traitsexist(mate, ["repsize"]) && continue
-            mate.size >= mate.traits["repsize"] && continue
+            !traitsexist(mate, ["repsize"]) && continue
+            mate.size < mate.traits["repsize"] && continue
             mate.isnew && continue
             !iscompatible(mate, ind) && continue
-            append!(posspartners, mate)
+            push!(posspartners, mate)
+            length(posspartners) >= 1 && break
         end
         idx += 1
     end

### 2018-04-26 src/subprocs.jl
can compare lengths of dicts directly

@@ -39,7 +39,7 @@ function chrms2traits(mtraits::Dict{String,Float64}, ptraits::Dict{String,Float6
             traitdict[traitname] = mean([mtraits[traitname], ptraits[traitname]])
         end
     else
-        length(keys(mtraits)) >= length(keys(ptraits)) ? traitdict = mtraits : traitdict = ptraits
+        length(mtraits) >= length(ptraits) ? traitdict = mtraits : traitdict = ptraits
         warn(STDERR, "parental traits do not match. Returning homozygote.")
     end
     traitdict

### 2018-04-26 paper/source.tex
Created TODO list and some other support files.
Namely: .gitignore, source.tex and run.sh

%% LyX 2.1.4 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}

\makeatletter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage[T1]{fontenc}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{xcolor}

%%
%% Julia definition (c) 2014 Jubobs
%%
\lstdefinelanguage{Julia}%
  {morekeywords={abstract,break,case,catch,const,continue,do,else,elseif,%
      end,export,false,for,function,immutable,import,importall,if,in,%
      macro,module,otherwise,quote,return,switch,true,try,type,typealias,%
      using,while},%
   sensitive=true,%
   alsoother={$},%
   morecomment=[l]\#,%
   morecomment=[n]{\#=}{=\#},%
   morestring=[s]{"}{"},%
   morestring=[m]{'}{'},%
}[keywords,comments,strings]%

\lstset{%
    language=julia,
    breaklines=true,
    frame=lines,
    basicstyle={\tiny\ttfamily},
    numbers=left,
    stepnumber=10,
    numberstyle={\tiny},
    tabsize=4,
    commentstyle={\color{gray}\itshape},
    keywordstyle={\color{blue}},
    stringstyle={\color{teal}},
    showstringspaces=false
}

\makeatother

\usepackage{babel}
\usepackage{listings}
\begin{document}

\title{Island Speciation Model Source Code}

\author{Ludwig Leidinger, Daniel Vedder}

\maketitle
\lstinputlisting[caption={islandsim.jl}]{../islandsim.jl}

\lstinputlisting[caption={extractinfo.jl}]{../extractinfo.jl}

\lstinputlisting[caption={getdata.jl}]{../getdata.jl}

\lstinputlisting[caption={GeMM.jl}]{../src/GeMM.jl}

\lstinputlisting[caption={types.jl}]{../src/types.jl}

\lstinputlisting[caption={init.jl}]{../src/init.jl}

\lstinputlisting[caption={mainprocs.jl}]{../src/mainprocs.jl}

\lstinputlisting[caption={subprocs.jl}]{../src/subprocs.jl}

\lstinputlisting[caption={io.jl}]{../src/io.jl}
\end{document}

### 2018-04-26 run.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+
+echo "Starting simulation run."
+julia islandsim.jl --maps invasion.map --linkage "none" --nniches 2 --tolerance "none" --dest tests
+echo "Done."

### 2018-04-26 src/init.jl
remove unnecessary trait calculation

@@ -21,8 +21,8 @@ function genesis(settings::Dict{String,Any},
         meangenes = length(traitnames)
         ngenes = rand(Poisson(meangenes))
         ngenes < 1 && (ngenes = 1)
-        traits = createtraits(traitnames, settings)
-        genes = creategenes(ngenes,traits)
+        traitdict = createtraits(traitnames, settings)
+        genes = creategenes(ngenes,traitdict)
         if settings["linkage"] == "none"
             nchrms = length(genes)
         elseif settings["linkage"] == "full"
@@ -31,7 +31,6 @@ function genesis(settings::Dict{String,Any},
             nchrms = rand(1:length(genes))
         end
         chromosomes = createchrs(nchrms,genes)
-        traitdict = chrms2traits(traits, traits)
         popsize = round(fertility * traitdict["repsize"]^(-1/4) * exp(-act/(boltz*traitdict["tempopt"]))) # population size determined by adult size and temperature niche optimum
         popmass = popsize * traitdict["seedsize"]
         if totalmass + popmass > nspecs # stop loop if cell is full

### 2018-05-14 paper/draft.tex
merge JSC revisions from Feb

--- 2018-02-20 paper/draft.tex
@@
 
-\title{The relative roles of genetic, ecological and environmental factors in species radiations}
+\title{The relative roles of genetic, ecological and environmental factors in insular species radiations}
 \author{Ludwig Leidinger}
@@
 \begin{abstract}
-  Islands feature peculiar and fascinating biotas, often very distinct from any other island or continental area.
+  Islands are known for their peculiar and fascinating biotas, typically very distinct from any other landmasses, be it neighboring islands or continental areas.
   Although island biogeography theory has seen an ever increasing development over the years,
@@
 %% [NEW ANGLE:]
-Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation. (e.g., Losos \& Ricklefs 2009) %TODO: ref
-While often featuring unique species communities and harboring countless endemics, %TODO: refs
-islands are especially threatened by climate change, human impact and invasive species (Fordham \& Brook 2010). %TODO: refs
-This makes understanding the functioning and evolution of island communities a high priority objective.
+Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation (e.g.
+Losos \& Ricklefs 2009; Warren et al.
+2015  Islands as model systems in ecology and evolution: prospects fifty years after MacArthur‐Wilson.
+EcoLett).
+%TODO: ref
+Part of this value as study system originates by typically remarkable species.
+In fact, islands often possess unique species communities and can exceede mainland areas by a factor of 9.5 and 8.1 in plant and animal  endemism, respectively (Kier et al.
+2005.
+A global assessment of endemism and species richness across island and mainland regions.
+PNAS).
+%TODO: refs
+Such high endemism makes islands  especially threatened by human impacts, in particular climate change and invasive species (Fordham \& Brook 2010). %TODO: refs
+Hence, understanding the origin, maintenance, functioning and evolution of island communities is of a high priority objective.
 
-One of the first and to this day highly influential formal theories of island biodiversity is MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB). %TODO: ref
+For the prediction of island community size, several theoretical frameworks have been proposed, including one of the first and to this day still highly influential classical ecological theories, namely MacArthur and Wilson's equilibrium theory of island biogeography (ETIB).
+%TODO: ref
 In it, species richness on islands is the result of opposing rates of immigration and extinction, both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
@@
 , but also island like systems, such as coral reefs, lakes/ponds or forest fragments (Kelly et al.\ 1989, Peay et al.\ 2007, ). %TODO: refs
-This general applicability (Lomolino 2000), however, cannot hide the fact that the ETIB fails to deliver a mechanistic explanation of the processes at play.
-It has particularly been criticized for its lack of considering speciation (see e.g.\ Losos \& Schluter 2000). %TODO: refs
+This general applicability (Lomolino 2000), however, cannot hide the fact that the ETIB fails to formalize a mechanistic explanation of how extinction and immigration are modified by area and isolation, respectively.
+Additionally, it has been demonstrated more recently that area also affects immigration via target effects (REFs) and isolation affects extinction via rescue-effects (REFs).
+Finally, ETIB has particularly been criticized for its lack of considering speciation (see e.g.\ Losos \& Schluter 2000).
+%TODO: refs
 
@@
 The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
-whith islands emerging from sea level, increasing over time in size and topographic complexity, before eroding and subsiding in the ocean again.
+whith islands emerging from sea level via volcanic hotspot activity, increasing over time in size and topographic complexity, then eroding and subsiding in the ocean again when the island is not over the hotspot anymore.
 In doing so, the model incorporates concepts of niche theory, speciation and species succession.
-Yet, there are still many processes and drivers deemed important, which are not considered by any formal theory and often little investigated.
+Yet, there are still many processes and drivers deemed important, which are not considered by any formal theory and often little investigated [if you say this, you have to mention at leastr a couple os examples, otherwise it is an empty claim].
 
 The problem of investigating these processes is the time frame and study extent needed for a proper evaluation.
-Because islands biogeography takes into account evolutionary time frames (millions of years),
-comprehensive and meaningful data is difficult to obtain.
-
-In a recent review %\cite{leidinger2017biodiversity}
-(Leidinger \& Sarmento Cabral 2017), we analyzed a powerful alternative to field experiments in island biogeography: mechanistic (/simulation) models.
-Although many of the reviewed models integrate multiple processes at a time, unfortunately, most models failed to include more explicit genetic or micro evolution processes.
-Instead, some models used concepts such as ``protracted speciaton'' (Rosindell \& Harmon 2011) %\cite{rosindellXXX},
+Because islands biogeography takes place at evolutionary time frames (millions of years),
+comprehensive and meaningful data is difficult and virtually impossible to obtain.
+Therefore, there has been an interest in building predictive models for islands.
+In a recent review, we indeed found a rapid increase in number of mechanistic simulaiton models for island biogeography %\cite{leidinger2017biodiversity}
+(Leidinger \& Sarmento Cabral 2017).
+Although many of the reviewed models integrate multiple processes at a time, unfortunately, most models failed to include microevolutionary or more explicit genetic processes %\cite{leidinger2017biodiversity}
+(Leidinger \& Sarmento Cabral 2017).
+One direction towards this is given by models using concepts such as ``protracted speciaton'' (Rosindell \& Harmon 2011) %\cite{rosindellXXX},
 where speciation occurs with a delay from the actual
 divergence event if the population remains isolated during that time.
-But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005) and genetic mechanisms can have important ramifications for speciation/radiation and species richness.
-
-Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al.\ 2017).
-A subset of these questions deals with the topic of speciation and radiations on islands.
+But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005) and genetic mechanisms should also play a central role for speciation/radiation and species richness.
+Questions, such as "[cite two questions from Patino et al.
+2017]", thus remain largely open (Patiño et al.\ 2017) and may require explicit consideration of evolutionary processes.
 %% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact of explosively radiating lineages (Hughes \& Eastwood 2006, Kocher 2004)
@@
 phylogenetic similarity \cite{losos2008phylogenetic}
-Still, the proposed mechanisms so far include only ecological or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
+Still, the most mechanisms integrated in island models [?] so far include only ecological or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
 while genetic properties are hardly recognized in the island context.
-At the same time, an increasing number of studies investigates genetic processes and
-their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein).
+This is surprising given the essential role islands have played in the development of evolutionary theories.
+Nevertheless, an increasing number of studies investigates genetic processes and
+their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein) as well as evolutionary rescue provided by the feedback with ecological processes (Schiffers et al.
+2013).
 %%TODO: here come a list of processes with examples
@@
 Besides arthropods (Howthorne \& Via 2001), these processes can be observed in plants as well.
+%Are you talking about islands here?
+Be clearer, because you said about the lack of geentic studies in island contexts.
 For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
@@
 levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption.
+%what Schiffers et al say about linkage and evolutionary rescue (basically the opposite of diversification?)
 
@@
 
-To address this hypothesis, we present a mechanistic model that tries to remedy above mentioned shortcomings
+To address this hypothesis, 
+we present a mechanistic model that overcomes above-mentioned shortcomings (LIST THEM HERE IN SHORT)
 by implementing genetic processes on a niche-explicit background
-using an island biogeography framework.%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
+using an island biogeography simulation arena.%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
 %% and what trait combinations constitute successful colonizer and radiating species, respectively.
-We will adress the questions of what relative impact does (a) genetic linkage and
-(b) reproductive isolation have on emerging species numbers and diversity,
+We will address the questions of what relative impact does (a) genetic linkage and
+(b) reproductive isolation have on emerging species richenss and composition,
 and (c) which genetic and functional trait characteristics distinguish successful colonizer,
@@
 \section{Material and Methods}
-We use an individual based metacommunity model written in julia %\cite{JULIALANG}.
-Each individual is characterized by a set of genetically controlled traits and is submitted to processes of
+We use an individual-based metacommunity model written in julia %\cite{JULIALANG}.
+Each individual is characterized by a set of genetically controlled (i.e. genomically-coded) traits, which influence ecological processes of
 establishment, competition, growth, (density independent) mortality, reproduction and dispersal.
-Each of which process is affected by the individuals' trait characteristics.
-The simulation arena consists of a larger landmass (the mainland) holding an initial random species pool and a smaller, initially unpopulated island some distance apart.
+The simulation arena consists of a larger landmass (the mainland) holding an initial random species pool and a smaller, initially unpopulated island some distance apart (user-defined in the experimental design).
+% it would be nice to hav a table with all model configurations that can be defined by the user upon designing the experiment.
+This might help readers see that the model is general enough to be applied to any ecological arena (mainland-island being one).
 During the course of the simulation, individuals may disperse from the mainland to the island, thus colonizing it. 
@@
 
-Building on the model of Cabral et al. (2017a, b), most of the biological rates are controlled following the metabolic theory
-of ecology (MTE, Brown et al.\ 2004).
+Building on the model of Cabral et al. (2017a, b), most of the biological rates follow the metabolic theory
+of ecology (MTE, Brown et al.\ 2004). % instead of saying 'most', say which follow and which don't
 In MTE, a biological rate $b$ depends on the temperature $T$ and individual mass $M$, scaling a base rate $b_0$ as:
@@
 
-Using the thus simulated data, we calculate a phylogenies of the colonizing lineages, species numbers, trait spaces etc. [WHAT ELSE?]
+Using the thus simulated data, we calculate phylogenies of the colonizing lineages, species numbers, trait spaces etc. [WHAT ELSE?]
 
@@
 (Dynamic) resource availability: assumed constant in model, but petentially more variation, complexity in community dynamics.
-Interaction: in present model only indirect interaction via competition, but predation, parasitation, herbivory, mutualism will increase complexity.
-Especially pollination is assumed to be an important interaction, because having a high impact on evolution and establishment of species due to specialization.
+Interaction: in present model only resource competition, % I would call just say resource competition
+but predation, parasitatism, herbivory, mutualism will increase complexity, and should be rather aimed at future modle developments once mechanistic consequences of competition have been better investigated.
+In particular, future model developments may concetrate on pollination, which may have a high impact on evolution and establishment of both plant and animal species due to specialization.
+%examples for islands?
 %[maybe some words on trophic interactions.]
@@
 recombination between populations to speciation (Via 2011, but see Feder \& Nosil 2010).
-During incipient speciation, however, th
+%During incipient speciation, however, th
 
 In this work we concentrated on genetic influences on speciation.
-However, there is ongoing debate on how strongly genetic factors also impact extinction processes (Jamieson 2007).
+However, there is ongoing debate on how strongly genetic factors also impact extinction processes (Jamieson 2007). For example, [discuss the evolutionary rescue studied by Schiffers et al. 2013]
 Seeing that observed speciation rates are in fact net speciation rates highlights the need to mechanistically explore

### 2018-05-14 src/constants.jl
decrease mutation rate

@@ -7,7 +7,7 @@ const mortality = exp(22) # global base mortality from Brown et al. 2004 is 26.3
 const fertility = exp(30.0) # global base reproduction rate 23.8 from Brown et al. 2004, alternatively 25.0
 const phylconstr = 10 #parse(ARGS[2])
 # const meangenes = 20 # mean number of genes per individual
-const mutationrate = 1e-3 * 0.3e11 # 1 base in 1000, correction factor for metabolic function
+const mutationrate = 1e-3 * 1.0e10 # 1 base in 1000, correction factor for metabolic function
 const isolationweight = 3 # additional distance to be crossed when dispersing from or to isolated patches
 const cellsize = 1e8 # ca. 100 tons carrying cap.
 const maxdispmean = 10 # maximum mean dispersal distance

### 2018-05-14 paper/draft.tex
remove some space and comments

@@
   and which of them resulted in radiations.
-
 \end{abstract}
@@
 \section{Introduction}
-%% island biota have always fascinated and inspired researchers (Darwin, Wallace, ...)
-
-%% This interest in islands and their recognition as highly suitable model systems resulted in the formulation of
-%% important theoretical frameworks, foremost MacArthur and Wilson's seminal equilibrium theory of island biogeography (ETIB).
-%% Later ... GDM.
-
-%% These theories have in common a focus on species numbers and respective rates of migration, extinction and speciation.
-%% While they are successful in describing species numbers, they are neutral in design (in the UNTB sense) and thus were not meant to consider species traits and compositions.
-
-%% These patterns have been explained theoretically [here, here and here], but usually one at a time.
-%% In the following, we aim to produce and thus explain these forementioned patterns using a single framework, and thus following the same set of (rather simple) rules.
-
-%% [NEW ANGLE:]
 Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation (e.g.

### 2018-05-14 paper/draft.tex
add some comments for intro and MM from PubClub

@@
 \section{Introduction}
+%% Intro subsections (==paragraphs):
+%% 1. Major topic (include only background relevant for exp. design)
+%% 2. Islands
+%% 3. Forces
+%% 4. Models
+%% 5. Aim, Objectives
+%% 2+3 may be combined, as well as 1+2(+3)
+
 Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation (e.g.
@@
 \section{Material and Methods}
+%% Include simplified model description ``based on ODD''
+%% scheduling (order of processes)
+%% justify each decision with reference
+%% Experimental design
 We use an individual-based metacommunity model written in julia %\cite{JULIALANG}.

### 2018-05-14 paper/draft.tex
add some more comments for intro from PubClub

@@
 %% 2+3 may be combined, as well as 1+2(+3)
+
+%% Cite & look at Katja Schiffer paper
+%% "species rich vs. species poor lineage''
+%% ``emerging systematic, functional, phylogenetic patterns''
+%% Hurlbert
 
@@
 
+%% In this study, we aim to address...
 To address this hypothesis, 
@@
 (b) reproductive isolation have on emerging species richenss and composition,
-and (c) which genetic and functional trait characteristics distinguish successful colonizer,
+and (c) which genetic and functional trait syndromes %% Which?
+distinguish successful colonizer,
 strong competitor and diversifying lineages.

### 2018-05-14 paper/draft.tex
comment intro subsections, add line breaks for better readability

@@
 
-Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation (e.g.
-Losos \& Ricklefs 2009; Warren et al.
-2015  Islands as model systems in ecology and evolution: prospects fifty years after MacArthur‐Wilson.
-EcoLett).
-%TODO: ref
+%% major topic: besides classical mechanisms (ecological, environmental), genetic mechanisms affect speciation/radiations
+
+%% Islands
+Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics
+like isolation (e.g. Losos \& Ricklefs 2009; Warren et al. 2015  %%Islands as model systems in ecology and evolution: prospects fifty years after MacArthur‐Wilson. EcoLett). %TODO: ref
 Part of this value as study system originates by typically remarkable species.
-In fact, islands often possess unique species communities and can exceede mainland areas by a factor of 9.5 and 8.1 in plant and animal  endemism, respectively (Kier et al.
-2005.
-A global assessment of endemism and species richness across island and mainland regions.
-PNAS).
-%TODO: refs
+In fact, islands often possess unique species communities and can exceede mainland areas
+by a factor of 9.5 and 8.1 in plant and animal  endemism, respectively
+(Kier et al. 2005. %A global assessment of endemism and species richness across island and mainland regions. PNAS). %TODO: refs
 Such high endemism makes islands  especially threatened by human impacts, in particular climate change and invasive species (Fordham \& Brook 2010). %TODO: refs
@@
 
-For the prediction of island community size, several theoretical frameworks have been proposed, including one of the first and to this day still highly influential classical ecological theories, namely MacArthur and Wilson's equilibrium theory of island biogeography (ETIB).
-%TODO: ref
-In it, species richness on islands is the result of opposing rates of immigration and extinction, both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
+For the prediction of island community size, several theoretical frameworks have been proposed,
+including one of the first and to this day still highly influential classical ecological theories,
+namely MacArthur and Wilson's equilibrium theory of island biogeography (ETIB). %TODO: ref
+In it, species richness on islands is the result of opposing rates of immigration and extinction,
+both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
 This relationship was since successfully applied to a multitude of islands (Lomolina \& Weiser 2001, Triantis et al.\ 2011)%TODO: refs
 , but also island like systems, such as coral reefs, lakes/ponds or forest fragments (Kelly et al.\ 1989, Peay et al.\ 2007, ). %TODO: refs
-This general applicability (Lomolino 2000), however, cannot hide the fact that the ETIB fails to formalize a mechanistic explanation of how extinction and immigration are modified by area and isolation, respectively.
-Additionally, it has been demonstrated more recently that area also affects immigration via target effects (REFs) and isolation affects extinction via rescue-effects (REFs).
-Finally, ETIB has particularly been criticized for its lack of considering speciation (see e.g.\ Losos \& Schluter 2000).
-%TODO: refs
+This general applicability (Lomolino 2000), however, cannot hide the fact
+that the ETIB fails to formalize a mechanistic explanation of how extinction and immigration are modified by area and isolation, respectively.
+Additionally, it has been demonstrated more recently that area also affects immigration via target effects (REFs) a
+nd isolation affects extinction via rescue-effects (REFs).
+Finally, ETIB has particularly been criticized for its lack of considering speciation (see e.g.\ Losos \& Schluter 2000). %TODO: refs
 
@@
 The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
-whith islands emerging from sea level via volcanic hotspot activity, increasing over time in size and topographic complexity, then eroding and subsiding in the ocean again when the island is not over the hotspot anymore.
+whith islands emerging from sea level via volcanic hotspot activity,
+increasing over time in size and topographic complexity,
+then eroding and subsiding in the ocean again when the island is not over the hotspot anymore.
 In doing so, the model incorporates concepts of niche theory, speciation and species succession.
-Yet, there are still many processes and drivers deemed important, which are not considered by any formal theory and often little investigated [if you say this, you have to mention at leastr a couple os examples, otherwise it is an empty claim].
+Yet, there are still many processes and drivers deemed important,
+which are not considered by any formal theory and often little investigated
+%[if you say this, you have to mention at leastr a couple os examples, otherwise it is an empty claim].
 
+%% Models
 The problem of investigating these processes is the time frame and study extent needed for a proper evaluation.
@@
 (Leidinger \& Sarmento Cabral 2017).
-Although many of the reviewed models integrate multiple processes at a time, unfortunately, most models failed to include microevolutionary or more explicit genetic processes %\cite{leidinger2017biodiversity}
+Although many of the reviewed models integrate multiple processes at a time, unfortunately,
+most models failed to include microevolutionary or more explicit genetic processes %\cite{leidinger2017biodiversity}
 (Leidinger \& Sarmento Cabral 2017).
@@
 divergence event if the population remains isolated during that time.
-But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005) and genetic mechanisms should also play a central role for speciation/radiation and species richness.
-Questions, such as "[cite two questions from Patino et al.
-2017]", thus remain largely open (Patiño et al.\ 2017) and may require explicit consideration of evolutionary processes.
+But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005)
+and genetic mechanisms should also play a central role for speciation/radiation and species richness.
+Questions, such as "[cite two questions from Patino et al. 2017]", thus remain largely open (Patiño et al.\ 2017)
+and may require explicit consideration of evolutionary processes.
 %% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact of explosively radiating lineages (Hughes \& Eastwood 2006, Kocher 2004)
 %%, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
-
 %% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
@@
 phylogenetic similarity \cite{losos2008phylogenetic}
-Still, the most mechanisms integrated in island models [?] so far include only ecological or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
+Still, the most mechanisms integrated in island models [?] so far include only ecological
+or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
 while genetic properties are hardly recognized in the island context.
@@
 Nevertheless, an increasing number of studies investigates genetic processes and
-their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein) as well as evolutionary rescue provided by the feedback with ecological processes (Schiffers et al.
-2013).
+their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein)
+as well as evolutionary rescue provided by the feedback with ecological processes (Schiffers et al. 2013).
 %%TODO: here come a list of processes with examples
@@
 
+%% Aim, objectives:
 %% In this study, we aim to address...
@@
 We use an individual-based metacommunity model written in julia %\cite{JULIALANG}.
-Each individual is characterized by a set of genetically controlled (i.e. genomically-coded) traits, which influence ecological processes of
+Each individual is characterized by a set of genetically controlled (i.e. genomically-coded) traits,
+which influence ecological processes of
 establishment, competition, growth, (density independent) mortality, reproduction and dispersal.
-The simulation arena consists of a larger landmass (the mainland) holding an initial random species pool and a smaller, initially unpopulated island some distance apart (user-defined in the experimental design).
+The simulation arena consists of a larger landmass (the mainland) holding an initial random species pool and a smaller,
+initially unpopulated island some distance apart (user-defined in the experimental design).
 % it would be nice to hav a table with all model configurations that can be defined by the user upon designing the experiment.
-This might help readers see that the model is general enough to be applied to any ecological arena (mainland-island being one).
+% This might help readers see that the model is general enough to be applied to any ecological arena (mainland-island being one).
 During the course of the simulation, individuals may disperse from the mainland to the island, thus colonizing it. 
@@
 Interaction: in present model only resource competition, % I would call just say resource competition
-but predation, parasitatism, herbivory, mutualism will increase complexity, and should be rather aimed at future modle developments once mechanistic consequences of competition have been better investigated.
-In particular, future model developments may concetrate on pollination, which may have a high impact on evolution and establishment of both plant and animal species due to specialization.
-%examples for islands?
+but predation, parasitatism, herbivory, mutualism will increase complexity,
+and should be rather aimed at future modle developments once mechanistic consequences of competition have been better investigated.
+In particular, future model developments may concetrate on pollination,
+which may have a high impact on evolution and establishment of both plant and animal species due to specialization. %examples for islands?
 %[maybe some words on trophic interactions.]
@@
 In this work we concentrated on genetic influences on speciation.
-However, there is ongoing debate on how strongly genetic factors also impact extinction processes (Jamieson 2007). For example, [discuss the evolutionary rescue studied by Schiffers et al. 2013]
+However, there is ongoing debate on how strongly genetic factors also impact extinction processes (Jamieson 2007).
+For example, [discuss the evolutionary rescue studied by Schiffers et al. 2013]
 Seeing that observed speciation rates are in fact net speciation rates highlights the need to mechanistically explore

### 2018-05-14 paper/draft.tex
begin restructuring intro

@@
 %% major topic: besides classical mechanisms (ecological, environmental), genetic mechanisms affect speciation/radiations
+Island radiations like the Darwin finches on Galápagos or \textit{Aeonium} on the Canaries have fascinated researchers throughout history %% cite? Darwin? ...?
+To this date, however, it is unclear why these genera produced this multitude of species as opposed to most other species groups.
+%%
+Nevertheless, an increasing number of studies investigates genetic processes and
+their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein)
+as well as evolutionary rescue provided by the feedback with ecological processes (Schiffers et al. 2013).
+%%TODO: here come a list of processes with examples
+%% - genetic processes in general
+%% - isolation processes
+Besides arthropods (Howthorne \& Via 2001), these processes can be observed in plants as well.
+%Are you talking about islands here?
+Be clearer, because you said about the lack of geentic studies in island contexts.
+For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
+seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al.\ 2013). % cite properly!
+Considering these phenomena, we expect speciation rates to be highest at intermediate levels of gene linkage --- low
+levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption.
+%what Schiffers et al say about linkage and evolutionary rescue (basically the opposite of diversification?)
 
@@
 Hence, understanding the origin, maintenance, functioning and evolution of island communities is of a high priority objective.
-
 For the prediction of island community size, several theoretical frameworks have been proposed,
@@
 Finally, ETIB has particularly been criticized for its lack of considering speciation (see e.g.\ Losos \& Schluter 2000). %TODO: refs
-
 About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM, Whittaker et al.\ 2007, 2008). %TODO: ref
@@
 This is surprising given the essential role islands have played in the development of evolutionary theories.
-Nevertheless, an increasing number of studies investigates genetic processes and
-their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein)
-as well as evolutionary rescue provided by the feedback with ecological processes (Schiffers et al. 2013).
-%%TODO: here come a list of processes with examples
-%% - genetic processes in general
-%% - isolation processes
-Besides arthropods (Howthorne \& Via 2001), these processes can be observed in plants as well.
-%Are you talking about islands here?
-Be clearer, because you said about the lack of geentic studies in island contexts.
-For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
-seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al.\ 2013). % cite properly!
-Considering these phenomena, we expect speciation rates to be highest at intermediate levels of gene linkage --- low
-levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption.
-%what Schiffers et al say about linkage and evolutionary rescue (basically the opposite of diversification?)
 

### 2018-05-15 islandsim.jl
Added config file parsing

@@ -5,6 +5,9 @@
 # Ludwig Leidinger 2018
 # <l.leidinger@gmx.net>
 #
+# Daniel Vedder 2018
+# <daniel.vedder@stud-mail.uni-wuerzburg.de>
+#
 # For a list of options, run `julia islandsim.jl --help`
 #
 # <MAPFILE> is a textfile containing information about the simulation arena
@@ -15,6 +18,9 @@ thisDir = pwd() * "/src"
 any(path -> path == thisDir, LOAD_PATH) || push!(LOAD_PATH, thisDir)
 
 using GeMM, ArgParse
+using ProfileView #DEBUG
+
+Profile.init(n=10^8, delay=0.005) #DEBUG
 
 function parsecommandline()
     s = ArgParseSettings()
@@ -26,7 +32,11 @@ function parsecommandline()
         "--maps", "-m"
             help = "list of map files, comma separated"
             arg_type = String
-            required = true
+            required = false
+        "--config", "-c"
+            help = "name of the config file"
+            arg_type = String
+            required = false
         "--linkage", "-l"
             help = "gene linkage (\"none\", \"random\" or \"full\")"
             arg_type = String
@@ -53,55 +63,80 @@ function parsecommandline()
         "--static"
             help = "static mainland. Turns off any dynamics on the continent"
             action = :store_true
-#        "arg1"
-#            help = "a positional argument"
-#            required = false
         end
     return parse_args(s)
 end
 
+function parseconfig(configfilename::String, settings::Dict{String,Any})
+    params = String["seed", "maps", "linkage", "nniches", "tolerance", "dest",
+                    "static", "propagule-pressure", "carrying-capacity", "disturbance"]
+    configfile = open(configfilename)
+    config = readlines(configfile)
+    close(configfile)
+    filter!(x -> isempty(strip(x)) || (x[1] != '#'), config)
+    config = map(split, config)
+    for c in config
+        if length(c) != 2
+            warn("Bad config file syntax: $c")
+        elseif c[1] in params
+            settings[c[1]] = parse(c[2])
+        else
+            warn(c[1]*" is not a recognized parameter!")
+        end
+    end
+    for p in params
+        !(p in keys(settings)) && warn("Parameter $p is not set!")
+    end
+    cp(configfilename, settings["dest"]*"/"*configfilename, remove_destination=true)
+    return settings
+end
 
[…]

### 2018-05-15 run.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+echo "Starting simulation run."
+
+date
+#TODO Test for destination commandline parameter
+#rm results/tests/* #TODO Remove this again
+time julia islandsim.jl --config test.config
+#TODO archive results
+date
+
+echo "Done."

### 2018-05-18 paper/draft.tex
add comment with ref
new additions, ideas for intro, comment disc

@@
 Island radiations like the Darwin finches on Galápagos or \textit{Aeonium} on the Canaries have fascinated researchers throughout history %% cite? Darwin? ...?
-To this date, however, it is unclear why these genera produced this multitude of species as opposed to most other species groups.
-%%
-Nevertheless, an increasing number of studies investigates genetic processes and
-their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein)
+To this date, however, it is unclear why these genera produced this multitude of species as opposed to most other genera that remained species poor.
+Until recently, researchers focused on environmental and ecological factors to explain different speeds and ocurrences of speciation.
+%% HERE COME SOME EXAMPLES
+Yet, these mechanisms often fail to explain how the rapid establishment of reproductive barriers or isolation arises,
+which ultimately results in sympatric speciation. %CLAIM! PROVIDE REFERENCES!
+%% somewhere here Patino et al 2018 anagenesis, cladogenesis, etc islands
+With the growing importance of genetic data and analyses, an increasing number of studies also investigates genetic processes and
+their effect on population variality. %EXAMPLES
+Some of these studies highlight the importance of genetic linkage or ``hitchhiking'' on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein)
 as well as evolutionary rescue provided by the feedback with ecological processes (Schiffers et al. 2013).
@@
 seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al.\ 2013). % cite properly!
-Considering these phenomena, we expect speciation rates to be highest at intermediate levels of gene linkage --- low
-levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption.
-%what Schiffers et al say about linkage and evolutionary rescue (basically the opposite of diversification?)
+In an island context, genetic studies are even rarer and almost exclusively of a descriptive nature.
 
@@
 strong competitor and diversifying lineages.
+Considering these phenomena, we expect speciation rates to be highest at intermediate levels of gene linkage --- low
+levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption.
+%what Schiffers et al say about linkage and evolutionary rescue (basically the opposite of diversification?)
 
@@
   \caption{Number of linkage units and number of species per lineage XXX timesteps after colonization.
-    [Histogram (with threshold category: strongly diversified lineages) or scatter plot. Two panels/color codes: high and low degree of environmental heterogeneity]}
+    [Histogram (with threshold category: strongly diversified lineages) or scatter plot. ]}%Two panels/color codes: high and low degree of environmental heterogeneity]}
   \label{specieslinkage}
@@
 \section{Discussion}
+%Assortative mating
+
 Shortcomings. (/Processes/Factors not considered.)

### 2018-05-24 islandsim.jl
Revert "Worked on `genesis()`"
This reverts commit 50e69a42f1ddbe830a7fd5b066bfbac91fe13c27.

@@ -35,9 +35,9 @@ function defaultSettings()
           ("mutate", true), # mutations occur
           ("cellsize", 100), # maximum biomass per patch in tonnes
     	  ("initadults", false), # individuals are initiated with adults size
-    	  # invasion specific settings
+          # invasion specific settings
           ("propagule-pressure", 0), # TODO
-          ("disturbance", 0)]) # percentage of individuals killed per update per patch
+          ("disturbance", 0)]) # percentage of individuals killed per update per cell
 end
 
 function parsecommandline()

### 2018-05-24 src/init.jl

@@ -15,7 +15,7 @@ function genesis(settings::Dict{String,Any},
                                                 "temptol"]) # minimal required traitnames
     community = Individual[]
     totalmass = 0.0
-    while totalmass < settings["cellsize"] 
+    while true 
         lineage = randstring(4)
         meangenes = length(traitnames)
         ngenes = rand(Poisson(meangenes))
@@ -34,13 +34,14 @@ function genesis(settings::Dict{String,Any},
         popsize = round(fertility * traitdict["repsize"]^(-1/4) * exp(-act/(boltz*traitdict["tempopt"]))) # population size determined by adult size and temperature niche optimum
         settings["initadults"]? indsize = traitdict["repsize"] : indsize = traitdict["seedsize"]
         popmass = popsize * indsize
-        totalmass += popmass
-        if totalmass + popmass > settings["cellsize"]
-            popmass = settings["cellsize"] - totalmass
+        if totalmass + popmass > settings["cellsize"] # stop loop if cell is full
+            break
         end
+        totalmass += popmass
         newind = Individual(lineage, chromosomes, traitdict, 0, false, 1.0, indsize)
         for i in 1:popsize
-            push!(community, deepcopy(newind))
+            !settings["static"] && (newind = deepcopy(newind))
+            push!(community, newind)
         end
     end
     community

### 2018-05-24 src/mainprocs.jl

@@ -159,7 +159,7 @@ end
 species-independent mortality due to disturbance on patch `p`
 """
 function disturb!(patch::Patch, intensity::Int64)
-    (intensity == 0 || length(patch.community) == 0) && return
+    intensity == 0 && return
     intensity > 100 && error("intensity must be less than 100%")
     deaths = Integer(round(length(patch.community) * (intensity/100)))
     dead = unique(rand(1:length(patch.community), deaths))
@@ -278,7 +278,7 @@ end
     reproduce!(w, p)
 Reproduction of individuals in a patch `p` whithin a world (array of patches) `w`
 """
-function reproduce!(world::Array{Patch,1}, patch::Patch) #TODO: refactor!
+function reproduce!(world::Array{Patch,1}, patch::Patch) #TODO: refactorize!
     idx = 1
     temp = patch.altitude
     seedbank = Individual[]
@@ -301,7 +301,7 @@ function reproduce!(world::Array{Patch,1}, patch::Patch) #TODO: refactor!
                 posspartners = findposspartners(world, patch.community[idx], patch.location) # this effectively controls frequency of reproduction
                 # length(posspartners) == 0 && push!(posspartners, patch.community[idx]) # selfing if no partners # CAVE!
                 if length(posspartners) > 0
-                    partner = rand(posspartners) #XXX There is at most one posspartner!
+                    partner = rand(posspartners)
                     parentmass = currentmass - noffs * seedsize # subtract offspring mass from parent
                     if parentmass <= 0
                         idx += 1 #splice!(patch.community, idx)
@@ -327,7 +327,6 @@ function reproduce!(world::Array{Patch,1}, patch::Patch) #TODO: refactor!
         end
         idx += 1
     end
-    info("Patch $(patch.id): $(length(patch.community)) individuals, $(length(seedbank)) offspring.")
     append!(patch.community, seedbank)
 end
 

### 2018-05-24 src/subprocs.jl

@@ -187,9 +187,7 @@ function findposspartners(world::Array{Patch,1}, ind::Individual, location::Tupl
     posspartners
 end
 
-function createtraits(traitnames::Array{String,1}, settings::Dict{String,Any})
-    #XXX: this is all very ugly. (case/switch w/ v. 2.0+?)
-    #TODO Default trait values need to appear elsewhere!
+function createtraits(traitnames::Array{String,1}, settings::Dict{String,Any}) #TODO: this is all very ugly. (case/switch w/ v. 2.0+?)
     traits = Trait[]
     seedsize = exp(-7 + 20 * rand()) # corresponds to 1mg to 22kg
     repsize = exp(0 + 17 * rand()) # 1g to 24t

### 2018-05-24 src/subprocs.jl
include warning in traitsexist() and add method for Dict instead of Ind

@@ -31,9 +31,22 @@ function chrms2traits(chrms::Array{Chromosome,1})
     traitdict
 end
 
+function traitsexist(traits::Dict{String, Float64}, traitnames::Array{String, 1})
+    for trait in traitnames
+        if !haskey(traits, trait)
+            warn("Missing trait \"", trait, "\". Individual might be killed.")
+            return false
+        end
+    end
+    true
+end
+
 function traitsexist(ind::Individual, traitnames::Array{String, 1})
     for trait in traitnames
-        !haskey(ind.traits, trait) && return false
+        if !haskey(ind.traits, trait)
+            warn("Individual is missing trait \"", trait, "\". Might be killed.")
+            return false
+        end
     end
     true
 end
@@ -173,7 +186,7 @@ function findposspartners(world::Array{Patch,1}, ind::Individual, location::Tupl
             mate = targetpatch[1].community[mateidx]
             mate.age == 0 && continue
             mate.lineage != ind.lineage && continue
-            !traitsexist(mate, ["repsize"]) && continue
+            !traitsexist(mate.traits, ["repsize"]) && continue
             mate.size < mate.traits["repsize"] && continue
             mate.isnew && continue
             !iscompatible(mate, ind) && continue

### 2018-05-24 src/mainprocs.jl
remove info messages to adapt to new traitsexist()

@@ -82,8 +82,7 @@ function establish!(patch::Patch, nniches::Int64=1)
     while idx <= size(patch.community,1)
         if patch.community[idx].isnew || patch.community[idx].age == 0
             fitness = 1
-            if !traitsexist(patch.community[idx], ["temptol", "tempopt"]) # CAVE: should check for more traits!
-                info(STDERR, "Individual killed due to missing trait(s).")
+            if !traitsexist(patch.community[idx].traits, ["temptol", "tempopt"]) # CAVE: should check for more traits!
                 splice!(patch.community, idx) # kill it!
                 continue
             end
@@ -93,8 +92,7 @@ function establish!(patch::Patch, nniches::Int64=1)
             fitness > 1 && (fitness = 1) # should be obsolete
             fitness < 0 && (fitness = 0) # should be obsolete
             if nniches >= 2
-                if !traitsexist(patch.community[idx], ["prectol", "precopt"]) # CAVE: should check for more traits!
-                    info(STDERR, "Individual killed due to missing trait(s).")
+                if !traitsexist(patch.community[idx].traits, ["prectol", "precopt"]) # CAVE: should check for more traits!
                     splice!(patch.community, idx) # kill it!
                     continue
                 end
@@ -186,8 +184,7 @@ function grow!(patch::Patch)
     temp = patch.altitude
     idx = 1
     while idx <= size(patch.community,1)
-        if !traitsexist(patch.community[idx], ["repsize"])
-            info(STDERR, "Individual killed due to missing trait(s).")
+        if !traitsexist(patch.community[idx].traits, ["repsize"])
             splice!(patch.community, idx)
             continue
         elseif !patch.community[idx].isnew
@@ -223,8 +220,7 @@ function disperse!(world::Array{Patch,1}, static::Bool = true) # TODO: additiona
     for patch in world
         idx = 1
         while idx <= size(patch.community,1)
-            if !traitsexist(patch.community[idx], ["dispmean", "dispshape"])
-                info(STDERR, "Individual killed due to missing trait(s).")
+            if !traitsexist(patch.community[idx].traits, ["dispmean", "dispshape"])
                 splice!(patch.community,idx)
                 continue
             elseif !patch.community[idx].isnew && patch.community[idx].age == 0
@@ -284,8 +280,7 @@ function reproduce!(world::Array{Patch,1}, patch::Patch) #TODO: refactorize!
     temp = patch.altitude
     seedbank = Individual[]
[…]

### 2018-05-24 islandsim.jl
Added `outfreq` parameter and moved settings to constants.jl

@@ -23,7 +23,7 @@ function simulation!(world::Array{Patch,1}, settings::Dict{String,Any}, mapfile:
     info("Starting simulation")
     for t in 1:timesteps
         info("UPDATE $t, population size $(sum(x -> length(x.community), world))")
-        (t == 1 || mod(t, 1000) == 0) && writedata(world, settings, t)
+        (t == 1 || mod(t, settings["outfreq"]) == 0) && writedata(world, settings, t)
         establish!(world, settings["nniches"], settings["static"])
         checkviability!(world, settings["static"])
         compete!(world, settings["static"])

### 2018-05-24 src/constants.jl

@@ -1,4 +1,4 @@
-# Constants for GeMM
+# Constants and settings for GeMM
 
 const boltz = 1.38064852e-23 # J/K = m2⋅kg/(s2⋅K)
 const act = 1e-19 # activation energy /J, ca. 0.63eV - Brown et al. 2004
@@ -11,4 +11,26 @@ const mutationrate = 1e-3 * 0.3e11 # 1 base in 1000, correction factor for metab
 const isolationweight = 3 # additional distance to be crossed when dispersing from or to isolated patches
 const maxdispmean = 10 # maximum mean dispersal distance
 const genelength = 20 # sequence length of genes
+
+# Return the default settings. All parameters must be registered here.
+function defaultSettings()
+    Dict(# general software settings
+          "seed" => 1, # for the RNG, seed = 0 -> random seed
+          "maps" => nothing, # comma-separated list of map files
+          "config" => nothing, # configuration file name
+          "fasta" => true, # record fasta data?
+          "dest" => string(Dates.today()), # output folder name
+          "outfreq" => 1000, # output frequency
+          # main model settings
+          "linkage" => "random", # gene linkage type
+          "nniches" => 2, # number of environmental niches (max. 3)
+          "tolerance" => "low", # sequence similarity threshold for reproduction
+          "static" => true, # mainland sites don't undergo eco-evolutionary processes
+          "mutate" => true, # mutations occur
+          "initadults" => false, # initialize organisms as adults
+          "cellsize" => 100, # maximum biomass per cell in tonnes
+          # invasion specific settings
+          "propagule-pressure" => 0, # TODO
+          "disturbance" => 0) # percentage of individuals killed per update per cell
+end
 

### 2018-05-24 src/io.jl

@@ -1,26 +1,5 @@
 # IO functions for GeMM
 
-# Return the default settings. All parameters must be registered here.
-function defaultSettings()
-    Dict(# general software settings
-          "seed" => 1, # for the RNG, seed = 0 -> random seed
-          "maps" => nothing, # comma-separated list of map files
-          "config" => nothing, # configuration file name
-          "fasta" => true, # record fasta data?
-          "dest" => string(Dates.today()), # output folder name
-          # main model settings
-          "linkage" => "random", # gene linkage type
-          "nniches" => 2, # number of environmental niches (max. 3)
-          "tolerance" => "low", # sequence similarity threshold for reproduction
-          "static" => true, # mainland sites don't undergo eco-evolutionary processes
-          "mutate" => true, # mutations occur
-          "initadults" => false, # initialize organisms as adults
-          "cellsize" => 100, # maximum biomass per cell in tonnes
-          # invasion specific settings
-          "propagule-pressure" => 0, # TODO
-          "disturbance" => 0) # percentage of individuals killed per update per cell
-end
-
 function parsecommandline()
     defaults = defaultSettings()
     s = ArgParseSettings()

### 2018-05-24 src/subprocs.jl
add more methods to traitsexist()

@@ -51,6 +51,22 @@ function traitsexist(ind::Individual, traitnames::Array{String, 1})
     true
 end
 
+function traitsexist(traits::Dict{String, Float64}, traitname::String)
+    if !haskey(traits, traitname)
+        warn("Missing trait \"", trait, "\". Individual might be killed.")
+        return false
+    end
+    true
+end
+
+function traitsexist(ind::Individual, traitname::String)
+    if !haskey(ind.traits, traitname)
+        warn("Individual is missing trait \"", trait, "\". Might be killed.")
+        return false
+    end
+    true
+end
+
 function gausscurve(b::Float64, c::Float64, x::Float64, a::Float64=1.0)
     if c != 0 && a != 1.0
         a = 1 / (c * sqrt(2 * pi))

### 2018-05-24 src/subprocs.jl
fix

@@ -53,7 +53,7 @@ end
 
 function traitsexist(traits::Dict{String, Float64}, traitname::String)
     if !haskey(traits, traitname)
-        warn("Missing trait \"", trait, "\". Individual might be killed.")
+        warn("Missing trait \"", traitname, "\". Individual might be killed.")
         return false
     end
     true
@@ -61,7 +61,7 @@ end
 
 function traitsexist(ind::Individual, traitname::String)
     if !haskey(ind.traits, traitname)
-        warn("Individual is missing trait \"", trait, "\". Might be killed.")
+        warn("Individual is missing trait \"", traitname, "\". Might be killed.")
         return false
     end
     true

### 2018-05-24 islandsim.jl
Settings are now explicitly saved to file.
Also put `mapfile` back into the argument list of write*()

@@ -23,7 +23,7 @@ function simulation!(world::Array{Patch,1}, settings::Dict{String,Any}, mapfile:
     info("Starting simulation")
     for t in 1:timesteps
         info("UPDATE $t, population size $(sum(x -> length(x.community), world))")
-        (t == 1 || mod(t, settings["outfreq"]) == 0) && writedata(world, settings, t)
+        (t == 1 || mod(t, settings["outfreq"]) == 0) && writedata(world, settings, mapfile, t)
         establish!(world, settings["nniches"], settings["static"])
         checkviability!(world, settings["static"])
         compete!(world, settings["static"])
@@ -49,7 +49,7 @@ function runit(settings::Dict{String,Any})
         i == 1 && (world = createworld(maptable, settings))
         i > 1 && updateworld!(world,maptable,settings["cellsize"])
         simulation!(world, settings, settings["maps"][i], timesteps)
-        writedata(world, settings, -1)
+        writedata(world, settings, settings["maps"][i], -1)
         println("WORLD POPULATION: $(sum(x -> length(x.community), world))") #DEBUG
         println("WORLD MEMORY: $(round(Base.summarysize(world)/1024^2, 2)) MB") #DEBUG
     end
@@ -57,7 +57,7 @@ end
 
 ## Settings
 const allargs = parsecommandline()
-if haskey(allargs, "config") && allargs["config"] != nothing
+if isfile(allargs["config"])
     allargs = parseconfig(allargs["config"], allargs)
 end
 

### 2018-05-24 src/constants.jl

@@ -17,7 +17,7 @@ function defaultSettings()
     Dict(# general software settings
           "seed" => 1, # for the RNG, seed = 0 -> random seed
           "maps" => nothing, # comma-separated list of map files
-          "config" => nothing, # configuration file name
+          "config" => "simulation.config", # configuration file name
           "fasta" => true, # record fasta data?
           "dest" => string(Dates.today()), # output folder name
           "outfreq" => 1000, # output frequency

### 2018-05-24 src/io.jl

@@ -181,9 +181,6 @@ function makefasta(world::Array{Patch, 1}, io::IO = STDOUT, sep::String = "", on
                             traits *= trait.name * string(trait.value) * ","
                         end
                     end
-                    #XXX Don't use string interpolation:
-                    # https://docs.julialang.org/en/stable/manual/performance-tips/#Avoid-string-interpolation-for-I/O-1
-                    #header = ">$counter x$(patch.location[1]) y$(patch.location[2]) $(ind.lineage) c$chrmno g$geneno $traits"
                     header = ">"*counter*" x"*string(patch.location[1])*" y"*string(patch.location[2])
                     header *= " "*ind.lineage*" c"*str(chrmno)*" g"*string(geneno)*" "*string(traits)
                     println(io, header)
@@ -210,9 +207,19 @@ function setupdatadir(settings::Dict{String, Any})
     else
         info("Setting up output directory $(settings["dest"])")
         mkpath(settings["dest"])
-        cp(settings["config"], joinpath(settings["dest"], settings["config"]))
+        writesettings(settings)
         for m in settings["maps"]
             cp(m, joinpath(settings["dest"], m))
+        end
+    end
+end
+
+function writesettings(settings::Dict{String, Any})
+    open(joinpath(settings["dest"], settings["config"]), "w") do f
+        println(f, "#\n# Island speciation model settings")
+        println(f, "# Run on $(Dates.format(Dates.now(), "d u Y H:M:S"))\n#\n")
+        for k in keys(settings)
+            println(f, "$k: $(settings[k])")
         end
     end
 end
@@ -223,8 +230,8 @@ end
 writes simulation output from `world` to separate table and fasta files.
 `timestep` and `setting` information is used for file name creation.
 """
-function writedata(world::Array{Patch,1}, settings::Dict{String, Any}, timestep::Int64)
-    basename = "t" * string(timestep) * "_s" * string(settings["seed"])
+function writedata(world::Array{Patch,1}, settings::Dict{String, Any}, mapfile::String, timestep::Int64)
+    basename = mapfile * "_t" * string(timestep) * "_s" * string(settings["seed"])
     basename = joinpath(settings["dest"], basename)
     filename = basename * ".tsv"
     println("Writing data \"$filename\"")
@@ -246,8 +253,8 @@ end
 writes raw julia data of the complete simulation state from `world` to file in the output directory.
[…]

### 2018-05-24 islandsim.jl
Outputted config file can now be used as input.

@@ -40,7 +40,10 @@ function simulation!(world::Array{Patch,1}, settings::Dict{String,Any}, mapfile:
 end
 
 function runit(settings::Dict{String,Any})
-    settings["seed"] == 0 ? srand() : srand(seed)
+    if settings["seed"] == 0
+        settings["seed"] = abs(rand(Int32))
+    end
+    srand(settings["seed"])
     settings["maps"] = map(x->String(x),split(settings["maps"],","))
     settings["cellsize"] *= 1e6 #convert tonnes to grams
     setupdatadir(settings)

### 2018-05-24 src/io.jl

@@ -217,9 +217,19 @@ end
 function writesettings(settings::Dict{String, Any})
     open(joinpath(settings["dest"], settings["config"]), "w") do f
         println(f, "#\n# Island speciation model settings")
-        println(f, "# Run on $(Dates.format(Dates.now(), "d u Y H:M:S"))\n#\n")
+        println(f, "# Run on $(Dates.format(Dates.now(), "d u Y HH:MM:SS"))\n#\n")
         for k in keys(settings)
-            println(f, "$k: $(settings[k])")
+            value = settings[k]
+            if isa(value, String)
+                value = '"'*value*'"'
+            elseif isa(value, Array)
+                vstr = '"'
+                for x in value
+                    vstr *= string(x)*","
+                end
+                value = vstr[1:end-1]*'"'
+            end
+            println(f, "$k $value")
         end
     end
 end

### 2018-05-24 src/subprocs.jl
patch to reflect local changes

@@ -203,7 +203,7 @@ function findposspartners(world::Array{Patch,1}, ind::Individual, location::Tupl
             mate = targetpatch[1].community[mateidx]
             mate.lineage != ind.lineage && continue #XXX Changed position for speed
             mate.age == 0 && continue #XXX obsolete (new individuals only added after reproduction)
-            !traitsexist(mate, ["repsize"]) && continue #TODO should be an error!
+            !traitsexist(mate.traits, ["repsize"]) && continue
             mate.size < mate.traits["repsize"] && continue
             mate.isnew && continue
             !iscompatible(mate, ind) && continue

### 2018-05-24 src/io.jl
getsettings() to combine settings, parseconfig no longer modifies settings

@@ -1,5 +1,13 @@
 # IO functions for GeMM
 
+function getsettings()
+    defaults = defaultSettings()
+    commandline = parsecommandline()
+    configs = parseconfig(commandline["config"])
+    merge(defaults, configs, commandline)
+end
+
+
 function parsecommandline()
     defaults = defaultSettings()
     s = ArgParseSettings()
@@ -43,10 +51,10 @@ function parsecommandline()
             help = "static mainland. Turns off any dynamics on the continent"
             action = :store_true
         end
-    return merge(defaults, parse_args(s))
+    parse_args(s)
 end
 
-function parseconfig(configfilename::String, settings::Dict{String,Any})
+function parseconfig(configfilename::String)
     # Read in the config file
     params = keys(defaultSettings())
     open(configfilename) do configfile
@@ -57,13 +65,14 @@ function parseconfig(configfilename::String, settings::Dict{String,Any})
     config = map(s -> strip(split(s, '#')[1]), config)
     config = map(split, config)
     # Parse parameters
+    settings = Dict{String, Any}()
     for c in config
         if length(c) != 2
             warn("Bad config file syntax: $c")
         elseif c[1] in params
             settings[c[1]] = parse(c[2])
         else
-            warn(c[1]*" is not a recognized parameter!")
+            warn(c[1]*" is not a recognized parameter!") # XXX maybe parse anyway
         end
     end
     return settings

### 2018-05-24 islandsim.jl
adapt to new settings parsing

@@ -40,9 +40,6 @@ function simulation!(world::Array{Patch,1}, settings::Dict{String,Any}, mapfile:
 end
 
 function runit(settings::Dict{String,Any})
-    if settings["seed"] == 0
-        settings["seed"] = abs(rand(Int32))
-    end
     srand(settings["seed"])
     settings["maps"] = map(x->String(x),split(settings["maps"],","))
     settings["cellsize"] *= 1e6 #convert tonnes to grams
@@ -59,9 +56,6 @@ function runit(settings::Dict{String,Any})
 end
 
 ## Settings
-const allargs = parsecommandline()
-if isfile(allargs["config"])
-    allargs = parseconfig(allargs["config"], allargs)
-end
+const settings = getsettings()
 
-@time runit(allargs)
+@time runit(settings)

### 2018-05-24 src/io.jl
move file verification and random seed generation to getsettings()

@@ -3,8 +3,16 @@
 function getsettings()
     defaults = defaultSettings()
     commandline = parsecommandline()
-    configs = parseconfig(commandline["config"])
-    merge(defaults, configs, commandline)
+    if isfile(commandline["config"])
+        configs = parseconfig(commandline["config"])
+    else
+        configs = Dict{String, Any}()
+    end
+    settings = merge(defaults, configs, commandline)
+    if settings["seed"] == 0
+        settings["seed"] = abs(rand(Int32))
+    end
+    settings
 end
 
 

### 2018-05-24 src/GeMM.jl
update function export

@@ -17,8 +17,7 @@ export
     # types:
     Patch,
     # functions:
-    parseconfig,
-    parsecommandline,
+    getsettings,
     setupdatadir,
     recordcolonizers,
     readmapfile,

### 2018-05-24 src/io.jl
move cellsize and maps processing to getsettings()

@@ -12,6 +12,8 @@ function getsettings()
     if settings["seed"] == 0
         settings["seed"] = abs(rand(Int32))
     end
+    settings["maps"] = map(x->String(x),split(settings["maps"],","))
+    settings["cellsize"] *= 1e6 #convert tonnes to grams
     settings
 end
 

### 2018-05-24 src/io.jl
do not require existing config file for saving

@@ -234,7 +234,7 @@ function setupdatadir(settings::Dict{String, Any})
 end
 
 function writesettings(settings::Dict{String, Any})
-    open(joinpath(settings["dest"], settings["config"]), "w") do f
+    open(joinpath(settings["dest"], "settings.conf", "w") do f
         println(f, "#\n# Island speciation model settings")
         println(f, "# Run on $(Dates.format(Dates.now(), "d u Y HH:MM:SS"))\n#\n")
         for k in keys(settings)

### 2018-05-24 src/constants.jl
empty default config setting to avoid errors + ensure type stability

@@ -16,8 +16,8 @@ const genelength = 20 # sequence length of genes
 function defaultSettings()
     Dict(# general software settings
           "seed" => 1, # for the RNG, seed = 0 -> random seed
-          "maps" => nothing, # comma-separated list of map files
-          "config" => "simulation.config", # configuration file name
+          "maps" => "", # comma-separated list of map files
+          "config" => "", # configuration file name
           "fasta" => true, # record fasta data?
           "dest" => string(Dates.today()), # output folder name
           "outfreq" => 1000, # output frequency

### 2018-05-24 src/io.jl
fix missing bracket

@@ -234,7 +234,7 @@ function setupdatadir(settings::Dict{String, Any})
 end
 
 function writesettings(settings::Dict{String, Any})
-    open(joinpath(settings["dest"], "settings.conf", "w") do f
+    open(joinpath(settings["dest"], "settings.conf"), "w") do f
         println(f, "#\n# Island speciation model settings")
         println(f, "# Run on $(Dates.format(Dates.now(), "d u Y HH:MM:SS"))\n#\n")
         for k in keys(settings)

### 2018-05-24 src/io.jl
temporarily change priorities of settings

@@ -8,7 +8,7 @@ function getsettings()
     else
         configs = Dict{String, Any}()
     end
-    settings = merge(defaults, configs, commandline)
+    settings = merge(defaults, commandline, configs) # XXX commandline should have highest priority, but overwrites with defaults
     if settings["seed"] == 0
         settings["seed"] = abs(rand(Int32))
     end

### 2018-05-25 islandsim.jl
move settings processing to getsettings()

@@ -41,8 +41,6 @@ end
 
 function runit(settings::Dict{String,Any})
     srand(settings["seed"])
-    settings["maps"] = map(x->String(x),split(settings["maps"],","))
-    settings["cellsize"] *= 1e6 #convert tonnes to grams
     setupdatadir(settings)
     for i in 1:length(settings["maps"])
         timesteps,maptable = readmapfile(settings["maps"][i])

### 2018-05-25 src/io.jl
move settings processing to getsettings(), improve flow and generality in parseconfig()

@@ -12,7 +12,7 @@ function getsettings()
     if settings["seed"] == 0
         settings["seed"] = abs(rand(Int32))
     end
-    settings["maps"] = map(x->String(x),split(settings["maps"],","))
+    settings["maps"] = map(x -> String(x), split(settings["maps"], ","))
     settings["cellsize"] *= 1e6 #convert tonnes to grams
     settings
 end
@@ -71,7 +71,9 @@ function parseconfig(configfilename::String)
         config = readlines(configfile)
     end
     # Remove comments and tokenize
-    filter!(x -> isempty(strip(x)) || (x[1] != '#'), config)
+    config = map(x -> strip(x), config)
+    filter!(x -> !isempty(x), config)
+    filter!(x -> (x[1] != '#'), config)
     config = map(s -> strip(split(s, '#')[1]), config)
     config = map(split, config)
     # Parse parameters
@@ -80,7 +82,7 @@ function parseconfig(configfilename::String)
         if length(c) != 2
             warn("Bad config file syntax: $c")
         elseif c[1] in params
-            settings[c[1]] = parse(c[2])
+            settings[c[1]] = parse(c[2]) # XXX this is dangerous regarding type stability
         else
             warn(c[1]*" is not a recognized parameter!") # XXX maybe parse anyway
         end

### 2018-05-25 src/types.jl
add lineage mapping

@@ -35,15 +35,25 @@ mutable struct Patch
     nicheb::Float64 # e.g. precipitation
     community::Array{Individual, 1}
     isolated::Bool
+    whoiswho::Dict{String, Array{Int64, 1}}
 end
 
 # constructors:
-Patch(id,location,altitude,area,isisland,nichea,nicheb,community) = Patch(id,location,altitude,area,isisland,nichea,nicheb,community,false)
-Patch(id,location,altitude,area,isisland,nichea,nicheb) = Patch(id,location,altitude,area,isisland,nichea,nicheb,Individual[],false)
-Patch(id,location,altitude,area,isisland,nichea) = Patch(id,location,altitude,area,isisland,nichea,0,Individual[],false)
-Patch(id,location,altitude,area,isisland) = Patch(id,location,altitude,area,isisland,0,0,Individual[],false)
-Patch(id,location,altitude,area) = Patch(id,location,altitude,area,false,0,0,Individual[],false)
-Patch(id,location,altitude) = Patch(id,location,altitude,100,false,0,0,Individual[],false)
-Patch(id,location) = Patch(id,location,298,100,false,0,0,Individual[],false)
+Patch(id,location,altitude,area,isisland,nichea,nicheb,community, isolated) =
+    Patch(id,location,altitude,area,isisland,nichea,nicheb,community, isolated, Dict{String, Array{Int64, 1}}())
+Patch(id,location,altitude,area,isisland,nichea,nicheb,community) =
+    Patch(id,location,altitude,area,isisland,nichea,nicheb,community, false, Dict{String, Array{Int64, 1}}())
+Patch(id,location,altitude,area,isisland,nichea,nicheb) =
+    Patch(id,location,altitude,area,isisland,nichea,nicheb,Individual[], false, Dict{String, Array{Int64, 1}}())
+Patch(id,location,altitude,area,isisland,nichea) =
+    Patch(id,location,altitude,area,isisland,nichea,0,Individual[], false, Dict{String, Array{Int64, 1}}())
+Patch(id,location,altitude,area,isisland) =
+    Patch(id,location,altitude,area,isisland,0,0,Individual[], false, Dict{String, Array{Int64, 1}}())
+Patch(id,location,altitude,area) =
+    Patch(id,location,altitude,area,false,0,0,Individual[], false, Dict{String, Array{Int64, 1}}())
+Patch(id,location,altitude) =
[…]

### 2018-05-25 src/constants.jl
Reverted default config file setting.

@@ -17,7 +17,7 @@ function defaultSettings()
     Dict(# general software settings
           "seed" => 1, # for the RNG, seed = 0 -> random seed
           "maps" => "", # comma-separated list of map files
-          "config" => "", # configuration file name
+          "config" => "simulation.conf", # configuration file name
           "fasta" => true, # record fasta data?
           "dest" => string(Dates.today()), # output folder name
           "outfreq" => 1000, # output frequency

### 2018-05-25 src/io.jl

@@ -34,6 +34,7 @@ function parsecommandline()
             help = "name of the config file"
             arg_type = String
             required = false
+            default = defaults["config"]
         "--linkage", "-l"
             help = "gene linkage (\"none\", \"random\" or \"full\")"
             arg_type = String
@@ -236,8 +237,9 @@ function setupdatadir(settings::Dict{String, Any})
 end
 
 function writesettings(settings::Dict{String, Any})
-    open(joinpath(settings["dest"], "settings.conf"), "w") do f
-        println(f, "#\n# Island speciation model settings")
+    open(joinpath(settings["dest"], settings["config"]), "w") do f
+        println(f, "#\n# --- Island speciation model settings ---")
+        println(f, "# This file was generated automatically.")
         println(f, "# Run on $(Dates.format(Dates.now(), "d u Y HH:MM:SS"))\n#\n")
         for k in keys(settings)
             value = settings[k]

### 2018-05-25 islandsim.jl
Commandline parameters now have priority over config file.

@@ -49,7 +49,7 @@ function runit(settings::Dict{String,Any})
         simulation!(world, settings, settings["maps"][i], timesteps)
         writedata(world, settings, settings["maps"][i], -1)
         println("WORLD POPULATION: $(sum(x -> length(x.community), world))") #DEBUG
-        println("WORLD MEMORY: $(round(Base.summarysize(world)/1024^2, 2)) MB") #DEBUG
+        #println("WORLD MEMORY: $(round(Base.summarysize(world)/1024^2, 2)) MB") #DEBUG
     end
 end
 

### 2018-05-25 src/io.jl

@@ -8,12 +8,12 @@ function getsettings()
     else
         configs = Dict{String, Any}()
     end
-    settings = merge(defaults, commandline, configs) # XXX commandline should have highest priority, but overwrites with defaults
+    settings = merge(defaults, configs, commandline)
     if settings["seed"] == 0
         settings["seed"] = abs(rand(Int32))
     end
     settings["maps"] = map(x -> String(x), split(settings["maps"], ","))
-    settings["cellsize"] *= 1e6 #convert tonnes to grams
+    isa(settings["cellsize"], Integer) && settings["cellsize"] *= 1e6 #convert tonnes to grams
     settings
 end
 
@@ -25,7 +25,6 @@ function parsecommandline()
         "--seed", "-s"
             help = "inital random seed"
             arg_type = Int
-            default = defaults["seed"]
         "--maps", "-m"
             help = "list of map files, comma separated"
             arg_type = String
@@ -34,35 +33,34 @@ function parsecommandline()
             help = "name of the config file"
             arg_type = String
             required = false
-            default = defaults["config"]
         "--linkage", "-l"
             help = "gene linkage (\"none\", \"random\" or \"full\")"
             arg_type = String
             range_tester = x->in(x,["none", "random", "full"])
             required = false
-            default = defaults["linkage"]
         "--nniches", "-n"
             help = "number of environmental niche traits (1 -- 3)"
             arg_type = Int
             range_tester = x -> x > 0 && x <= 3
             required = false
-            default = defaults["nniches"]
         "--tolerance", "-t"
             help = "tolerance of sequence identity when reproducing (\"high\", \"evo\", \"low\" or \"none\")"
             arg_type = String
             range_tester = x->in(x,["high", "evo", "low", "none"])
             required = false
-            default = defaults["tolerance"]
         "--dest", "-d"
             help = "output directory. Defaults to current date"
             arg_type = String
             required = false
-            default = defaults["dest"]
         "--static"
             help = "static mainland. Turns off any dynamics on the continent"
             action = :store_true
-        end
-    parse_args(s)
+    end
+    args = parse_args(s)
+    for a in keys(args)
+        (args[a] == nothing) && delete!(args, a)
+    end
+    args
 end
 
 function parseconfig(configfilename::String)
@@ -240,7 +238,7 @@ function writesettings(settings::Dict{String, Any})
[…]

### 2018-05-25 src/constants.jl
Changed cellsize to 20 tonnes (literature value)

@@ -28,7 +28,7 @@ function defaultSettings()
           "static" => true, # mainland sites don't undergo eco-evolutionary processes
           "mutate" => true, # mutations occur
           "initadults" => false, # initialize organisms as adults
-          "cellsize" => 100, # maximum biomass per cell in tonnes
+          "cellsize" => 20, # maximum biomass per hectare in tonnes (based on Clark et al. 2001)
           # invasion specific settings
           "propagule-pressure" => 0, # TODO
           "disturbance" => 0) # percentage of individuals killed per update per cell

### 2018-05-25 src/io.jl

@@ -13,7 +13,9 @@ function getsettings()
         settings["seed"] = abs(rand(Int32))
     end
     settings["maps"] = map(x -> String(x), split(settings["maps"], ","))
-    isa(settings["cellsize"], Integer) && settings["cellsize"] *= 1e6 #convert tonnes to grams
+    if isa(settings["cellsize"], Integer)
+        settings["cellsize"] *= 1e6 #convert tonnes to grams
+    end
     settings
 end
 

### 2018-05-25 src/mainprocs.jl
add identifyAdults!()

@@ -276,6 +276,7 @@ end
 Reproduction of individuals in a patch `p` whithin a world (array of patches) `w`
 """
 function reproduce!(world::Array{Patch,1}, patch::Patch) #TODO: refactorize!
+    identifyAdults!(patch)
     idx = 1
     temp = patch.altitude
     seedbank = Individual[]

### 2018-05-25 src/subprocs.jl
implement identifyAdults!()

@@ -155,6 +155,15 @@ function checkborderconditions!(world::Array{Patch,1},xdest::Float64,ydest::Floa
     xdest,ydest
 end
 
+function identifyAdults!(patch::Patch)
+    adultspeciesidx = Dict{String, Array{Int64, 1}}
+    for i in eachindex(patch.community)
+        patch.community[i].size < patch.community[i].traits["repsize"] && continue
+        adultspeciesidx = merge(append!, adultspeciesidx, Dict(patch.community[i].lineage => [i]))
+    end
+    patch.whoiswho = adultspeciesidx
+end
+
 function iscompatible(mate::Individual, ind::Individual)
     tolerance = ind.traits["reptol"]
     indgene = ""
@@ -197,7 +206,7 @@ function findposspartners(world::Array{Patch,1}, ind::Individual, location::Tupl
     while length(posspartners) == 0
         idx > length(coordinates) && break
         targetpatch = filter(l -> l.location == coordinates[idx], world)
-        length(targetpatch) >= 1 ? communityidxs = collect(eachindex(targetpatch[1].community)) : communityidxs = []
+        length(targetpatch) >= 1 ? communityidxs = collect(eachindex(targetpatch[1].community)) : communityidxs = [] # ...? communityidxs = targetpatch[1].whosiwho[]
         shuffle!(communityidxs)
         for mateidx in communityidxs
             mate = targetpatch[1].community[mateidx]

### 2018-05-25 src/subprocs.jl
adapt findposspartners() to use adult species indices

@@ -158,6 +158,7 @@ end
 function identifyAdults!(patch::Patch)
     adultspeciesidx = Dict{String, Array{Int64, 1}}
     for i in eachindex(patch.community)
+        !traitsexist(patch.community[i].traits, ["repsize"]) && continue
         patch.community[i].size < patch.community[i].traits["repsize"] && continue
         adultspeciesidx = merge(append!, adultspeciesidx, Dict(patch.community[i].lineage => [i]))
     end
@@ -206,14 +207,14 @@ function findposspartners(world::Array{Patch,1}, ind::Individual, location::Tupl
     while length(posspartners) == 0
         idx > length(coordinates) && break
         targetpatch = filter(l -> l.location == coordinates[idx], world)
-        length(targetpatch) >= 1 ? communityidxs = collect(eachindex(targetpatch[1].community)) : communityidxs = [] # ...? communityidxs = targetpatch[1].whosiwho[]
+        if length(targetpatch) >= 1 && haskey(targetpatch[1].whoiswho, ind.lineage)
+            communityidxs = (targetpatch[1].whoiswho[ind.lineage])
+        else
+            communityidxs = []
+        end
         shuffle!(communityidxs)
         for mateidx in communityidxs
             mate = targetpatch[1].community[mateidx]
-            mate.lineage != ind.lineage && continue #XXX Changed position for speed
-            mate.age == 0 && continue #XXX obsolete (new individuals only added after reproduction)
-            !traitsexist(mate.traits, ["repsize"]) && continue
-            mate.size < mate.traits["repsize"] && continue
             mate.isnew && continue
             !iscompatible(mate, ind) && continue
             push!(posspartners, mate)

### 2018-05-25 src/subprocs.jl
fix empty declaration

@@ -156,7 +156,7 @@ function checkborderconditions!(world::Array{Patch,1},xdest::Float64,ydest::Floa
 end
 
 function identifyAdults!(patch::Patch)
-    adultspeciesidx = Dict{String, Array{Int64, 1}}
+    adultspeciesidx = Dict{String, Array{Int64, 1}}()
     for i in eachindex(patch.community)
         !traitsexist(patch.community[i].traits, ["repsize"]) && continue
         patch.community[i].size < patch.community[i].traits["repsize"] && continue

### 2018-05-25 src/io.jl
Cut `basicparser` from `parseconfig`, to be reused in `readmapfile`.

@@ -65,30 +65,55 @@ function parsecommandline()
     args
 end
 
-function parseconfig(configfilename::String)
-    # Read in the config file
-    params = keys(defaultSettings())
-    open(configfilename) do configfile
-        config = readlines(configfile)
+"""
+    basicparser(f)
+Do elementary parsing on a config or map file.
+
+Reads in the file, strips whole-line and inline comments
+and separates lines by whitespace.
+Returns a 2d array representing the tokens in each line.
+"""
+function basicparser(filename::String)
+    # Read in the file
+    println("Reading file \"$filename\"...")
+    lines = String[]
+    open(filename) do file
+        lines = readlines(file)
     end
     # Remove comments and tokenize
-    config = map(x -> strip(x), config)
-    filter!(x -> !isempty(x), config)
-    filter!(x -> (x[1] != '#'), config)
-    config = map(s -> strip(split(s, '#')[1]), config)
-    config = map(split, config)
-    # Parse parameters
+    lines = map(x -> strip(x), lines)
+    filter!(x -> !isempty(x), lines)
+    filter!(x -> (x[1] != '#'), lines)
+    lines = map(s -> strip(split(s, '#')[1]), lines)
+    lines = map(split, lines)
+    map(l -> map(s -> convert(String, s), l), lines)
+end
+
+function parseconfig(configfilename::String)
+    config = basicparser(configfilename)
     settings = Dict{String, Any}()
+    defaults = defaultSettings()
     for c in config
         if length(c) != 2
             warn("Bad config file syntax: $c")
-        elseif c[1] in params
-            settings[c[1]] = parse(c[2]) # XXX this is dangerous regarding type stability
+        elseif c[1] in keys(defaults)
+            value = parse(c[2])
+            if !isa(value, typeof(defaults[c[1]]))
+                warn("$(c[1]): expected $(typeof(settings[c[1]])), got $(typeof(value)).")
+            end
+            settings[c[1]] = value
         else
             warn(c[1]*" is not a recognized parameter!") # XXX maybe parse anyway
         end
     end
     return settings
+end
+
+function readmapfilenew(mapfilename::String)
+    mapfile = basicparser(mapfilename)
+    timesteps = parse(mapfile[1][1])
+    !isa(timesteps, Integer) && error("Invalid timestep argument in the mapfile: $timesteps")
+    #TODO
 end
 
 function readmapfile(filename::String)

### 2018-05-25 src/io.jl
Rewrote `readmapfile`.

@@ -109,32 +109,14 @@ function parseconfig(configfilename::String)
     return settings
 end
 
-function readmapfilenew(mapfilename::String)
+function readmapfile(mapfilename::String)
     mapfile = basicparser(mapfilename)
     timesteps = parse(mapfile[1][1])
-    !isa(timesteps, Integer) && error("Invalid timestep argument in the mapfile: $timesteps")
-    #TODO
-end
-
-function readmapfile(filename::String)
-    println("Reading file \"$filename\"...")
-    mapstrings = String[]
-    open(filename) do file
-        mapstrings = readlines(file)
-    end
-    mapstrings = filter(x->!all(isspace,x),mapstrings) # remove empty lines
-    mapstrings = filter(x->x[1]!='#',mapstrings) # remove comment lines
-    mapsubstrings = map(split,mapstrings)
-    mapentries = map(x->map(String,x),mapsubstrings)
-    timesteps = 0
-    try
-        timesteps = parse(Int,filter(x->size(x,1)==1,mapentries)[1][1])
-    catch
+    if length(mapfile[1]) != 1 || !isa(timesteps, Integer)
         timesteps = 1000
-        warn("your mapfile \"$filename\" does not include timestep information. Assuming $timesteps timesteps.")
+        warn("Invalid timestep information in the mapfile. Setting timesteps to 1000.")
     end
-    mapentries = filter(x->size(x,1)>1,mapentries)
-    return timesteps,mapentries
+    return timesteps,mapfile[2:end]
 end
 
 """

### 2018-05-25 paper/draft.tex
comments, can input citations via betterbibtex/zotero
include some proper refs

@@
 To this date, however, it is unclear why these genera produced this multitude of species as opposed to most other genera that remained species poor.
-Until recently, researchers focused on environmental and ecological factors to explain different speeds and ocurrences of speciation.
+Until recently, researchers focused on environmental and ecological factors to explain different speeds and ocurrences of speciation.\cite{krugSpeciesGenusRatios2008}\cite{hurlbertProcessesGeneratingLatitudinal2014}\cite{willisSpeciesDiversityScale2002}
 %% HERE COME SOME EXAMPLES
@@
 With the growing importance of genetic data and analyses, an increasing number of studies also investigates genetic processes and
-their effect on population variality. %EXAMPLES
-Some of these studies highlight the importance of genetic linkage or ``hitchhiking'' on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein)
-as well as evolutionary rescue provided by the feedback with ecological processes (Schiffers et al. 2013).
+their effect on population variability. %EXAMPLES
+Some of these studies highlight the importance of genetic linkage or ``hitchhiking'' on reproductive/ecological isolation of populations \cite{hawthorneGeneticLinkageEcological2001}\cite{viaGeneticMosaicSuggests2008}%(e.g., Via \& West 2008, Via 2001 and references therein)
+as well as evolutionary rescue provided by the feedback with ecological processes \cite{schiffersLimitedEvolutionaryRescue2012}.
 %%TODO: here come a list of processes with examples
@@
 %% - isolation processes
-Besides arthropods (Howthorne \& Via 2001), these processes can be observed in plants as well.
+Besides arthropods \cite{hawthorneGeneticLinkageEcological2001}, these processes can be observed in plants as well.
 %Are you talking about islands here?
@@
 For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
-seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al.\ 2013). % cite properly!
+seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits \cite{hermannTightGeneticLinkage2013}
 In an island context, genetic studies are even rarer and almost exclusively of a descriptive nature.
@@
 Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics
-like isolation (e.g. Losos \& Ricklefs 2009; Warren et al. 2015  %%Islands as model systems in ecology and evolution: prospects fifty years after MacArthur‐Wilson. EcoLett). %TODO: ref
+like isolation (e.g. \cite{lososAdaptationDiversificationIslands2009b}; \cite{warrenIslandsModelSystems}.
 Part of this value as study system originates by typically remarkable species.
 In fact, islands often possess unique species communities and can exceede mainland areas
-by a factor of 9.5 and 8.1 in plant and animal  endemism, respectively
-(Kier et al. 2005. %A global assessment of endemism and species richness across island and mainland regions. PNAS). %TODO: refs
-Such high endemism makes islands  especially threatened by human impacts, in particular climate change and invasive species (Fordham \& Brook 2010). %TODO: refs
+by a factor of 9.5 and 8.1 in plant and animal  endemism, respectively \cite{kierGlobalAssessmentEndemism2009}
+Such high endemism makes islands  especially threatened by human impacts, in particular climate change and invasive species \cite{fordhamWhyTropicalIsland2010}
 Hence, understanding the origin, maintenance, functioning and evolution of island communities is of a high priority objective.
@@
 
+%Clonality
+
 Shortcomings. (/Processes/Factors not considered.)

### 2018-07-03 paper/draft.tex
include additional thoughts and refs in intro

@@
 %% Hurlbert
+% !! Hypothesis: Quick establishment of reproductive barriers (+ population variability) facilitates/speeds up adaptation. !!
 
@@
 Island radiations like the Darwin finches on Galápagos or \textit{Aeonium} on the Canaries have fascinated researchers throughout history %% cite? Darwin? ...?
-To this date, however, it is unclear why these genera produced this multitude of species as opposed to most other genera that remained species poor.
-Until recently, researchers focused on environmental and ecological factors to explain different speeds and ocurrences of speciation.\cite{krugSpeciesGenusRatios2008}\cite{hurlbertProcessesGeneratingLatitudinal2014}\cite{willisSpeciesDiversityScale2002}
+However, it is often unclear why these genera produced this multitude of species as opposed to most other genera that remained species poor.
+To explain speed and occurences of radiations, researchers still mainly focus on environmental and ecological factors \cite{krugSpeciesGenusRatios2008}\cite{hurlbertProcessesGeneratingLatitudinal2014}\cite{willisSpeciesDiversityScale2002}.
+% when radiations DID happen!
+While these factors often correlate strongly with genetic divergence of species and populations, % REFS!
+in some cases the reason for maintenance of population diversity remains in the dark. \cite{chenConcordanceGeneticDiversity}
 %% HERE COME SOME EXAMPLES

### 2018-07-03 islandsim.jl
Added `burn-in` setting to control invasion start.

@@ -33,7 +33,7 @@ function simulation!(world::Array{Patch,1}, mapfile::String, timesteps::Int=1000
         compete!(world, settings["static"])
         reproduce!(world)
         settings["mutate"] && mutate!(world)
-        invade!(world)
+        settings["burn-in"] < t && invade!(world)
         colonizers = disperse!(world, settings["static"])
         # model output
         map(p -> simlog("Patch $(p.id): $(length(p.community)) individuals.", 'd'), world)

### 2018-07-03 src/constants.jl

@@ -39,6 +39,7 @@ function defaultSettings()
          "propagule-pressure" => 0, # number of non-native individuals introduced per invasion event
          "global-species-pool" => 0, # size of the global species pool (invasion source)
          "disturbance" => 0, # percentage of individuals killed per update per cell
+         "burn-in" => 1000, # timesteps before invasion starts
          # species trait names
          "traitnames" => ["compat",
                           "dispmean",

### 2018-07-03 paper/draft.tex
add remark to species concepts in disc

@@
 
+It should be noted that observed speciation and radiation events hinge upon arbitrary species concepts.
+Increasingly more available population genome data make apparent that population genetic structure does not
+always mirror classic systematics.
+Therefore, we expect future studies to continue focussing on population level variablity when conducting genetic analyses.
 \printbibliography

### 2018-07-03 paper/draft.tex
changes to intro

@@
 % !! Hypothesis: Quick establishment of reproductive barriers (+ population variability) facilitates/speeds up adaptation. !!
+% Rationale: If genetic factors have impact on systems with clearly defined isolation (islands),
+% they should be even more important on continental systems.
 
 %% major topic: besides classical mechanisms (ecological, environmental), genetic mechanisms affect speciation/radiations
-Island radiations like the Darwin finches on Galápagos or \textit{Aeonium} on the Canaries have fascinated researchers throughout history %% cite? Darwin? ...?
-However, it is often unclear why these genera produced this multitude of species as opposed to most other genera that remained species poor.
-To explain speed and occurences of radiations, researchers still mainly focus on environmental and ecological factors \cite{krugSpeciesGenusRatios2008}\cite{hurlbertProcessesGeneratingLatitudinal2014}\cite{willisSpeciesDiversityScale2002}.
+Island radiations like the classical Darwin finches on Galápagos or \textit{Aeonium} on the Canaries have fascinated researchers throughout history. %% cite? Darwin? ...?
+However, it is often unclear why these genera produced this multitude of species as opposed to most other genera that remain species poor.
+In islands, as well as continental and aquatic systems, researchers still mainly focus on environmental and ecological factors to explain speed and occurences of radiations \cite{krugSpeciesGenusRatios2008}\cite{hurlbertProcessesGeneratingLatitudinal2014}\cite{willisSpeciesDiversityScale2002}.
 % when radiations DID happen!
-While these factors often correlate strongly with genetic divergence of species and populations, % REFS!
-in some cases the reason for maintenance of population diversity remains in the dark. \cite{chenConcordanceGeneticDiversity}
+In fact, these factors often correlate strongly with the timing of genetic divergence of species and populations, % REFS!
+still, in some cases the reason for maintenance of population diversity remains in the dark. \cite{chenConcordanceGeneticDiversity}
 %% HERE COME SOME EXAMPLES
-Yet, these mechanisms often fail to explain how the rapid establishment of reproductive barriers or isolation arises,
-which ultimately results in sympatric speciation. %CLAIM! PROVIDE REFERENCES!
+Additionally, these studies highlight the cases of successful radiations, while a conclusive explanation for why these radiations could occur in the first place is still missing.
+This becomes especially apparent on islands, where species generally experience nearly identical environmental conditions.%REF
+%mechanisms often fail to explain how the rapid establishment of reproductive barriers or isolation arises,
+%which ultimately results in sympatric speciation. %CLAIM! PROVIDE REFERENCES!
 %% somewhere here Patino et al 2018 anagenesis, cladogenesis, etc islands
-With the growing importance of genetic data and analyses, an increasing number of studies also investigates genetic processes and
+In this context, genetic data and analyses gain growing importance,
+with an increasing number of studies also investigating genetic processes and
 their effect on population variability. %EXAMPLES
@@
 The problem of investigating these processes is the time frame and study extent needed for a proper evaluation.
-Because islands biogeography takes place at evolutionary time frames (millions of years),
+Because island biogeography takes place at evolutionary time frames (millions of years),
 comprehensive and meaningful data is difficult and virtually impossible to obtain.

### 2018-07-03 paper/draft.tex
edit intro

@@
 \section{Introduction}
-%% Intro subsections (==paragraphs):
-%% 1. Major topic (include only background relevant for exp. design)
-%% 2. Islands
-%% 3. Forces
-%% 4. Models
-%% 5. Aim, Objectives
-%% 2+3 may be combined, as well as 1+2(+3)
 
@@
 %% somewhere here Patino et al 2018 anagenesis, cladogenesis, etc islands
+%% Islands:
+%% ========
+%% Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics
+%% like isolation (e.g. \cite{lososAdaptationDiversificationIslands2009b}; \cite{warrenIslandsModelSystems}.
+%% Part of this value as study system originates by typically remarkable species.
+%% In fact, islands often possess unique species communities and can exceede mainland areas
+%% by a factor of 9.5 and 8.1 in plant and animal  endemism, respectively \cite{kierGlobalAssessmentEndemism2009}
+%% Such high endemism makes islands  especially threatened by human impacts, in particular climate change and invasive species \cite{fordhamWhyTropicalIsland2010}
+%% Hence, understanding the origin, maintenance, functioning and evolution of island communities is of a high priority objective.
+%% [Move this maybe to model section:]
+%% For the prediction of island community size, several theoretical frameworks have been proposed,
+%% including one of the first and to this day still highly influential classical ecological theories,
+%% namely MacArthur and Wilson's equilibrium theory of island biogeography (ETIB). %TODO: ref
+%% In it, species richness on islands is the result of opposing rates of immigration and extinction,
+%% both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
+%% This relationship was since successfully applied to a multitude of islands (Lomolina \& Weiser 2001, Triantis et al.\ 2011)%TODO: refs
+%% , but also island like systems, such as coral reefs, lakes/ponds or forest fragments (Kelly et al.\ 1989, Peay et al.\ 2007, ). %TODO: refs
+%% This general applicability (Lomolino 2000), however, cannot hide the fact
+%% that the ETIB fails to formalize a mechanistic explanation of how extinction and immigration are modified by area and isolation, respectively.
+%% Additionally, it has been demonstrated more recently that area also affects immigration via target effects (REFs) a
+%% nd isolation affects extinction via rescue-effects (REFs).
+%% Finally, ETIB has particularly been criticized for its lack of considering speciation (see e.g.\ Losos \& Schluter 2000). %TODO: refs
+%% About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM, Whittaker et al.\ 2007, 2008). %TODO: ref
+%% The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
+%% whith islands emerging from sea level via volcanic hotspot activity,
+%% increasing over time in size and topographic complexity,
+%% then eroding and subsiding in the ocean again when the island is not over the hotspot anymore.
+%% In doing so, the model incorporates concepts of niche theory, speciation and species succession.
+%% Yet, there are still many processes and drivers deemed important,
+%% which are not considered by any formal theory and often little investigated
+%% %[if you say this, you have to mention at leastr a couple os examples, otherwise it is an empty claim].
 In this context, genetic data and analyses gain growing importance,
@@
 
-%% Islands
-Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics
-like isolation (e.g. \cite{lososAdaptationDiversificationIslands2009b}; \cite{warrenIslandsModelSystems}.
-Part of this value as study system originates by typically remarkable species.
-In fact, islands often possess unique species communities and can exceede mainland areas
-by a factor of 9.5 and 8.1 in plant and animal  endemism, respectively \cite{kierGlobalAssessmentEndemism2009}
-Such high endemism makes islands  especially threatened by human impacts, in particular climate change and invasive species \cite{fordhamWhyTropicalIsland2010}
-Hence, understanding the origin, maintenance, functioning and evolution of island communities is of a high priority objective.
-For the prediction of island community size, several theoretical frameworks have been proposed,
-including one of the first and to this day still highly influential classical ecological theories,
-namely MacArthur and Wilson's equilibrium theory of island biogeography (ETIB). %TODO: ref
-In it, species richness on islands is the result of opposing rates of immigration and extinction,
-both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
-This relationship was since successfully applied to a multitude of islands (Lomolina \& Weiser 2001, Triantis et al.\ 2011)%TODO: refs
-, but also island like systems, such as coral reefs, lakes/ponds or forest fragments (Kelly et al.\ 1989, Peay et al.\ 2007, ). %TODO: refs
-This general applicability (Lomolino 2000), however, cannot hide the fact
-that the ETIB fails to formalize a mechanistic explanation of how extinction and immigration are modified by area and isolation, respectively.
-Additionally, it has been demonstrated more recently that area also affects immigration via target effects (REFs) a
-nd isolation affects extinction via rescue-effects (REFs).
-Finally, ETIB has particularly been criticized for its lack of considering speciation (see e.g.\ Losos \& Schluter 2000). %TODO: refs
-About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM, Whittaker et al.\ 2007, 2008). %TODO: ref
-The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
-whith islands emerging from sea level via volcanic hotspot activity,
-increasing over time in size and topographic complexity,
-then eroding and subsiding in the ocean again when the island is not over the hotspot anymore.
-In doing so, the model incorporates concepts of niche theory, speciation and species succession.
-Yet, there are still many processes and drivers deemed important,
-which are not considered by any formal theory and often little investigated
-%[if you say this, you have to mention at leastr a couple os examples, otherwise it is an empty claim].
-
 %% Models
-The problem of investigating these processes is the time frame and study extent needed for a proper evaluation.
-Because island biogeography takes place at evolutionary time frames (millions of years),
+The problem of investigating speciation processes is the time frame and study extent needed for a proper evaluation.
+Because island biogeography processes and thus radiations generally take place at evolutionary time frames (millions of years),
 comprehensive and meaningful data is difficult and virtually impossible to obtain.
-Therefore, there has been an interest in building predictive models for islands.
-In a recent review, we indeed found a rapid increase in number of mechanistic simulaiton models for island biogeography %\cite{leidinger2017biodiversity}
+Therefore, there has been an interest in building process-based island models to test the causality of assumptions and hypotheses.
+In a recent review, we indeed found a rapid increase in number of mechanistic simulation models for island biogeography \cite{leidingerBiodiversityDynamicsIslands2017}
 (Leidinger \& Sarmento Cabral 2017).
 Although many of the reviewed models integrate multiple processes at a time, unfortunately,
-most models failed to include microevolutionary or more explicit genetic processes %\cite{leidinger2017biodiversity}
+most models failed to include micro-evolutionary or more explicit genetic processes %\cite{leidinger2017biodiversity}
 (Leidinger \& Sarmento Cabral 2017).
-One direction towards this is given by models using concepts such as ``protracted speciaton'' (Rosindell \& Harmon 2011) %\cite{rosindellXXX},
+One direction towards this is given by models using concepts such as ``protracted speciaton'' \cite{rosindellUnifiedModelSpecies2013}(Rosindell \& Harmon 2011)
 where speciation occurs with a delay from the actual
 divergence event if the population remains isolated during that time.
-But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005)
+But nature is usually more complex than this (cf.\ \cite{silvertownGhostCompetitionPhylogeny2004} Silvertown 2004, but see \cite{herbenGhostHybridizationNiche2005} Herben et al.\ 2005)
 and genetic mechanisms should also play a central role for speciation/radiation and species richness.
-Questions, such as "[cite two questions from Patino et al. 2017]", thus remain largely open (Patiño et al.\ 2017)
+Questions, such as "What functional traits (e.g. relating to dispersal capacity,
+reproduction, trophic ecology) are associated with high diversification rates within and across island systems?"
+or "What is the influence of gene flow among islands and/or
+between islands and mainland areas on speciation rates?", thus remain largely open \cite{patinoRoadmapIslandBiology2017}(Patiño et al.\ 2017)
 and may require explicit consideration of evolutionary processes.
@@
 while genetic properties are hardly recognized in the island context.
-This is surprising given the essential role islands have played in the development of evolutionary theories.
+Given the essential role islands have played in the development of evolutionary theories, it is surprising that these processes
+are not implemented more often in island simulations.
 
@@
 %% In this study, we aim to address...
-To address this hypothesis, 
-we present a mechanistic model that overcomes above-mentioned shortcomings (LIST THEM HERE IN SHORT)
+To assess the role of genetic processes, namely genetic linkage, for island radiations,
+we present a mechanistic model that overcomes above-mentioned shortcomings of missing micro-evolutinary processes and lack of integrative design %(LIST THEM HERE IN SHORT)
 by implementing genetic processes on a niche-explicit background

### 2018-07-04 paper/draft.tex
come up with flow for first part of intro

@@
 %% major topic: besides classical mechanisms (ecological, environmental), genetic mechanisms affect speciation/radiations
-Island radiations like the classical Darwin finches on Galápagos or \textit{Aeonium} on the Canaries have fascinated researchers throughout history. %% cite? Darwin? ...?
-However, it is often unclear why these genera produced this multitude of species as opposed to most other genera that remain species poor.
-In islands, as well as continental and aquatic systems, researchers still mainly focus on environmental and ecological factors to explain speed and occurences of radiations \cite{krugSpeciesGenusRatios2008}\cite{hurlbertProcessesGeneratingLatitudinal2014}\cite{willisSpeciesDiversityScale2002}.
+Island radiations like the classical Darwin finches on Galápagos or \textit{Aeonium} on the Canaries have fascinated researchers throughout history. %% Maybe other examples?
+But why were these genera able to produce this multitude of species as opposed to most other genera that remain species poor?
+For a population to produce separate species, that population must
+(a) have sufficient variability in any one characteristic, e.g. traits or genome, and
+(b) have reproductive barriers (e.g. environmental/geographical, ecological or behavioral) that facilitate establishment of distinct distribution patterns of said characteristic.
+In islands, as well as continental and aquatic systems, researchers still mainly focus on reproductive barriers like environmental and ecological factors to explain speed and occurences of radiations \cite{krugSpeciesGenusRatios2008}\cite{hurlbertProcessesGeneratingLatitudinal2014}\cite{willisSpeciesDiversityScale2002}.
 % when radiations DID happen!
-In fact, these factors often correlate strongly with the timing of genetic divergence of species and populations, % REFS!
-still, in some cases the reason for maintenance of population diversity remains in the dark. \cite{chenConcordanceGeneticDiversity}
+In fact, these factors indeed often correlate strongly with the timing of genetic divergence of species and populations. % REFS!
+But even then, in some cases the reason for maintenance of population diversity remains in the dark. \cite{chenConcordanceGeneticDiversity}
 %% HERE COME SOME EXAMPLES
-Additionally, these studies highlight the cases of successful radiations, while a conclusive explanation for why these radiations could occur in the first place is still missing.
+Additionally, these studies highlight the cases of successful radiations, while a conclusive explanation for why these population variablities and thus radiations could occur in the first place is still missing.
 This becomes especially apparent on islands, where species generally experience nearly identical environmental conditions.%REF
+Yet, only a fraction of them manage to radiate into multiple species through the course of their evolutionary history.
 %mechanisms often fail to explain how the rapid establishment of reproductive barriers or isolation arises,
@@
 %% %[if you say this, you have to mention at leastr a couple os examples, otherwise it is an empty claim].
-In this context, genetic data and analyses gain growing importance,
-with an increasing number of studies also investigating genetic processes and
-their effect on population variability. %EXAMPLES
+A promising avenue to this with growing importance is genetic data and analyses.
+Genetic analyses have already helped clearing up phylogenies of many species groups %, like bats or cichlids REFS!
+and more and more complete genomes are published on a regular basis.
 Some of these studies highlight the importance of genetic linkage or ``hitchhiking'' on reproductive/ecological isolation of populations \cite{hawthorneGeneticLinkageEcological2001}\cite{viaGeneticMosaicSuggests2008}%(e.g., Via \& West 2008, Via 2001 and references therein)

### 2018-07-04 paper/draft.tex
add refs, clean up

@@
 % when radiations DID happen!
-In fact, these factors indeed often correlate strongly with the timing of genetic divergence of species and populations. % REFS!
+In fact, these factors indeed often correlate strongly with the timing of genetic divergence of species and populations \cite{hembryEvolutionaryBiogeographyTerrestrial}. % REFS!
 But even then, in some cases the reason for maintenance of population diversity remains in the dark. \cite{chenConcordanceGeneticDiversity}
@@
 This becomes especially apparent on islands, where species generally experience nearly identical environmental conditions.%REF
-Yet, only a fraction of them manage to radiate into multiple species through the course of their evolutionary history.
+Yet, only a fraction of them manage to radiate into multiple species through the course of their evolutionary history \cite{hembryEvolutionaryBiogeographyTerrestrial}.
 %mechanisms often fail to explain how the rapid establishment of reproductive barriers or isolation arises,
@@
 A promising avenue to this with growing importance is genetic data and analyses.
-Genetic analyses have already helped clearing up phylogenies of many species groups %, like bats or cichlids REFS!
-and more and more complete genomes are published on a regular basis.
-Some of these studies highlight the importance of genetic linkage or ``hitchhiking'' on reproductive/ecological isolation of populations \cite{hawthorneGeneticLinkageEcological2001}\cite{viaGeneticMosaicSuggests2008}%(e.g., Via \& West 2008, Via 2001 and references therein)
+Genetic analyses have already helped clearing up phylogenies of many species groups \cite{valenteUsingMolecularPhylogenies2018}%, like bats or cichlids REFS!
+and more and more complete genomes are published on a regular basis \cite{alonso-blanco135GenomesReveal2016} \cite{lewinEarthBioGenomeProject2018}.
+Some genetic studies highlight the importance of genetic linkage or ``hitchhiking'' on reproductive/ecological isolation of populations \cite{hawthorneGeneticLinkageEcological2001}\cite{viaGeneticMosaicSuggests2008}
 as well as evolutionary rescue provided by the feedback with ecological processes \cite{schiffersLimitedEvolutionaryRescue2012}.
@@
 Therefore, there has been an interest in building process-based island models to test the causality of assumptions and hypotheses.
-In a recent review, we indeed found a rapid increase in number of mechanistic simulation models for island biogeography \cite{leidingerBiodiversityDynamicsIslands2017}
-(Leidinger \& Sarmento Cabral 2017).
+In a recent review, we indeed found a rapid increase in number of mechanistic simulation models for island biogeography \cite{leidingerBiodiversityDynamicsIslands2017}.
 Although many of the reviewed models integrate multiple processes at a time, unfortunately,
@@
 (Leidinger \& Sarmento Cabral 2017).
-One direction towards this is given by models using concepts such as ``protracted speciaton'' \cite{rosindellUnifiedModelSpecies2013}(Rosindell \& Harmon 2011)
+One direction towards this is given by models using concepts such as ``protracted speciaton'' \cite{rosindellUnifiedModelSpecies2013}.
 where speciation occurs with a delay from the actual
 divergence event if the population remains isolated during that time.
-But nature is usually more complex than this (cf.\ \cite{silvertownGhostCompetitionPhylogeny2004} Silvertown 2004, but see \cite{herbenGhostHybridizationNiche2005} Herben et al.\ 2005)
+But nature is usually more complex than this (cf.\ \cite{silvertownGhostCompetitionPhylogeny2004} but see \cite{herbenGhostHybridizationNiche2005}) 
 and genetic mechanisms should also play a central role for speciation/radiation and species richness.
@@
 or "What is the influence of gene flow among islands and/or
-between islands and mainland areas on speciation rates?", thus remain largely open \cite{patinoRoadmapIslandBiology2017}(Patiño et al.\ 2017)
+between islands and mainland areas on speciation rates?", thus remain largely open \cite{patinoRoadmapIslandBiology2017}
 and may require explicit consideration of evolutionary processes.
@@
 To assess the role of genetic processes, namely genetic linkage, for island radiations,
-we present a mechanistic model that overcomes above-mentioned shortcomings of missing micro-evolutinary processes and lack of integrative design %(LIST THEM HERE IN SHORT)
+we present a mechanistic model that overcomes above-mentioned shortcomings like missing micro-evolutinary processes and lack of integrative design
 by implementing genetic processes on a niche-explicit background
-using an island biogeography simulation arena.%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
-%% and what trait combinations constitute successful colonizer and radiating species, respectively.
+using an island biogeography simulation arena.
 We will address the questions of what relative impact does (a) genetic linkage and
-(b) reproductive isolation have on emerging species richenss and composition,
+(b) reproductive isolation have on emerging species richness and composition,
 and (c) which genetic and functional trait syndromes %% Which?

### 2018-07-04 analyse.R
Modified experiment.py to circumvent slurm.

@@ -24,10 +24,10 @@ plotTraits = function(timestep=-1, toFile=TRUE) {
     traitfilepath = paste(outdir, traitfile, sep="/")
     ts = read.table(traitfilepath, header=T)
     jpeg(paste(outdir, "/", simname, "_traits.jpg", sep=""), height=720, width=1800)
-    boxplot(ts$fitness*10, log(ts$size), log(ts$seedsize), log(ts$repsize), ts$lnkgunits/10,
+    boxplot(ts$fitness*100, log(ts$size), log(ts$seedsize), log(ts$repsize), ts$lnkgunits/10,
             ts$ngenes/10, ts$temptol, ts$tempopt/100, ts$prectol, ts$precopt, ts$compat*10,
             ts$repradius*10, ts$dispmean, ts$dispshape,
-            names=c("Fitness (x10)", "log(Size)", "log(Seed size)", "log(Reprod. size)",
+            names=c("Fitness (x100)", "log(Size)", "log(Seed size)", "log(Reprod. size)",
                     "Chromosome (x0.1)", "Genes (x0.1)", "Temp. tolerance",
                     "Temp. opt. (x0.01)", "Precip. tolerance",
                     "Precip. optimum", "Compatibility (x10)", "Rep. radius (x10)",

### 2018-07-04 experiment.py

@@ -4,7 +4,7 @@
 # Daniel Vedder, 19/6/2018
 #
 
-import os, sys, time
+import os, sys, time, random
 
 global simname, replicates
 
@@ -20,9 +20,8 @@ if len(sys.argv) >= 3:
     replicates = int(sys.argv[2])
 
 # These settings stay constant throughout all simulation runs
-constant_settings = {"seed":0,
-                     "maps":'"invasion.map"',
-                     "outfreq":10,
+constant_settings = {"maps":'"invasion.map"',
+                     "outfreq":20,
                      "logging":"true",
                      "debug":"false",
                      "fasta":"false",
@@ -34,33 +33,38 @@ constant_settings = {"seed":0,
                      "static":"false",
                      "mutate":"false",
                      "initadults":"true",
-                     "metabolicpopsize":"false",
+                     "initpopsize":'"bodysize"',
+                     "burn-in": 1000,
                      "global-species-pool":100}
 
 # These settings are varied (the first value is the default,
 # every combination of the rest is tested)
-varying_settings = {"cellsize":[20, 10, 100],
+varying_settings = {"cellsize":[20, 10, 50],
                     "propagule-pressure":[0,1,10],
                     "disturbance":[0,1,10]}
 
 def slurm(config):
     "Send a job to slurm"
-    cmd = "sbatch -c 2 --mem 50GB ./islandsim.jl --config "+str(config)
-    os.system(cmd)
-    time.sleep(3) #prevent output folder merges
+    if "gaia" in os.uname().nodename:
+        cmd = "sbatch -c 2 --mem 50GB ./islandsim.jl --config "+str(config)
+        os.system(cmd)
+        os.system("rm "+config) #cleanup
+        time.sleep(3) #prevent output folder merges
+    else:
+        print("Not on gaia, slurm is probably not available. Retaining job "+config)
 
-def write_config(config, cellsize, prop_pressure, disturbance):
+def write_config(config, cellsize, prop_pressure, disturbance, seed):
     "Write out a config file with the given values"
-    global simname
     cf = open(config, 'w')
     cf.write("# Island speciation model for invasion experiments\n")
     cf.write("# This config file was generated automatically.\n")
     cf.write("# "+time.asctime()+"\n")
-    cf.write('\ndest "results/'+simname+'"\n')
+    cf.write('\ndest "results/'+config.replace(".conf", "")+'"\n')
     cf.write("\n# Constant settings:\n")
     for k in constant_settings.keys():
[…]

### 2018-07-04 src/constants.jl

@@ -27,7 +27,7 @@ function defaultSettings()
          "quiet" => false, # don't write output to screen
          "debug" => false, # write out debug statements
          # main model settings
-         "linkage" => "random", # gene linkage type
+         "linkage" => "random", # gene linkage type (random/full/none)
          "nniches" => 2, # number of environmental niches (max. 3)
          "tolerance" => "low", # sequence similarity threshold for reproduction
          "static" => true, # mainland sites don't undergo eco-evolutionary processes

### 2018-07-04 src/init.jl

@@ -16,7 +16,7 @@ function genesis()
             quarterpopsize = Integer(floor((settings["cellsize"] / newind.traits["repsize"]) / 4))
             popsize = rand(0:quarterpopsize)
         elseif contains(settings["initpopsize"], "minimal")
-            popsize = 2 #Takes two to tangle ;-)
+            popsize = 2 #Takes two to tangle ;-) #XXX No reproduction occurs!
         else
             simlog("Invalid value for `initpopsize`: $(settings["initpopsize"])", 'e')
         end

### 2018-07-04 src/io.jl

@@ -370,7 +370,7 @@ function recordstatistics(world::Array{Patch,1})
     lineages = unique(reduce(vcat, map(p -> collect(keys(p.whoiswho)), world)))
     div = round.(diversity(world),3)
     space = freespace(world)
-    simlog("Population size: $popsize")
+    simlog("Population size: $popsize, lineages: $(length(lineages))")
     simlog("$popsize,$space,$(length(lineages)),$(div[1]),$(div[2]),$(div[3])",
            'i', "diversity.log", true)
 end

### 2018-07-05 paper/draft.tex
more edits to intro, include refs

@@
 
-%% Cite & look at Katja Schiffer paper
-%% "species rich vs. species poor lineage''
+%% Q24. What is the relative importance of ecological versus geographical speciation on islands?
+%% => Patino hardly considering genetic mechanisms
 %% ``emerging systematic, functional, phylogenetic patterns''
 %% Hurlbert
-% !! Hypothesis: Quick establishment of reproductive barriers (+ population variability) facilitates/speeds up adaptation. !!
+
 % Rationale: If genetic factors have impact on systems with clearly defined isolation (islands),
@@
 %% major topic: besides classical mechanisms (ecological, environmental), genetic mechanisms affect speciation/radiations
-Island radiations like the classical Darwin finches on Galápagos or \textit{Aeonium} on the Canaries have fascinated researchers throughout history. %% Maybe other examples?
-But why were these genera able to produce this multitude of species as opposed to most other genera that remain species poor?
+To this day, island radiations like the classical Darwin finches or Galápagos tortoises %%or \textit{Aeonium} on the Canaries
+still engage researchers throughout the world \cite{roman-palaciosTortoiseFinchTesting}.
+But why were these lineages able to produce this multitude of species as opposed to most other lineages that remain species poor?
 For a population to produce separate species, that population must
 (a) have sufficient variability in any one characteristic, e.g. traits or genome, and
-(b) have reproductive barriers (e.g. environmental/geographical, ecological or behavioral) that facilitate establishment of distinct distribution patterns of said characteristic.
+(b) have gene-flow restricting factors (e.g. environmental/geographical, ecological or behavioral) that facilitate establishment of distinct distribution patterns of said characteristic.
 In islands, as well as continental and aquatic systems, researchers still mainly focus on reproductive barriers like environmental and ecological factors to explain speed and occurences of radiations \cite{krugSpeciesGenusRatios2008}\cite{hurlbertProcessesGeneratingLatitudinal2014}\cite{willisSpeciesDiversityScale2002}.
@@
 A promising avenue to this with growing importance is genetic data and analyses.
-Genetic analyses have already helped clearing up phylogenies of many species groups \cite{valenteUsingMolecularPhylogenies2018}%, like bats or cichlids REFS!
+Genetic analyses have already helped clearing up phylogenies and island biogeography of many species groups \cite{austinReconstructingIslandRadiation2004}\cite{emersonDiversificationForestBeetle2005}\cite{valenteUsingMolecularPhylogenies2018}%, like bats or cichlids REFS!
 and more and more complete genomes are published on a regular basis \cite{alonso-blanco135GenomesReveal2016} \cite{lewinEarthBioGenomeProject2018}.
-Some genetic studies highlight the importance of genetic linkage or ``hitchhiking'' on reproductive/ecological isolation of populations \cite{hawthorneGeneticLinkageEcological2001}\cite{viaGeneticMosaicSuggests2008}
+Some genetic studies from continental species highlight the importance of genetic linkage or ``hitchhiking'' on reproductive/ecological isolation of populations \cite{hawthorneGeneticLinkageEcological2001}\cite{viaGeneticMosaicSuggests2008}
 as well as evolutionary rescue provided by the feedback with ecological processes \cite{schiffersLimitedEvolutionaryRescue2012}.
@@
 seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits \cite{hermannTightGeneticLinkage2013}
-In an island context, genetic studies are even rarer and almost exclusively of a descriptive nature.
+Thus, genetic characteristics presents a powerful opportunity for a deeper understanding of speciation processes.
 
 %% Models
-The problem of investigating speciation processes is the time frame and study extent needed for a proper evaluation.
+However, the problem of investigating speciation processes is the time frame and study extent needed to go beyond mere correlation \cite{dormannCorrelationProcessSpecies}.
 Because island biogeography processes and thus radiations generally take place at evolutionary time frames (millions of years),
-comprehensive and meaningful data is difficult and virtually impossible to obtain.
+comprehensive and meaningful data is difficult to obtain (\cite{didierLikelihoodTreeTopologies2017}, but see \cite{mitchellInferringDiversificationRate})
 Therefore, there has been an interest in building process-based island models to test the causality of assumptions and hypotheses.
@@
 But nature is usually more complex than this (cf.\ \cite{silvertownGhostCompetitionPhylogeny2004} but see \cite{herbenGhostHybridizationNiche2005}) 
-and genetic mechanisms should also play a central role for speciation/radiation and species richness.
+and we expect genetic mechanisms to play a central role in speciation/radiation and species richness.
 Questions, such as "What functional traits (e.g. relating to dispersal capacity,
-reproduction, trophic ecology) are associated with high diversification rates within and across island systems?"
+reproduction, trophic ecology) are associated with high diversification rates within and across island systems?" %better another...
 or "What is the influence of gene flow among islands and/or

### 2018-08-01 analyse.R
plotMap() checks for empty files

@@ -82,7 +82,7 @@ plotMap = function(timestep=-1, compensate=TRUE, showinvaders=TRUE) {
         tsvfile = grep(paste("t", timestep, sep=""), list.files(outdir), value=T)
     }
     tsvfilepath = paste(outdir, tsvfile, sep="/")
-    if (!file.exists(tsvfilepath)) {
+    if (!file.exists(tsvfilepath) || file.info(tsvfilepath)$size == 0) {
         print(paste("WARNING: tsvfile not found", tsvfilepath))
         return()
     }

### 2018-08-01 analyse.R
Made analyse.R more robust

@@ -6,8 +6,9 @@ library(ggplot2)
 
 # The working directory may be specified via the commandline, otherwise it
 # defaults to results/tests
+resultdir = "results"
 simname = commandArgs()[length(commandArgs())]
-outdir = paste0("results/", simname)
+outdir = paste0(resultdir, "/", simname)
 
 
 # Plot the distribution of traits in the population at the given timestep (-1 => END)
@@ -20,7 +21,7 @@ plotTraits = function(timestep=-1, toFile=TRUE) {
         traitfile = grep(paste("t", timestep, sep=""), list.files(outdir), value=T)
     }
     traitfilepath = paste(outdir, traitfile, sep="/")
-    if (!file.exists(traitfilepath)) {
+    if (!file.exists(traitfilepath) || file.info(traitfilepath)$size == 0) {
         print(paste("WARNING: traitfile not found", traitfilepath))
         return()
     }
@@ -42,7 +43,7 @@ plotTraits = function(timestep=-1, toFile=TRUE) {
 # Plot population size and diversity indices over time
 plotDiversity = function(logfile="diversity.log") {
     logfile = paste(outdir, logfile, sep="/")
-    if (!file.exists(logfile)) {
+    if (!file.exists(logfile) || file.info(logfile)$size == 0) {
         print(paste("WARNING: logfile not found", logfile))
         return()
     }
@@ -141,10 +142,10 @@ visualize = function(toFile=TRUE) {
 
 # If the simname is given as 'all', process every folder in 'results'
 if (simname == "all") {
-    for (f in list.files("results")) {
+    for (f in list.files(resultdir)) {
         print(paste("Processing", f))
         simname = f
-        outdir = paste0("results/", simname)
+        outdir = paste0(resultdir, "/", simname)
         if (file.info(outdir)$isdir) visualize()
     }
     print("Done.")
@@ -153,7 +154,7 @@ if (simname == "all") {
     # the specified doesn't exist)
     if (!(file.exists(outdir) && file.info(outdir)$isdir)) {
         simname = "tests"
-        outdir = paste0("results/", simname)
+        outdir = paste0(resultdir, "/", simname)
     }
     visualize()
     print("Done.")

### 2018-08-01 src/io.jl
fix phylo output

@@ -378,16 +378,16 @@ Save the phylogeny of individuals currently in world `w`.
 function writephylo(world::Array{Patch,1}, settings::Dict{String, Any}, timestep::Int)
     if !isfile(joinpath(settings["dest"], "phylo.tsv"))
         open(joinpath(settings["dest"], "phylo.tsv"), "w") do f
-            println(f, "t\tid\tfirstparent\tsecondparent")
+            println(f, "t\tpatch\tid\tparent")
         end
     end
     open(joinpath(settings["dest"], "phylo.tsv"), "a") do f
         for p in world
             for i in 1:size(p.phylo, 1)
                 print(f, timestep, "\t")
+                print(f, p.id, "\t")
                 print(f, p.phylo[i, 1], "\t")
-                print(f, p.phylo[i, 2], "\t")
-                println(f, p.phylo[i, 3])
+                println(f, p.phylo[i, 2])
             end
         end
     end

### 2018-08-01 paper/draft.tex
updated refs, new title

@@
 
-\title{The relative roles of genetic, ecological and environmental factors in insular species radiations}
+\title{The effect of genetic linkage on island endemics richness}%The relative roles of genetic, ecological and environmental factors in insular species radiations}
 \author{Ludwig Leidinger}
@@
 \begin{abstract}
-  Islands are known for their peculiar and fascinating biotas, typically very distinct from any other landmasses, be it neighboring islands or continental areas.
+  Islands often witness fascinating species radiations, which are usually correlated to the islands' unique
+  geo-morphological histories.
+  Yet, why these radiations occur in the first place is still under debate.
+  Besides eco-environmental factors, there is evidence that genetic characteristics can have important
+  effects on the speciation process.
+  One of those characteristics is genetic linkage and describes how interdependent genes are known for their peculiar and fascinating biotas, typically very distinct from any other landmasses, be it neighboring islands or continental areas.
   Although island biogeography theory has seen an ever increasing development over the years,

### 2018-08-01 src/constants.jl
increase fertility, syntax harmonization

@@ -5,7 +5,7 @@ const boltz = 1.38064852e-23 # J/K = m2⋅kg/(s2⋅K)
 const act = 1e-19 # activation energy /J, ca. 0.63eV - Brown et al. 2004
 const growthrate = exp(25.2) # global base growth/biomass production from Brown et al. 2004
 const mortality = exp(22) # global base mortality from Brown et al. 2004 is 26.3, but competition and dispersal introduce add. mort.
-const fertility = exp(26.5) # global base reproduction rate 23.8 from Brown et al. 2004, alternatively 25.0, default 30.0
+const fertility = exp(27.0) # global base reproduction rate 23.8 from Brown et al. 2004, alternatively 25.0, default 30.0
 const phylconstr = 10 #parse(ARGS[2])
 # const meangenes = 20 # mean number of genes per individual
 const mutationrate = 1e7 # one mutation per generation with 200bp genes, corrected for metabolic function

### 2018-08-01 src/types.jl

@@ -5,8 +5,7 @@ mutable struct Trait
     value::Float64
 end
 
-abstract type AbstractGene
-end
+abstract type AbstractGene end
 
 mutable struct Gene <: AbstractGene
     sequence::Int

### 2018-08-01 src/mainprocs.jl
fix probability calculation

@@ -26,7 +26,7 @@ function mutate!(ind::Individual, temp::Float64, settings::Dict{String, Any})
         for idx in eachindex(chrm.genes)
             charseq = collect(num2seq(chrm.genes[idx].sequence))
             for i in eachindex(charseq)
-                if rand() <= ind.traits["mutprob"] * exp(-act/(boltz*temp))
+                if rand() <= 1 - exp(-ind.traits["mutprob"] * exp(-act/(boltz*temp)))
                     newbase = rand(collect("acgt"),1)[1]
                     while newbase == charseq[i]
                         newbase = rand(collect("acgt"),1)[1]

### 2018-08-03 paper/draft.tex
add comments with references to Rangel2018 paper
add to and edit MM

@@
 %% Models
+%from Rangel et al 2018 (science, cradles/museums/graves):
+%''spatially and temporally explicit, process-based models (18, 19),
+%founded on a comprehensive suite of well-studied, widely accepted mechanisms,
+%have the greatest potential to assess the complex and sometimes indeterminate
+%interactions among underlying processes''
+%and:
+%"The second-most-influential model parameter was the maximum sustainable
+%evolutionary rate realizable by a population (Hmax),
+%which limits the adaptability of niche limits and evolutionary rescue (82–85)
+%in the face of changing climates (figs. S16 to S19 and table S7).
+%Low Hmax values indicate that niche traits have low genetic variance,
+%low population growth rates, or both—preventing species from tracking
+%and adapting to changing climates."
+%==> what affects evolutionary rate?
+
 However, the problem of investigating speciation processes is the time frame and study extent needed to go beyond mere correlation \cite{dormannCorrelationProcessSpecies}.
@@
 %% Experimental design
-We use an individual-based metacommunity model written in julia %\cite{JULIALANG}.
-Each individual is characterized by a set of genetically controlled (i.e. genomically-coded) traits,
-which influence ecological processes of
-establishment, competition, growth, (density independent) mortality, reproduction and dispersal.
-The simulation arena consists of a larger landmass (the mainland) holding an initial random species pool and a smaller,
-initially unpopulated island some distance apart (user-defined in the experimental design).
-% it would be nice to hav a table with all model configurations that can be defined by the user upon designing the experiment.
-% This might help readers see that the model is general enough to be applied to any ecological arena (mainland-island being one).
-During the course of the simulation, individuals may disperse from the mainland to the island, thus colonizing it. 
-The complete state of the simulation is recorded every 1000 time steps for analysis.
 
-Building on the model of Cabral et al. (2017a, b), most of the biological rates follow the metabolic theory
+\begin{figure}
+  %\includegraphics{}
+  \caption{Schematic of the model's individuals genetic architecture.
+    Individuals are composed of a diploid set of linkage units which combine the genes coding for functional traits.}
+  \label{schematic}
+\end{figure}
+
+We use an individual-based spatially and temporally explicit metacommunity model written in julia %\cite{JULIALANG}.
+The model considers competition between individuals, environmental niche adaptation, density independent mortality,
+dispersal, sexual reproduction and mutation.
+Individuals are characterized by a lineage identifier,
+body mass, age and fitness.
+Each individual's life history traits (table) are coded by mutable genes \cref{fig:schematic}.
+The genes, in turn, may be combined to linkage units (``chromosomes'').
+Genes in a linkage unit are considered a heredetary unit.
+This means that they will always be inherited in junction.
+Thus, only complete linkage units will be subject to recombination.
+
+Building on the model of Cabral et al. (2017a, b), reproduction rates (as number of offspring),
+growth rates and mortality probabilities follow the metabolic theory
 of ecology (MTE, Brown et al.\ 2004). % instead of saying 'most', say which follow and which don't
@@
 
+Dispersal is controlled by a logistic dispersal kernel \cite{bullockjamesm.SynthesisEmpiricalPlant2016}
+with parameters PAR1 and PAR2
+for mean dispersal kernel and kernel shape (tail fatness), respectively.
+
 A detailed protocol can be found in the appendix (\cite{grimm2010odd}, XXX).
+
+The simulation arena consists of a larger landmass (the mainland) holding an initial random species pool and a smaller,
+initially unpopulated island some distance apart (user-defined in the experimental design).
+% it would be nice to hav a table with all model configurations that can be defined by the user upon designing the experiment.
+% This might help readers see that the model is general enough to be applied to any ecological arena (mainland-island being one).
+During the course of the simulation, individuals may disperse from the mainland to the island, thus colonizing it. 
+The complete state of the simulation is recorded every 1000 time steps for analysis.
 
@@
 In the simulations, we modify the value and extent of
-(a) gene linkage,
-(b) reproductive isolation (or reproduction compatibility),
-(c) habitat diversity, and
-(d) topographical complexity
+%(a)
+gene linkage
+%(b) reproductive isolation (or reproduction compatibility),
+%(c) habitat diversity, and
+%(d) topographical complexity
 to asses their relative contribution on community composition.
+Additionally, we simulate lineages in isolation with varying degrees of genetic linkage
+and analyize emerging species numbers.
 \cref{tab:scenarios} summarizes the parameters which were varied for the scenarios, their meaning and their values. %TODO: make table.
 Each parameter combination was replicated %TODO: how much?
-times, setting a different random seed for each replicate.
+times.%, setting a different random seed for each replicate.
 
 \subsection{Data recording}
-We record the complete state of our simulation world at the start and end of a simulation run and each time the arena changes (geomorphological dynamics).
-Additionally, we store the state of the simulation arena each 1000 %TODO: how much exactly?
-time steps and record every individual that colonizes the island to track changes.
-
-Using the thus simulated data, we calculate phylogenies of the colonizing lineages, species numbers, trait spaces etc. [WHAT ELSE?]
+We record individual data (properties and genome) at the start and end of a simulation run
+and at intervals of 1000 time steps.% and each time the arena changes (geomorphological dynamics).
+Additionally, we keep track of the genealogy of our individuals through time,
+allowing us to validate inferred phylogenies.
+Using the thus simulated data, we calculate phylogenies of the colonizing lineages, emerging species numbers and
+trait space. % etc. [WHAT ELSE?]
 
@@
   %\includegraphics{}
-  \caption{Number of linkage units and number of species per lineage XXX timesteps after colonization.
-    [Histogram (with threshold category: strongly diversified lineages) or scatter plot. ]}%Two panels/color codes: high and low degree of environmental heterogeneity]}
+  \caption{[2 x 2]
+    Extinctions and numbers of species per lineage after XXXX timesteps against degree of genetic linkage.
+    Top row: lineages simulated in isolation.
+    Bottom row: simulated communities.}
   \label{specieslinkage}
@@
   %\includegraphics{}
-  \caption{Exemplary phylogenies of lineages without co-occurring competing lineages calculated from neutral sequences of
+  \caption{[2 x 3]
+    Exemplary genealogies of lineages in isolation for different degrees of genetic linkage.
     extant individuals.
-    (A) Lineage with low gene linkage,
-    (B) lineage with intermediate gene linkage,
-    (C) lineage with high gene linkage.}
+    Columns: phylogeny and genealogy
+    Rows: Lineage with low gene linkage,
+    lineage with intermediate gene linkage,
+    lineage with high gene linkage.}
   \label{phylogenies}

### 2018-08-03 analyse.R
Rearranged analyse.R and started on plotEstablishment()

@@ -11,7 +11,81 @@ simname = commandArgs()[length(commandArgs())]
 outdir = paste0(resultdir, "/", simname)
 
 
-# Plot the distribution of traits in the population at the given timestep (-1 => END)
+### ANALYSE THE WHOLE EXPERIMENT
+
+collateSpeciesTable = function(rundir, timestep, compensate=TRUE, showinvaders=TRUE) {
+    ## load raw data
+    tsvfile = grep(".tsv", grep(paste0("t", timestep, "_"), list.files(rundir), value=T), value=T)
+    if (length(tsvfile) == 0 && compensate) {
+        # If the desired timestep doesn't exist, take the newest timestep we have
+        timestep = (length(grep(".tsv", list.files(rundir), value=T))-1) * 10
+        tsvfile = grep(paste("t", timestep, sep=""), list.files(rundir), value=T)
+    }
+    tsvfilepath = paste(rundir, tsvfile, sep="/")
+    if (!file.exists(tsvfilepath) || file.info(tsvfilepath)$size == 0) {
+        print(paste("WARNING: tsvfile not found", tsvfilepath))
+        return()
+    }
+    ts = read.table(tsvfilepath, header=T)
+    ts$temp.C = ts$temp - 273
+    ## load comparison data (to show invasive species)
+    if (showinvaders && timestep < 1000 && timestep != -1) showinvaders = FALSE
+    if (showinvaders) {
+        tsvfile2 = grep(".tsv", grep("t1000_", list.files(rundir), value=T), value=T)
+        tsvfilepath2 = paste(rundir, tsvfile2, sep="/")
+        ts2 = read.table(tsvfilepath2, header=T)
+        nativespecs = unique(ts2$lineage)
+    }
+    ## create an index of species abundance
+    allspecies = c()
+    specidx = sort(unique(ts$lineage))
+    for (p in unique(ts$id)) {
+        patch = ts[which(ts$id == p),]
+        x = patch$xloc[1]
+        y = patch$yloc[1]
+        for (spec in unique(patch$lineage)) {
+            n = sum(which(patch$lineage == spec))
+            s = which(specidx == spec)
+            if (showinvaders && !(spec %in% nativespecs)) i = TRUE
+            else i = FALSE
+            allspecies = rbind(allspecies, c(x,y,s,n,i))
+        }
+    }
+    colnames(allspecies) = c("xloc", "yloc", "lineage", "abundance", "alien")
[…]

### 2018-08-03 analyse.R
Finished `analyseEstablishment()`
Also bugfixes in `collateSpeciesTable()`

@@ -22,10 +22,11 @@ collateSpeciesTable = function(rundir, timestep, compensate=TRUE, showinvaders=T
         tsvfile = grep(paste("t", timestep, sep=""), list.files(rundir), value=T)
     }
     tsvfilepath = paste(rundir, tsvfile, sep="/")
-    if (!file.exists(tsvfilepath) || file.info(tsvfilepath)$size == 0) {
+    if (!file.exists(tsvfilepath) || file.info(tsvfilepath)$isdir || file.info(tsvfilepath)$size == 0) {
         print(paste("WARNING: tsvfile not found", tsvfilepath))
         return()
     }
+    print(paste("Collating data from", tsvfilepath)) #DEBUG
     ts = read.table(tsvfilepath, header=T)
     ts$temp.C = ts$temp - 273
     ## load comparison data (to show invasive species)
@@ -38,50 +39,77 @@ collateSpeciesTable = function(rundir, timestep, compensate=TRUE, showinvaders=T
     }
     ## create an index of species abundance
     allspecies = c()
-    specidx = sort(unique(ts$lineage))
     for (p in unique(ts$id)) {
-        patch = ts[which(ts$id == p),]
+        patch = subset(ts, id == p)
         x = patch$xloc[1]
         y = patch$yloc[1]
-        for (spec in unique(patch$lineage)) {
-            n = sum(which(patch$lineage == spec))
-            s = which(specidx == spec)
-            if (showinvaders && !(spec %in% nativespecs)) i = TRUE
+        for (s in unique(patch$lineage)) {
+            n = length(subset(patch, lineage == s)$lineage)
+            if (showinvaders && !(s %in% nativespecs)) i = TRUE
             else i = FALSE
             allspecies = rbind(allspecies, c(x,y,s,n,i))
         }
     }
     colnames(allspecies) = c("xloc", "yloc", "lineage", "abundance", "alien")
     allspecies = as.data.frame(allspecies)
-    allspecies$alien = as.factor(allspecies$alien)
+    allspecies$abundance = as.numeric(as.character(allspecies$abundance))
     return(allspecies)
 }
 
-plotEstablishment = function() {
-    ##TODO
+analyseEstablishment = function() {
     print("Analysing invasion success")
-    results = array(dim=c(2,2,2,6), dimnames=list(temperature=c("T35", "T25"),
-                                                  disturbance=c("1DB", "10DB"),
-                                                  propagules=c("1PP", "10PP"),
[…]

### 2018-08-03 paper/draft.tex
add some refs

@@
 
-We use an individual-based spatially and temporally explicit metacommunity model written in julia %\cite{JULIALANG}.
+We use an individual-based spatially and temporally explicit metacommunity model written in julia \cite{bezansonJuliaFreshApproach2017}.
 The model considers competition between individuals, environmental niche adaptation, density independent mortality,
@@
 
-A detailed protocol can be found in the appendix (\cite{grimm2010odd}, XXX).
+A detailed protocol can be found in the appendix \cite{grimmStandardProtocolDescribing2006a} \cite{grimmODDProtocolReview2010}.
 

### 2018-08-03 src/io.jl
update fasta header creation

@@ -214,7 +214,6 @@ function makefasta(world::Array{Patch, 1}, io::IO = STDOUT, sep::String = "", on
         (onlyisland && !patch.isisland) && continue
         lineage = ""
         for ind in patch.community
-            counter += 1
             (!patch.isisland && ind.lineage == lineage) && continue # only one individual per species on mainland
             chrmno = 0
             for chrm in ind.genome
@@ -230,7 +229,7 @@ function makefasta(world::Array{Patch, 1}, io::IO = STDOUT, sep::String = "", on
                             traits *= string(trait.nameindex) * string(trait.value) * ","
                         end
                     end
-                    header = ">"*string(counter)*" x"*string(patch.location[1])*" y"*string(patch.location[2])
+                    header = ">"*string(ind.id)*" x"*string(patch.location[1])*" y"*string(patch.location[2])
                     header *= " "*ind.lineage*" c"*string(chrmno)*" g"*string(geneno)*" "*traits
                     println(io, header)
                     println(io, num2seq(gene.sequence))

### 2018-08-20 paper/draft.tex
results: temporal dynamics of radiations FL vs NL

@@
 
+\subsection{Single species scenarios}
+Species evolving on its own without interference from other species showed complex and dynamic speciation/differentiation patterns.
+While low-linkage lineages adapted to the new environmental conditions on the island within some thousand timesteps to produce distinct populations,
+high-linkage lineages were noticeably slower to differentiate.
+Several 10.000 years later, however, high-linkage lineages began with strong radiations, thus surpassing species numbers as compared to low-linkage scenarios.
+
+\begin{figure}
+  %\includegraphics{}
+  \caption{[2 x 2]
+    Emerging numbers of species (and average population sizes) over time.
+    Low-linkage lineages in COLOR1, high-linkage lineages in COLOR2.
+    Top row: lineages simulated in isolation.
+    Bottom row: simulated communities.}
+  \label{specieslinkage}
+\end{figure}
+
 \begin{figure}

### 2018-08-20 paper/draft.tex
disc: temporal dynamics of radiations FL vs NL

@@
 \section{Discussion}
+The fact that low-linkage (single) species speciated in general faster than their high-linkage counterparts
+is in line with expectations given their higher potential for fast evolutionary response [as exemplified REFS].
+What did come as a surprise, however was that, over time, high linkage lineages surpassed low linkage lineages
+in the number of emerging species.
+
 %Assortative mating

### 2018-08-20 paper/draft.tex
disc: reasons for temp dyns of radiations FL vs NL

@@
 in the number of emerging species.
+The reason for this is likely to be found in the high genetic variation throughout the high-linkage populations
+due to genetic hitchhiking.
+Combined with random drift, this genetic variation has a high potential of producing numerous species over time [non-adaptive radiation].
+[EXAMPLES]
+On the other hand, independent recombination of genes in low-linkage populations explains their quick adaptive response
+to the conditions in their respective environments.
+Once adapted and having reached a high degree of specialization, however, the genetic architecture leads to strong selection against
+new and mainly deleterious mutations.
+[EXAMPLES]
+
+[SPECULATION:] In mixed community scenarios this effect becomes less clear as inter-lineage competition and red-queen-dynamics
+present complex challenges.
 

### 2018-08-23 src/io.jl
actually print timestep info

@@ -152,13 +152,13 @@ function dumpinds(world::Array{Patch, 1}, settings::Dict{String, Any}, timestep:
             (!patch.isisland && settings["static"] && ind.lineage == lineage) && continue
             if header
                 #XXX Transfer to a dynamic system? (As in createworld()?)
+                print(io, "time", sep)
                 print(io, "patch_no", sep)
                 print(io, "xloc", sep)
                 print(io, "yloc", sep)
                 print(io, "temp", sep)
                 print(io, "area", sep)
                 print(io, "prec", sep)
-                ## print(io, "nicheb", sep)
                 print(io, "island", sep)
                 print(io, "isolation", sep)
                 print(io, "invasible", sep)
@@ -178,6 +178,7 @@ function dumpinds(world::Array{Patch, 1}, settings::Dict{String, Any}, timestep:
                 println(io)
                 header = false
             end
+            print(io, timestep, sep)
             print(io, patch.id, sep)
             print(io, patch.location[1], sep)
             print(io, patch.location[2], sep)

### 2018-08-23 src/io.jl
prevent multiple header printing

@@ -140,6 +140,7 @@ Output all data of individuals in `world` as table to `io`. Columns are separate
 """
 function dumpinds(world::Array{Patch, 1}, settings::Dict{String, Any}, timestep::Int, io::IO = STDOUT, sep::String = "\t")
     header = true
+    timestep > 0 && (header = false)
     traitkeys = []
     onlyisland = settings["static"] && timestep > 1
     for patch in world

### 2018-08-23 src/io.jl
get traitnames from settings

@@ -172,8 +172,7 @@ function dumpinds(world::Array{Patch, 1}, settings::Dict{String, Any}, timestep:
                 print(io, "size", sep)
                 print(io, "lnkgunits", sep)
                 print(io, "ngenes", sep)
-                traitkeys = keys(ind.traits) #XXX This will give problems if we lose traits
-                for key in traitkeys
+                for key in settings["traitnames"]
                     print(io, key, sep)
                 end
                 println(io)
@@ -199,7 +198,7 @@ function dumpinds(world::Array{Patch, 1}, settings::Dict{String, Any}, timestep:
             print(io, ind.size, sep)
             print(io, length(ind.genome), sep)
             print(io, sum(map(x -> length(x.genes), ind.genome)), sep)
-            for key in traitkeys
+            for key in settings["traitnames"]
                 try
                     print(io, ind.traits[key], sep)
                 catch

### 2018-08-23 src/io.jl
fix: print header only once

@@ -140,7 +140,7 @@ Output all data of individuals in `world` as table to `io`. Columns are separate
 """
 function dumpinds(world::Array{Patch, 1}, settings::Dict{String, Any}, timestep::Int, io::IO = STDOUT, sep::String = "\t")
     header = true
-    timestep > 0 && (header = false)
+    timestep != 0 && (header = false)
     traitkeys = []
     onlyisland = settings["static"] && timestep > 1
     for patch in world

### 2018-08-23 islandsim.jl
stop writing data at end of simulation

@@ -68,7 +68,6 @@ function runit(prerun::Bool = false)
         i > 1 && updateworld!(world,maptable,settings["cellsize"])
         !prerun && writedata(world, settings, settings["maps"][i], 0)
         simulation!(world, settings, settings["maps"][i], timesteps)
-        !prerun && writedata(world, settings, settings["maps"][i], -1)
     end
 end
 

### 2018-08-23 src/defaults.jl
tolerance setting as value

@@ -15,7 +15,8 @@ function defaultSettings()
          # main model settings
          "linkage" => "random", # gene linkage type (random/full/none)
          "nniches" => 2, # number of environmental niches (max. 3)
-         "tolerance" => "low", # sequence similarity threshold for reproduction
+         "tolerance" => 0.8, # sequence similarity threshold for reproduction
+         "fixtol" => true,
          "static" => true, # mainland sites don't undergo eco-evolutionary processes
          "mutate" => true, # mutations occur
          "initadults" => false, # initialize organisms as adults

### 2018-08-23 src/io.jl

@@ -55,9 +55,9 @@ function parsecommandline()
             range_tester = x -> x > 0 && x <= 3
             required = false
         "--tolerance", "-t"
-            help = "tolerance of sequence identity when reproducing (\"high\", \"evo\", \"low\" or \"none\")"
-            arg_type = String
-            range_tester = x->in(x,["high", "evo", "low", "none"])
+            help = "tolerance of sequence identity when reproducing"
+            arg_type = Float64
+            range_tester = x -> 0.0 <= x <= 1.0
             required = false
         "--dest", "-d"
             help = "output directory. Defaults to current date"
@@ -268,6 +268,7 @@ function setupdatadir(settings::Dict{String, Any})
                 cp(m, joinpath(settings["dest"], m), remove_destination = true) # most likely replicates with same parameters
             end
         end
+        cp(joinpath("src", "constants.jl"), joinpath(settings["dest"], "constants.jl"), remove_destination = true)
     end
 end
 

### 2018-08-23 src/mainprocs.jl

@@ -4,7 +4,7 @@ function mutate!(traits::Array{Trait, 1}, settings::Dict{String, Any})
     for trait in traits
         traitname = settings["traitnames"][trait.nameindex]
         (contains(traitname, "mutprob") && mutationrate != 0) && continue
-        contains(traitname, "reptol") && settings["tolerance"] != "evo" && continue # MARK CAVE!
+        contains(traitname, "reptol") && continue # MARK CAVE!
         oldvalue = trait.value
         contains(traitname, "tempopt") && (oldvalue -= 273)
         while oldvalue <= 0 # make sure sd of Normal dist != 0

### 2018-08-23 src/subprocs.jl

@@ -287,15 +287,7 @@ function createtraits(settings::Dict{String, Any}) #TODO: this is all very ugly.
         elseif contains(traitnames[idx], "prectol")
             push!(traits, Trait(idx, rand() + 0.39))
         elseif contains(traitnames[idx], "reptol")
-            if settings["tolerance"] == "high"
-                push!(traits, Trait(idx, 0.8))
-            elseif settings["tolerance"] == "low"
-                push!(traits, Trait(idx, 0.95)) #CAVE: code values elsewhere?
-            elseif settings["tolerance"] == "none"
-                push!(traits, Trait(idx, 0.01)) #CAVE: code values elsewhere?
-            else
-                push!(traits, Trait(idx, 0.5 + rand() * 0.5))
-            end
+            push!(traits, Trait(idx, settings["tolerance"]))
         else
             push!(traits, Trait(idx, rand()))
         end

### 2018-08-23 src/mainprocs.jl
use setting fixtol in mutate!()

@@ -4,7 +4,7 @@ function mutate!(traits::Array{Trait, 1}, settings::Dict{String, Any})
     for trait in traits
         traitname = settings["traitnames"][trait.nameindex]
         (contains(traitname, "mutprob") && mutationrate != 0) && continue
-        contains(traitname, "reptol") && continue # MARK CAVE!
+        contains(traitname, "reptol") && settings["fixtol"] && continue # MARK CAVE!
         oldvalue = trait.value
         contains(traitname, "tempopt") && (oldvalue -= 273)
         while oldvalue <= 0 # make sure sd of Normal dist != 0

### 2018-08-23 analysis/ana.R
revise analysis script

@@ -4,14 +4,14 @@
 ## Ludwig Leidinger <ludwig.leidinger@uni-wuerzburg.de>
 ## Dec., 2017
 
-library(stats) # what for?
+#library(stats) # what for?
 library(ape)
 library(ggplot2)
 library(ggtree)
 library(dplyr)
 library(vegan)
 library(viridis)
-#library(phyloseq)
+library(paleotree)
 
 
 args = commandArgs(trailingOnly = TRUE)
@@ -26,20 +26,9 @@ for(basename in basenames){
     ## read main data:
     allworld = read.table(paste0(basename, ".tsv"), header = T)
 
-    nameparts = unlist(strsplit(basename, "_"))
-    nameparts[length(nameparts) - 1] = "t0"
-    mlname = paste(nameparts, collapse = "_")
-    mlworld = read.table(paste0(mlname, ".tsv"), header = T)
-
     ## read sequences (fasta format!):
     allseqs = read.dna(file=paste0(basename, ".fa"), format="fasta")
     headers = names(allseqs[sapply(allseqs, length) == 200])[c(TRUE,FALSE)]
-
-    mlseqs = read.dna(file=paste0(mlname, ".fa"), format="fasta")
-    mlheaders = names(mlseqs[sapply(mlseqs, length) == 200])[c(TRUE,FALSE)]
-
-    ## make sure seqs and data have same length and elements:
-    allworld = allworld %>% filter(island == 1)
 
     allworld$tips = headers
     cols = ncol(allworld)
@@ -52,35 +41,8 @@ for(basename in basenames){
     allworld$linkage = "intermediate"
     allworld$linkage[allworld$lnkgunits == 2] = "full"
     allworld$linkage[allworld$lnkgunits >= 22] = "none"
-    nameparts = unlist(strsplit(basename, "_"))
-    time = nameparts[length(nameparts) - 1]
-    time = strsplit(time, "t")
-    time = as.numeric(rev(unlist(time))[1])
-    allworld$time = time
     allworld$FAD = abs(max(allworld$time) - allworld$time) + allworld$age
     allworld$LAD = abs(max(allworld$time) - allworld$time)
-
-    mlworld$tips = mlheaders
-    cols = ncol(mlworld)
-    mlworld = mlworld[,c(cols,1:(cols-1))]
-    names(mlworld)[names(mlworld) == "id"] = "ID"
[…]

### 2018-08-23 paper/draft.tex
discuss plant polyploidisation

@@
 
+In plants polyloidisation leads to increased speciation or adaptive radiation potential \cite{baduelPolyploidHopShifting2018}.
+Although our model considers linkage and thus ploidy level as static, we still can draw some conclusions relevant to the issue
+of why autopolyploids much rarer than paleopolyploids.
+[in the beginning great variation, but as specialisation increases, mutations outcompeted]
+[with high specialisation no evolutionary rescue in case of disturbance]
 [SPECULATION:] In mixed community scenarios this effect becomes less clear as inter-lineage competition and red-queen-dynamics

### 2018-08-24 paper/draft.tex
include JSC comments

@@
 \section{Introduction}
+%% WHY ARE GENERA DIFFERENTLY RICH BESIDE ECOLOGY AND GEOGRAPHY/GEOLOGY? GENETICS!
 
@@
 % when radiations DID happen!
-In fact, these factors indeed often correlate strongly with the timing of genetic divergence of species and populations \cite{hembryEvolutionaryBiogeographyTerrestrial}. % REFS!
-But even then, in some cases the reason for maintenance of population diversity remains in the dark. \cite{chenConcordanceGeneticDiversity}
+These factors indeed often correlate strongly with the timing of genetic divergence of species and populations \cite{hembryEvolutionaryBiogeographyTerrestrial}. % REFS!
+But even then, in some cases the reason for maintenance of population variability remains unclear. \cite{chenConcordanceGeneticDiversity}
 %% HERE COME SOME EXAMPLES
-Additionally, these studies highlight the cases of successful radiations, while a conclusive explanation for why these population variablities and thus radiations could occur in the first place is still missing.
+Additionally, these studies highlight the cases of successful radiations, while a conclusive explanation for why these population variabilities and thus radiations could occur in the first place is still missing.
 This becomes especially apparent on islands, where species generally experience nearly identical environmental conditions.%REF
@@
 
-However, the problem of investigating speciation processes is the time frame and study extent needed to go beyond mere correlation \cite{dormannCorrelationProcessSpecies}.
+The problem of investigating speciation processes is the time frame and study extent needed to go beyond mere correlation \cite{dormannCorrelationProcessSpecies}.
 Because island biogeography processes and thus radiations generally take place at evolutionary time frames (millions of years),
@@
 divergence event if the population remains isolated during that time.
-But nature is usually more complex than this (cf.\ \cite{silvertownGhostCompetitionPhylogeny2004} but see \cite{herbenGhostHybridizationNiche2005}) 
+But nature is usually more complex than this %(cf.\ \cite{silvertownGhostCompetitionPhylogeny2004} but see \cite{herbenGhostHybridizationNiche2005}) 
 and we expect genetic mechanisms to play a central role in speciation/radiation and species richness.
@@
 %% In this study, we aim to address...
-To assess the role of genetic processes, namely genetic linkage, for island radiations,
+To assess the role of genetic factors, namely genetic linkage, for island radiations
 we present a mechanistic model that overcomes above-mentioned shortcomings like missing micro-evolutinary processes and lack of integrative design
@@
 levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption.
+%Provide Refs!
 %what Schiffers et al say about linkage and evolutionary rescue (basically the opposite of diversification?)
+%After presenting the study questions and expectations, you should say how the questions are going to be answered  (e.g. short explain simulation experimental design).
 
@@
     Individuals are composed of a diploid set of linkage units which combine the genes coding for functional traits.}
-  \label{schematic}
+  \label{fig:schematic}
 \end{figure}
 
-We use an individual-based spatially and temporally explicit metacommunity model written in julia \cite{bezansonJuliaFreshApproach2017}.
+We use an individual-based spatially and temporally explicit metacommunity model written in julia \cite{bezansonJuliaFreshApproach2017} .  \ref{fig:schematic}
 The model considers competition between individuals, environmental niche adaptation, density independent mortality,
@@
 
-\subsection{scenarios}
+\subsection{Data recording}
+We record individual data (properties and genome) at the start and end of a simulation run
+and at intervals of 1000 time steps.% and each time the arena changes (geomorphological dynamics).
+Additionally, we keep track of the genealogy of our individuals through time,
+allowing us to validate inferred phylogenies.
+Using the thus simulated data, we calculate phylogenies of the colonizing lineages, emerging species numbers and
+trait space. % etc. [WHAT ELSE?]
+
+\subsection{Experimental design and scenarios}
 In the simulations, we modify the value and extent of
@@
 times.%, setting a different random seed for each replicate.
-
-\subsection{Data recording}
-We record individual data (properties and genome) at the start and end of a simulation run
-and at intervals of 1000 time steps.% and each time the arena changes (geomorphological dynamics).
-Additionally, we keep track of the genealogy of our individuals through time,
-allowing us to validate inferred phylogenies.
-Using the thus simulated data, we calculate phylogenies of the colonizing lineages, emerging species numbers and
-trait space. % etc. [WHAT ELSE?]
 
@@
     Top row: lineages simulated in isolation.
-    Bottom row: simulated communities.}
+    !Bottom row: simulated communities.!}
   \label{specieslinkage}
@@
   \caption{[2 x 2]
+    ???
     Extinctions and numbers of species per lineage after XXXX timesteps against degree of genetic linkage.
@@
   %\includegraphics{}
-  \caption{Trait space characteristics of lineages.
+  \caption{Trait space characteristics of lineages./Evolution of trait values over time against linkage
     (A) [Ordination] Trait space of successful colonizers compared to the mainland species pool. Each point represents a lineage.
@@
 \section{Discussion}
+%% GENETIC ARCHITECTURE/DYNAMICS STUDIED FOR DECADES. WITH GENETIC/BIG DATA, APPLY in (ISLAND) BIOGEOGRAPHY.
+
+% Hypothesis: Species rich lineages in island systems due to adaptive radiation, thus low linkage.
+% high linkage species groups maybe not even considered because different taxonomic level?
+
 The fact that low-linkage (single) species speciated in general faster than their high-linkage counterparts
@@
 %Clonality
+
+Linkage as dynamic process: polyploidy, TEs, etc.
+Here we investigate the long term effects of fixed linkage degrees on species richness.
 

### 2018-08-24 paper/draft.tex
some disc

@@
   \caption{[2 x 3]
-    Exemplary genealogies of lineages in isolation for different degrees of genetic linkage.
-    extant individuals.
-    Columns: phylogeny and genealogy
+    Exemplary phylogenies of lineages in isolation for different degrees of genetic linkage over time.
     Rows: Lineage with low gene linkage,
@@
 
+Besides patterns of species and genetic diversity, our model is able to produce all relevant data for
+numerous analyses like rank-abundance, species-area-relations, $\alpha$-, $\beta$-, $\gamma$-diversity etc.
+This creates additional possibilities of comparing, validating and calibrating the model to real world patterns \cite{SimpleModelsLead2013}. % other paper? -> check mech. isl. mods-review
+
 In reality linkage between genes is not a binary decision, but rather a consequence of the distance between those genes.
@@
 these processes as well.
-In fact, genetic processes might play an important role in the issue of extinction debt. %% cite Ludmilla, ask for refs
+In fact, genetic processes might play an important role in the issue of extinction debt. %% cite Ludmilla
 
